### 2004-09-29 classes/711/mp10/mp10.tex
initial import

\documentclass[12pt]
{article}

\usepackage{amssymb}
\usepackage{alltt}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{problem}[theorem]{Problem}

\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\newenvironment{solution}[1][Solution]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\newcommand{\qed}{\nobreak \ifvmode \relax \else
      \ifdim\lastskip<1.5em \hskip-\lastskip
      \hskip1.5em plus0em minus0.5em \fi \nobreak
      \vrule height0.75em width0.5em depth0.25em\fi}

\title{CSG711: Problem Set 10}
\author{Sam Tobin-Hochstadt}
\date{\today}

\begin{document}

\maketitle

\begin{problem}
The program is
\begin{verbatim}
x = x0; 
y = y0; 
acc = 0;
while (x != 0) 
{
  acc = acc+y; 
  x = x-1
}
\end{verbatim}
\end{problem}

\begin{solution}
The correct invariant is $x*y + acc = x_0*y_0$.  It is obvious that
this is true at the beginning of the while loop.  After each
execution, we subtract one from $x$, and add $y$ to $acc$.  Since
$(x-1)*y+y+acc = x*y + acc$ for all values of $x$,$y$ and $acc$, this
invariant is preserved by the while loop.  Finally, when $x=0$, then
$x*y + acc = acc$, and therefore $x=0 \Rightarrow acc = x_0*y_0$.
\end {solution}

\begin{problem}
Write a program to find integer square-root by binary search. Your
program should satisfy the same specification as the integer
square-root program in the notes, but it should run in time O(log
n). Write an invariant or invariants that describe the behavior of the
program.
\end {problem}

\begin{solution}
  The program is:
  \begin{verbatim}
z := 0
u := 0
v := 0
w := 1 ;
d := 2 ;
dd := x ;
while (dd > 0){
    w := w+w+w+w
    d := d+d
    dd := dd/2
} ;

while d > 0{
    if ((u + v + w) == x)
    then {
        z := z+d/2 ;
        d := 0
    }
    else if (u + v + w > x) 
    then {
        d := d/2 ;
        u := u ;
        v := v/2 ;
        w := w/4
    }
    else {
        z := z+d/2 ;
        d := d/2 ;
        u := u + v + w ;
        v := v/2 + w ;
        w := w/4
    }
}
  \end{verbatim}

The invariant is $\sqrt{x_0} \leq upper^2 ~\wedge~ \sqrt{x_0} \geq
lower^2 ~\land~ u = z^2 ~\land~ v = zd ~\land~ w = {d^2}/{4}$.  It is
obvious that if $upper = lower$, then $upper^2 = lower^2 = x_0$.
Further, from the invariant, it is clear that $u + v + w =
(z+d/2)^2$.

Now we must show that we initialize the variables $u,v,w$ correctly,
and maintain the invariant throughout the program. Since we start with
$z = 0$, it is clear that $u = 0$  and $v = 0$ are correct.  All that
remains is $w$, which equals $d^2/4$.  In doing this, we must ensure
that $d > \sqrt{x}$ to start with.  We do this by setting $d =
2^{\log_2 d}$ and $w=(4^{\log_2 d - 1})$ which ensures that $d \geq x$
and $w = d^2/4$ at the beginning of the program.

It is easy to show that each of the three cases of the while loop
preserves the invariant.  If $x = u+v+w$, then $z+d/2 = \sqrt{x}$, and
we are done.  Since $z' = z+d/2$ and $d' = 0$, $u$, $v$ and $w$ do not
change.    In the second case, we halve $d$, and thus do not change
$u$, halve $v$ (which is $zd$) and divide $w$ (which is $d^2/4$) by
4.  In the final case, $u' = z'^2 = (z+d/2)^2 = u + v + w$.  Also, $w$
is again quartered, and $v' = k'd' = (z+d/2)*d/2 = zd/2 + d^2/4 = v +
w/4$.  So the invariants are all preserved.

Finally, the above code was tested as a Python program (attached), and
actually works.
\end{solution}

\begin{problem}
The binary-search program in this week's lecture notes in fact works
fine even if the array A has multiple occurrences of x. Modify the
invariant so that it is preserved in this case.

The original invariant is

\[
(1 < lo) \land (hi \leq N) \land 
(\forall j)((0 < j < lo \Rightarrow A[j] < x)
\land (hi < j \leq N \Rightarrow  A[j] > x))
\]

\end {problem}

\begin{solution}
Merely modifying the original invariant to use $\leq$ instead of $<$
would not work, since then if $lo = x$ and $hi = x$, one of the sides
of the conjunction would be false.

Instead, we need to weaken the invariant slightly, by disjoining the
body of the $\forall$ with $A[lo] = x \lor A[hi] = x$, giving us the
final invariant

$(1 < lo) \land (hi \leq N) \land (\forall j)((0 < j < lo \Rightarrow A[j] < x)
\land (hi < j \leq N \Rightarrow  A[j] > x)
\lor((A[lo] = x) \lor (A[hi] = x)))$

This is correct for the cases where there are multiple x's in array
$A$, since if there are, then the only interesting case is when one of
the $lo$ or $hi$ references points to one of the x's.  The added case
deals with this situation.

\end{solution}

\end{document}

### 2004-09-29 classes/hopl/control/control.tex
initial import

\documentclass[12pt]{article}

\usepackage{amssymb}
\usepackage[all]{xy}
\usepackage{amsmath}
\usepackage{fullpage}
\usepackage{enumerate}

\title{Modern Control Operators}
\author{Sam Tobin-Hochstadt}
\date{\today}

\input xy
\xyoption{all}

\begin{document}
\bibliographystyle{plain}

\maketitle

\makebox{\hspace{21mm}
\xymatrix{{}& *++[F--]{~~~~~\cite{monads}~~~~~}& *++[F--]{~~~~~\cite{sitaram94thesis}~~~~~} & *++[F--]{~~~~~\cite{luc94partial}~~~~~}\\
& {} & *++[F]{~~~~~\cite{sitaram93handling}~~~~~}\ar[u] & *++[F--]{~~~~~\cite{queinnec93library}~~~~~} \ar[u]
\\ &{} & {} &*++[F]{~~~~~\cite{queinnec91dynamic}~~~~~} \ar[u] 
\\ &{} & {}& *++[F]{~~~~~\cite{hieb90continuations}~~~~~} \ar@/^/[uul]  \ar[u]   \\
& *++[F]{~~~~~\cite{danvy90abstracting}~~~~~} \ar@/^/[urr]  \ar[uuuu] \ar@//[uurr] \ar@//[uuurr] &*++[F]{~~~~~\cite{sitaram90control}~~~~~} \ar[ur] \ar[uuu] \ar[l] \ar@/^/[uuur]\\&
  {}&*++[F]{~~~~~\cite{felleisen87syntactic}~~~~~} \ar[u] \\
  &{}&*++[F--]{~~~~~\cite{felleisen86secd}~~~~~} \ar[u] \\
}
}

\subsection*{A Syntactic Theory of Sequential Control} 
Eugene Kohlbecker, Daniel Friedman, Matthias Felleisen and Bruce Duba 1987
\cite{felleisen87syntactic}

This paper lays out a semantics for higher order control based on
syntactic transformation of evaluation contexts.  Semantic analysis
with syntactic transformation was introduced primarily by Plotkin
\cite{gdp:cbn-cbv}, and evaluation contexts by Felleisen in
\cite{felleisen86secd}.  They present the control operators ${\mathcal
A}$ and ${\mathcal C}$, corresponding to {\tt abort} and {\tt
control}.  They then provide a context based semantics for these
operators, and prove a number of important properties, such as
Church-Rosser, for the new extended language, called $\lambda_C$.

This paper, referenced by many of the following papers in the area, lays
the groundwork for future work in the study of higher order control
operators.  It summarizes work published in a number of other settings
by the authors, places the previous results in a self-contained
context, and proves a number of key theorems about the system.

\subsection*{Control Delimiters and their Hierarchies}
Matthias Felleisen and Dorai Sitaram 1990 \cite{sitaram90control}

In this paper, Felleisen and Sitaram describe a new pair of control
operators, providing delimited continuations, first described in \cite{mf-prompts}.
Such continuations do not capture the entire rest of the computation,
but merely that back to an enclosing prompt.  Further, instead of
using the standard Scheme control operator {\tt call/cc}, they use the
operator $\mathcal{F}$, also previously defined by Felleisen.  This
operator is simpler, but provides all the power of the others.

The authors then provide a number of potential applications for such a
control operator, as well as implementation strategies.  Some of the
potential applications include coroutines, engines and implementations
of first-order control handlers such as exceptions.  Finally, the
authors discuss the possibilites of generating hierarchies of control
delimiters, so that continuations can be captured back to specific
delimiters, rather than the nearest enclosing delimiter.  They show
how this is easily expressible in terms of the original operators.

The sequence of papers culminating in this one introduced the delimited
continuation concept as a research topic.  Sitaram, Danvy and other
built on this work to create other, potentially more sophisticated
control operators.  However, the early papers by Felleisen and
collaborators represented their introduction.

\subsection*{Abstracting Control}
Olivier Danvy and Andre Filinski 1990 \cite{danvy90abstracting}

This paper is at once both very similar to, and very different from,
the previous paper.  It is similar in that it too introduces a new
control operator for delimited continutation, here called {\tt shift}
and {\tt reset}.  Further, shift and reset are very similar to the
previous control operators introduced by Sitaram and Felleisen.
Similar enough, in fact, that it takes serious effort to find programs
that distinguish them.

However, the paper is also very different.  Where the earlier paper
focused on implementation in Scheme, Danvy and Filinski focus on
semantics.  And where Felleisen and his collaborators focused on
operational semantics, this paper provides a more denotational focus,
defining the new control operators in terms of the CPS (and ECPS)
translation for the call-by-value $\lambda$-calculus.  However, this
paper also provides a number of examples of the control operators in
action, and provides a hierarchy of such operators.

{\tt shift} and {\tt reset} have seen significantly more use in actual
programming than the other delimited continutation operators.  For
example, Filinski uses them to implement a representation of monads
\cite{monads}, and Sperber and Gasbicher provide a direct
implementation in \cite{581504}.  However, they still have not
achieved even the popularity of {\tt call/cc}.

\subsection*{Continuations and Concurrency}
Robert Hieb and R. Kent Dybvig 1990 \cite{hieb90continuations}

This paper introduced a new delimited continuation operator, {\tt
spawn}, with significant differences from those seen previously.
First, this operator was introduced with the specific purpose of
handling concurrent process control.  However, the concurrency aspects
have been neglected in the further literature.  Also, instead of being
a pair of operators, there is only one.  The operator creates a new
control operator each time it is invoked, and passes this new operator
to its argument.  Further, the newly created control operator is
delimited by the particular {\tt spawn} where it was created.  This
new operator behaves identically to {\tt control} from
\cite{sitaram90control}, with the appropriate delimiter.

In addition to defining the new operator, the authors provide a
semantics for it.  To accomplish this, they add a labeling construct
to the standard context based reduction semantics seen earlier.  This
labeling construct allows jumps back to previously labeled points in
the program execution, saving and restoring the context.  While this
operator could have been useful for defining the semantics of control
operators, it has not seen further use.

\subsection*{A Dynamic Extent Control Operator for Partial
  Continuations}
Christian Queinnec and Bernard Serpette 1991 \cite{queinnec91dynamic}

In this paper, Queinnec and Serpette propose a control operator, {\tt
splitter}, very much like {\tt spawn}.  Again, it takes the form of a
delimiter which binds a new control operator each time it is invoked.
However, in this case, two new control operators are created.  One
reifies the current continuation back to the delimiter, but does not
involve aborting the current context.  The other merely aborts the
context back to the delimiter.  These are of course the two operations
performed together in all the other control operators.  Again, lexical
scoping plays a significant role, as it did for {\tt spawn}.
Unfortunately, no semantics are provided in the paper (although an
ACPS semantics is provided in a later survey paper \cite{queinnec93library}).

This paper represents a different direction from the main stream of
control operator research.  {\tt splitter} divides the two main
effects of continuation invocation in a way not seen elsewhere.  This
makes {\tt splitter}-created continutations even more ``composable''
than those created by {\tt shift}, which are often described as
``composable''.  Further, none of the other systems involve a
specialized delimited abort construct.  (Queinnec and Moreau devised a
similar control operator in \cite{luc94partial}).

The lexically scoped control operators are fundamentally different
from dynamically scoped operators.  It is not obvious which is the
correct choice.  \footnote{Note that this distinction is irrelevant
  for {\tt call/cc}, which is not scoped.}  Lexical scope has won out
for variable binding in all modern programming languages.  However,
exceptions in languages from C++ to SML are dynamically scoped.  This
suggests that higher order control operators be scoped similarly to
their first order cousins.  Finally, it is possible to provide a
purely functional implementation of lexically scoped control
operators in terms of dynamically scoped operators.  However, the
reverse is not the case.  These reasons together suggest that dynamic
scope is the correct choice.  However, in the abscence of extensive
programming experience with such operators, such a judgement can only
be provisional.

\subsection*{Handling Control}
Dorai Sitaram 1993 \cite{sitaram93handling}

This paper introduces yet another very different control operator,
{\tt fcontrol}, athough with a confusingly similar name to some
operators that we have seen before. Here, instead of the control
operator reifying the current continuation and passing that to its
argument, as all the previous operators did, the delimiter contains a
handler procedure.  When the control operator is invoked, control
passes to the handler, which recieves both the value supplied to the
control operator and the continuation at the point the operator was
invoked (back to the delimiter).  The handler is then free to do as it
wishes with these two arguments.

This is a radical departure from control operators previously
discussed here, but it is very like an exception handler.  When an
exception is raised, control passes to the dynamically nearest
enclosing handler, which is given the value of the exception.  Here,
the addition is to provide the current continuation of the exception
invocation as well.  This allows, for example, for resumption of the
computation with a different value.

This paper further provides a number of examples (including the
classic same-fringe) using the new operator.  In these examples, we
see that much of the computation acutally takes place in the handler,
rather than in the ``body'' of the code.  This is unlike normal
programming style with either exception handlers or standard control
operators.  This paper does not provide semantics for the new
operators, however such semantics are available in
\cite{sitaram94thesis}.

\bibliography{control}

\end{document}

### 2004-09-29 classes/hopl/logic/logic.tex
initial import

\documentclass[12pt]{article}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{fullpage}
\usepackage{enumerate}

\title{Origins of Logic Programming}
\author{Sam Tobin-Hochstadt}
\date{\today}

\begin{document}

\bibliographystyle{plain}

\maketitle

\subsection*{Predicate Logic as a Programming Language}
Kowalski, 1974 \cite{kow74}

In this paper, Kowalski lays the foundations for logic programming as
we know it.  The paper is still the standard reference for the ideas
behind logic programming.  The work in it was inspired by
collaboration with the authors of Prolog, Colmerauer and
Roussel. \cite{col93} He begins by laying out a syntax for logical
claims in the form of Horn clauses, and a way to compute with a set of
claims via refutation.

With the Horn clause syntax in place, he gives an operational
semantics for the language, which he describes as a ``procedural
interpretation''.  This semantics defines a top down method for
computing with a set of clauses and a goal statement, continuing until
a halt statement is reached.  For the language, the output at the end
is actually the substitution of the variables, since the final goal
statement is always the null statement.

With this semantics in place for predicate logic as a programming
language, Kowalski considers the other important issues.
Nondeterminism plays a major role, both in the multiple possibilities
for a successful substitution, and the non-deterministic choice of
procedure invocation.  Since a goal statement can have multiple
conjuncts, more than one of them could be reduced at any point.  Both
of these kinds of nondeterminism are discussed, and several options
for sequencing procedure invocations are discussed.  This can have
substantial performance impact.

\subsection*{Semantics of Predicate Logic as a Programming Language}
Van Emden and Kowalski, 1976 \cite{kow76}

The preceding paper on predicate logic as a programming language
contained a section on semantics, but these were far from formal,
especially by the standards of the day.  What that paper lacked, and
what this paper provides in quantity, is meaning functions.  Again,
this paper is foundational for the study of formal semantics of logic
programs.

The first meaning function (called ``denotation'' in the paper) is
defined by what is described as an operational semantics.  However,
this is not an operational semantics in the sense that we now think of
them, it is instead a set of values for which a predicate is true, for
a certain kind of truth.  This is how all the meaning functions are
defined.  For the operational meaning function $D_1$, the meaning is
given by provability, that is $X\vdash Y$ iff $X \wedge \overline{Y}$
is refutable.

The next semantics is a model theoretic one, which is given in terms
of Herbrand models of the set of clauses ${\mathbf A}$.  Here, a
predicate's denotation relative to ${\mathbf A}$ is given by those
terms which are true in all Herbrand models of ${\mathbf A}$.  Also, a
fixed-point semantics is given, where predicates are given denotations
by the least fixed point of the initial set of clauses ${\mathbf A}$.
Finally, the paper demonstrates all of these semantics to be
equivalent, and gives an interpretation of the operational semantics
in terms of Robinson's hyperresolution.

\subsection*{Algorithm = Logic + Control}
Kowalski, 1979 \cite{kow79}

By 1979, when this paper was published, logic programming had become a
significant part of the programming community.  However, there was as
yet still not very much understanding of the difference between how to
program in predicate logic versus the traditional methods of
programming in procedural languages.  In this paper, a different
understanding of algorithms is proposed, one which takes into account
the differences that separate logic from procedural programming.

The fundamental contribution of this paper is the equation $A = L +
C$, that is, that algorithms are the combination of logic and
control.  For Kowalski, logic is not specification but definition.  A
set of Horn clauses that make up a logic program is the logic of an
algorithm, as is the description of an algorithm that can be found in
a text book.  These concepts are familiar to procedural programmers.
However, control, the way in which the definitions are used to
compute, is not usually considered explicitly.

Further, algorithms do not necessarily have a unique decomposition
into logic and control.  If less is specified in the logic, more must
be left up to the control to decide.  For example, if the logic of a
program specifies quicksort, there is not much for the control to
decide.  It merely follows the procedures laid out in the logic.
However, if the logic simply consists of the statement that sorting
relates a list and an ordered permutation of that list, then the
control has a far more significant job to do.

Kowalski distinguishes two forms of control that are particularly
important for logic programs: bottom-up and top-down. Top down is the
form of control most familiar to programmers in non-logic languages.
It is also how the standard procedural interpretation of logic
programs computes.  However, there are cases where bottom up,
consisting of generating facts directly implied by constants and
assertions in the program, can actually compute faster and have other
desirable problems.  Kowalski analyzes a number of programs in this
context.

\subsection*{The Birth of Prolog}
Colmerauer and Roussel, 1993 \cite{col93}

This paper, while very interesting, is very much unlike a standard
computer science paper.  Instead, it is a survey of work that has been
done, not from the perspective of the work, but from the perspective
of the researchers.  The reader is thus given an inside look at the
development of logic programming.

The paper is divided into three stages in the history of Prolog.  It
begins with a natural language processing application developed in
Montreal, and continues through early Prolog into the final Prolog
much like that which we see 30 years later.

The original idea of Colmerauer was to develop as system that would
accept statements in French and then provide answers to natural
language queries such as ``What does Tom eat?'', to which the computer
might respond ``mice''.  This was implemented in terms of an
application called Q-systems, which were expressed as sets of
rewriting rules, written with variables.  The system then used
unification to find appropriate substitutions for the variables.

Having written the Q-systems, the group then focused on a system for
syntactic and semantic reasoning, which would allow them more power to
interpret and answer queries.  They settled on a predicate logic
formulation based on discussions with Robert Kowalski, who visited
Marseilles near that time and was in communication with Colmerauer and
Roussel.

The system they devised bears a close resemblance to the final Prolog,
including backtracking unification and cut.  However, this version
featured a number of added-on language features which would be removed
later.  For the final version, these features were removed and the
system was made more powerful generally, resulting in the Prolog we
know today.

\bibliography{logic}

\end{document}

### 2004-09-29 classes/theory/midterm.tex
initial import

\documentclass[12pt]
{article}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{enumerate}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{problem}[theorem]{Problem}

% General
\def\credittype{}
\def\RC{\credittype{RC}}
\def\EC{\credittype{EC}}
\def\NA{\credittype{NA}}

% Machines
\def\machine{\sf}
\def\mdfa{{\machine DFA}}
\def\mnfa{{\machine NFA}}
\def\mgnfa{{\machine GNFA}}
\def\mcfg{{\machine CFG}}
\def\mcfl{{\machine CFL}}
\def\mpda{{\machine PDA}}

% Grammar things
\newcommand\mygrammar[1]{
$$
\begin{array}{rcll}
#1
\end{array}
$$
}
\newcommand\gr[3]{\ensuremath{{\mit #1} &\rightarrow& #2 & ~~~~;~#3} \\}
\newcommand\gb[3]{\ensuremath{&|& #2 & ~~~~;~#3} \\}
\newcommand\dr[2]{\ensuremath{{\mit #1} &\Rightarrow& #2 &} \\}
\newcommand\db[1]{\ensuremath{&\Rightarrow& #1 &} \\}
\def\por{~|~}
\def\mt{\mathtt}

% Jay's 5 rules
\newcommand\jaarule[1]{\ensuremath{\mbox{{\sc JaysRule{#1}}}}}

% # in tt
\def\h{\ensuremath{\mt{\#}}}

% Graph stuff
\newcommand\pedge[3]{\ensuremath{#1,#2 \rightarrow #3}}
\def\pedgexxx{\pedge}
\newcommand\pedgeexx[2]{\pedge{\epsilon}{#1}{#2}}
\newcommand\pedgexex[2]{\pedge{#1}{\epsilon}{#2}}
\newcommand\pedgexxe[2]{\pedge{#1}{#2}{\epsilon}}
\newcommand\pedgexee[1]{\pedge{#1}{\epsilon}{\epsilon}}
\newcommand\pedgeexe[1]{\pedge{\epsilon}{#1}{\epsilon}}
\newcommand\pedgeeex[1]{\pedge{\epsilon}{\epsilon}{#1}}
\newcommand\pedgeeee[0]{\pedge{\epsilon}{\epsilon}{\epsilon}}

% Non-terminals
\newcommand\nterm[1]{\ensuremath{\langle\mbox{{\sc #1}}\rangle}}

\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\newcommand{\qed}{\nobreak \ifvmode \relax \else
      \ifdim\lastskip<1.5em \hskip-\lastskip
      \hskip1.5em plus0em minus0.5em \fi \nobreak
      \vrule height0.75em width0.5em depth0.25em\fi}

\title{CSG 713: Midterm}
\author{Sam Tobin-Hochstadt}
\date{\today}

\begin{document}

\maketitle

\begin{proof}[Problem 1] 45 points
\begin{enumerate}[i.]
\item $C - R$ is always context free, since $C-R = C \cap
  \overline{R}$, which is context free since complement preserves
  regularity, and intersection with a regular language preserves
  context freeness.
\item $R - C$ is not always regular, for example, $\Sigma^* -
  \overline{0^n1^n} = 0^n1^n$, which is not regular.
\item $N - R$ is not always non-regular, since $0^n1^n - \Sigma^*  =
  \{\}$, which is regular. 
\item $R - N$ is not always regular, since $\Sigma^* - 0^n1^n =
  \overline{0^n1^n}$, which is not regular.
\item $G - D$ is always recognizeable.  Given $M_G$ and $M_D$ which
  recognize the two languages, simply run $M_G$ on the input, and
  after it terminates, run $M_D$ on the input.  If both accept,
  accept.  Otherwise, reject.  This procedure will always terminate
  for all inputs in $G$, since $M_G$ terminates on those inputs, and
  $M_D$ terminates on all inputs.  
\item $D - G$ is not always decideable.  For example, $\Sigma^* -
  A_{TM} = \overline{A_{TM}}$ which is not even recognizeable.
\item Infinite unions of regular languages are not always regular.
  For example, $\displaystyle \bigcup_i a^ib^i = a^nb^n$ which is not
  regular, but all the individual sets are singleton strings, which
  are obviously regular.
\item The non-regular languages are not closed under concatention.
  For example, the language $A = a^p$ where $p$ is prime, is not
  regular.  However, $A$ concatenated with itself six times is $a^n$, where
  $n>11$, which is regular.
\end{enumerate}
\end{proof}

\begin{proof}[Problem 2] 30 points

\begin{enumerate}[i.]

\item Given a regular language $L$, we want to compute the language
  $Prefix(P)$.  To do this, consider $M_L$, an NFA which accepts
  $L$. First, add a state to $M_L$, an make it accepting.  Make all
  other states not accepting.  Then add $\epsilon$-transitions from
  all original states in $M_L$ from which a path existed to an
  accepting state to the new final state.

Claim:  This new NFA $M_L'$ accepts $Prefix(L)$.  Consider a string
$x$ in $Prefix(L)$.  By the definition of prefix, there exists a state
$p$ in $M_L$, which is the state the $M_L$ is in when it finishes
processing $x$.  By construction, $M_L'$ contains an
$\epsilon$-transition from $p$ to the accepting state.  Therefore,
$M_L'$ accepts $x$.  Thus, $L(M_L') \supseteq Prefix(L)$.

To prove that $L(M_L') \subseteq Prefix(L)$, consider a string
accepted by $M_L'$.  At the end of processing this string, the machine
must have taken an $\epsilon$-transition to the accepting state.
Consider the state the machine was in immediately preceeding this
transition.  By construction, there exists a path from this state to
an accepting state in $M_L$.  This path must be labeled by some string
$y$.  Then $xy \in L(M_L)$.  Therefore, $x \in Prefix(L)$.

\addtocounter{enumi}{1}
\item

\end{enumerate}

\end{proof}

\begin{proof}[Problem 3] 30 points

\end{proof}

\begin{proof}[Problem 4] 50 points

\begin{enumerate}[i.]
\item $L = 0^n1^m0^{n+m}$ is context-free but not regular.  It is
  generated by the following grammar:

\mygrammar{
\gr{S}{0S0 \por S'}{0^n1^m0^m0^n = 0^n1^m0^{n+m}}
\gr{S'}{1S'0 \por \epsilon}{1^n0^n}
}

Further, this language is clearly not regular.  Consider $L \cap
1^*0^* = 1^n0^n$. This is not regular, but $1^*0^*$ is.  So $L$ is not
regular.

\item $L = \{ww^Rx \por w \in \{0,1\}^+, x \in \{0,1\}^*\}$.  This
  language is context free, but not regular.  It is generated by the
  following grammar:

\mygrammar{
\gr{S}{1S'1X \por 0S'0X}{L}
\gr{S'}{1S'1 \por 0S'0 \por \epsilon}{ww^R}
\gr{X}{1X \por 0X \por \epsilon}{\Sigma^*}
}

$L$ is clearly not regular.  Consider $L \cap \{xy \por |x| = |y|\}$.
Let this language be $L'$.  If $L$ was regular, $L'$ would be as
well.  However, $FirstHalf(L') = ww^R$, which is not regular.
Therefore, $L$ is not regular.

\item   This language is decideable but not context-free.

\item This language is regular.

\end{enumerate}

\end{proof}

\begin{proof}[Problem 5] 30 points

The following grammar solves part (2).

\mygrammar{
\gr{S}{S^* \por T}{}
\gr{T}{TR \por R}{}
\gr{R}{R \cup P \por P}{}
\gr{P}{a \por b \por (S) \por \epsilon \por \emptyset}{}
}

\end{proof}

\begin{proof}[Problem 6] 30 points

If $L_1$ and $L_2$ are recognizeable, and both $L_1 \cup L_2$ and $L_1 \cap
L_2$ are decideable, then $L_1$ and $L_2$ are decideable.

Proof: Consider the following four machines: $A,B,C,D$, where $L(A) =
L_1, L(B) = L_2, L(C) = L_1 \cup L_2, L(D) = L_1 \cap L_2$.  Then $A$
and $B$ are recognizers, and $C$ and $D$ are deciders.  Then we can
construct the following machine which decides $L_1$.  The construction
for $L_2$ is analagous.

Given an input string $x$, run $C$ on $x$.  If $C$ rejects, halt and
reject.  If $C$ accepts, then run $D$ on $x$. If $D$ accepts, halt and
accept.  If $D$ rejects, then run $A$ and $B$ in parallel
\footnote{Meaning, take each machine one step each time, alternating
machines.}  on $x$ until one of them accepts.  If $A$ halts and
accepts, accept.  If $B$ halts and accepts, reject.  Since $C$
accepted, $x \in L_1$ or $x \in L_2$.  Since both
machines are recognizers, therefore, one of them must accept $x$.
Therefore, this procedure always finishes.

\end{proof}

\end{document}

### 2004-09-29 classes/theory/ps1.tex
initial import

\documentclass[12pt]
{article}

\usepackage{amssymb}
\usepackage{amsmath}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{problem}[theorem]{Problem}

\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\newcommand{\qed}{\nobreak \ifvmode \relax \else
      \ifdim\lastskip<1.5em \hskip-\lastskip
      \hskip1.5em plus0em minus0.5em \fi \nobreak
      \vrule height0.75em width0.5em depth0.25em\fi}

\title{CSG 713: Problem Set 1}
\author{Sam Tobin-Hochstadt}
\date{\today}

\begin{document}

\maketitle

\subsection*{Problem 1}

Consider the regular language $M$, accepted by a DFA $D_M =
(Q,\Sigma,\delta,q_0,F)$. Then let NFA $N_{M^R} =
(Q \cup \{q_0'\},\Sigma,\delta',q_0',\{q_0\})$, where $q_0'$ is a new
state and
$\delta' = \{ (m,n,a) ~|~ (n,m,a) \in \delta \} \cup \{
(q_0',f,\epsilon) ~|~ f \in F\}$.

To see that these two automata accept the same language, consider what
happens to each when processing a string.  For each string $w$, if there
is a path from the start state to a final state labeled with that
string in the DFA, then there must be a path from one of the states in
$F$ to $q_0$ labeled with $w^R$ in the NFA.  By construction,
therefore, there is path labeled with $\epsilon w^R$ from the start
state to a final state in the NFA.  Therefore, all strings in $W^R$
are accepted by $N_{M^R}$.

To see that all strings accepted by $N_{M^R}$ are in $W^R$, consider
what would happen if we applied the construction to a DFA with only
one start state, and added no epsilon transitions.  Then, if we
applied the construction again, we would get the original DFA.
Therefore, since the final automata accepts every string in the
reverse of the reverse of the first automata, as proved above, since
it clearly accepts no more strings than the first automata (since they
are isomorphic), the construction must not add any strings to the
language.

Finally, since NFAs can be made to have single start states without
loss of generality, and the construction also works for NFAs, the two
automata accept the same language. \qed

\subsection*{Problem 2}

\subsubsection*{Part A}

We solve this problem by proving that $C^R$ is regular.

For this problem, we need merely to keep track of whether there is a
carry bit or not, which means we need two states.  If the top is equal
to the bottom, plus a carry, then we move the the carry state.  If the
top plus the carry is equal to the bottom, we move the even state.  If
the top plus the carry is equal to the bottom plus a carry, we stay in
the carry state.  If the top is equal to the bottom, and there is no
carry, we stay in the same state.  Otherwise the machine fails.

This is implemented in an attached machine.

\subsubsection*{Part B}
Since $0 * 3^n$ and $2 * 3^n$ are even for all values of $n$, and $1 *
3^n$ is odd for all positive values of $n$, we merely need to
determine if the number of 1's in the string is even.  See the
attached figure for the automaton.

\subsection*{Problem 3}

\subsubsection*{Part A \small{I discussed this problem with Richard Cobbe}}

First, we show that multiplication by two yields a regular language.
This is trivial, since to multiply by two in binary we need merely to
shift left by one position, and compare for equality.  To do this with
an automata, we need merely to require an initial zero in the top row
when examining the string in reverse.

Since we also already are able to check for equality between the sum
of two rows and a third (Problem 2.A), we need merely to show that this combination
preserves regularity. To do this, consider a DFA that checks for
equality, and one that checks for being twice the size, when a zero is
added (this is multiplication by two).  If both accept, then accept.
This is possible because regular languages are closed under
intersection.

\subsubsection*{Part B}

For this problem, we need merely to see which row has a one where the
other has a zero first.  That row is the larger.  The automaton
implementing this is attached.

\subsection*{Problem 4}

First, note that if $(Q,\Sigma,\delta,q_0,F)$ is an NFA which accepts
language $M$, then it can also be considered as an all-paths-NFA,
accepting language $M'$.  To construct a DFA that accepts the same
language, we follow the construction for simulating an NFA with a
DFA.  However, we add one additional state, a dead state, with edges
from itself to itself on all inputs.  We add an edge to this state for
each time when the NFA would be unable to make a transition.  Further,
we restrict the accepting states to those subsets of $Q$ which are
also subsets of $F$.

This construction correctly simulates an all-paths-NFA for the same
reason that the construction works for NFAs.  If the NFA never
encounters a state where it is unable to make a transitions, then it
ends up in the state representing all possible ending states (this is
identical to the NFA case).  If all of those states are accepting, the
DFA will accept, which is the correct behavior for the
all-paths-NFA. If the all-paths-NFA ever reached a point where it was
unable to make a transition, the DFA would go to the dead state, and
reject, which is also the correct behavior.

Second, note that every DFA is an all-paths-NFA, with no
nondeterminism.  Therefore, any language that can be accepted by a DFA
can also be accepted by an all-paths-NFA.  Therefore, all-paths-NFAs
can accepts all regular languages.

From these two facts, all-paths-NFAs accepts precisely the regular
languages.  \qed

\subsection*{Problem 5 \small{I worked on this problem with Greg Pettyjohn}}
Consider a DFA $M$ which accepts language $A$.  Construct a new DFA
$M'$ by changing $F$ to $\overline{F}$.  This DFA accepts all proper
prefixes of $A$.  Call this operation $PP(A)$.

\subsubsection*{Part A}
$NOPREFIX(A) = A - PP(A)$, therefore it is regular, since the regular
languages are closed under set difference.

\subsubsection*{Part B}
$NOPREFIX(A) = \overline{PP(A)}$, therefore it is regular, since the regular
languages are closed under set complement.

\subsection*{Problem 6}

To prove that $Multiple_5$ is regular, we prove the $Multiple_5^R$ is
regular.  To do this, we construct an automaton.

Let $Q = \{0,1,2,3,4\} x \{1,2,4,3\}$.

Let $q_0 = (0,1)$.

Let $F = \{(0,x)~|~ x \in \{1,2,4,3\}\}$.

Let $\Sigma = \{0,1\}.$

Finally, we must construct the transition function.  For all states
labelled $(x,1)$, add an edge labeled 0 to the state $(x,2)$.
Similarly from $(x,2)$ to $(x,4)$ and from $(x,4)$ to $(x,3)$ and
finally from $(x,3)$ back to $(x,1)$.

Then add an edge for each state labeled $(m,n)$, add an edge labeled 1
to the state $(m+n\bmod 5, f(n))$, where $f$ takes 1 to 2, 2 to 4, 4
to 3 and 3 to 1, as in the above paragraph.

This automaton keeps track both of the number modulo 5, and how many
to add to the number if the next digit is a one.  Since the powers of
2 modulo 5 repeat the sequence $1,2,4,3$, we have a total of 20
states.  If the computation ends with the current modulus at 0, the
machine accepts.

\subsection*{Problem 7}

\subsubsection*{Part A}
The automaton is equivalent to the RE $((0 \cup 1)^*1) \cup ((0 \cup
1)^*1(00)^*)$.

\subsubsection*{Part B}
The automaton is equivalent to the RE $(b^*a) (((a \cup b)a) \cup ((a
\cup b)bb^*a))^*$.

\end{document}

### 2004-09-29 classes/theory/ps2.tex
initial import

\documentclass[12pt]
{article}

\usepackage{amssymb}
\usepackage{amsmath}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{problem}[theorem]{Problem}

% General
\def\credittype{}
\def\RC{\credittype{RC}}
\def\EC{\credittype{EC}}
\def\NA{\credittype{NA}}

% Machines
\def\machine{\sf}
\def\mdfa{{\machine DFA}}
\def\mnfa{{\machine NFA}}
\def\mgnfa{{\machine GNFA}}
\def\mcfg{{\machine CFG}}
\def\mcfl{{\machine CFL}}
\def\mpda{{\machine PDA}}

% Grammar things
\newcommand\mygrammar[1]{
$$
\begin{array}{rcll}
#1
\end{array}
$$
}
\newcommand\gr[3]{\ensuremath{{\mit #1} &\rightarrow& #2 & ~~~~;~#3} \\}
\newcommand\gb[3]{\ensuremath{&|& #2 & ~~~~;~#3} \\}
\newcommand\dr[2]{\ensuremath{{\mit #1} &\Rightarrow& #2 &} \\}
\newcommand\db[1]{\ensuremath{&\Rightarrow& #1 &} \\}
\def\por{~|~}
\def\mt{\mathtt}

% Jay's 5 rules
\newcommand\jaarule[1]{\ensuremath{\mbox{{\sc JaysRule{#1}}}}}

% # in tt
\def\h{\ensuremath{\mt{\#}}}

% Graph stuff
\newcommand\pedge[3]{\ensuremath{#1,#2 \rightarrow #3}}
\def\pedgexxx{\pedge}
\newcommand\pedgeexx[2]{\pedge{\epsilon}{#1}{#2}}
\newcommand\pedgexex[2]{\pedge{#1}{\epsilon}{#2}}
\newcommand\pedgexxe[2]{\pedge{#1}{#2}{\epsilon}}
\newcommand\pedgexee[1]{\pedge{#1}{\epsilon}{\epsilon}}
\newcommand\pedgeexe[1]{\pedge{\epsilon}{#1}{\epsilon}}
\newcommand\pedgeeex[1]{\pedge{\epsilon}{\epsilon}{#1}}
\newcommand\pedgeeee[0]{\pedge{\epsilon}{\epsilon}{\epsilon}}

% Non-terminals
\newcommand\nterm[1]{\ensuremath{\langle\mbox{{\sc #1}}\rangle}}

% non-terminals for 2.8
\def\NP{\nterm{NOUN-PHRASE}}
\def\VP{\nterm{VERB-PHRASE}}
\def\CN{\nterm{CMPLX-NOUN}}
\def\CV{\nterm{CMPLX-VERB}}
\def\AR{\nterm{ARTICLE}}
\def\N{\nterm{NOUN}}
\def\V{\nterm{VERB}}
\def\PP{\nterm{PREP-PHRASE}}
\def\P{\nterm{PREP}}
\def\SEN{\nterm{SENTENCE}}

% 2.21
\def\STMT{\nterm{STMT}}
\def\STMTLIST{\nterm{STMT-LIST}}
\def\IT{\nterm{IF-THEN}}
\def\A{\nterm{ASSIGN}}
\def\ITE{\nterm{IF-THEN-ELSE}}
\def\NIF{\nterm{NON-IF}}
\def\IF{\nterm{IF}}
\def\NEST{\nterm{NESTED-IF}}
\def\NNEST{\nterm{NON-NESTED-IF}}
\def\BEGIN{\nterm{BEGIN-END}}

%--
%Jeffrey Palm --> http://jeffpalm.com

\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\newcommand{\qed}{\nobreak \ifvmode \relax \else
      \ifdim\lastskip<1.5em \hskip-\lastskip
      \hskip1.5em plus0em minus0.5em \fi \nobreak
      \vrule height0.75em width0.5em depth0.25em\fi}

\title{CSG 713: Problem Set 3}
\author{Sam Tobin-Hochstadt}
\date{\today}

\begin{document}

\maketitle

For all grammars, $S$ is the start symbol.  Jeff Palm supplied the
\LaTeX{} macros that made this typesetting possible.

\subsection*{Problem 1}

\subsubsection*{Part A}

\mygrammar{  
  \gr{S}{aSb  \por aSbb \por \epsilon}{a^i b^j ~with~ i \leq j \leq 2i}  
}

\subsubsection*{Part B}

\mygrammar{
  \gr{R}{aRa \por bRb \por \epsilon}{x \in \{a,b\}^*, x = x^R}
  \gr{R_2}{aR_2a \por bR_2b \por \h A\h}{x\h A^*\h x^R}
  \gr{A}{Aa \por Ab \por A\h \por \epsilon}{\{a,b,\h\}^*}
  \gr{S_1}{A\h R_2\h A \por R_2}{A^*\h R_2\h A^*}
  \gr{S}{R \por S_1}{S_1 ~or ~a ~palindrome}
}

\subsection*{Problem 2}

See PDA diagrams on attached sheet.

\subsection*{Problem 3}
Twice as many a's as b's.
 \mygrammar{
   \gr{S}{ S_1 \por S_2 \por S_3 \epsilon }{}
   \gr{S_1}{Saba \por aSba \por abSa \por abaS}{}
   \gr{S_2}{Saab \por aSab \por aaSb \por aabS}{}
   \gr{S_3}{Sbaa \por bSaa \por baSa \por baaS}{}
}

\subsection*{Problem 4}
The complement of $a^nb^n$.

\mygrammar{
  \gr{S}{X \por Y}{a^nb^n~ complement}
  \gr{X}{aXb \por A \por B}{a^nb^m, m \neq n}
  \gr{A}{Aa \por a}{a^*}
  \gr{B}{Bb \por b}{b^*}
  \gr{Y}{ZbZaZ}{a^*b^* ~complement}
  \gr{Z}{Za \por Zb \por \epsilon}{\{a,b\}^*}
}

\newpage

\subsection*{Problem 5}
Two different derivations of ``the girl touches the boy with the
flower''.

\subsubsection*{First Derivation}
Meaning: The girl uses the flower to touch the boy.

\mygrammar{  
  \dr{\SEN}{\NP \; \VP}  
  \db{\CN \; \VP}  
  \db{\AR \; \N \; \VP}  
  \db{\mt{the} \; \N \; \VP}  
  \db{\mt{the} \; \mt{girl} \; \VP}  
  \db{\mt{the} \; \mt{girl} \; \CV \; \PP}
  \db{\mt{the} \; \mt{girl} \; \V \;\NP\; \PP}
  \db{\mt{the} \; \mt{girl} \; \mt{touches} \;\NP\; \PP}
  \db{\mt{the} \; \mt{girl} \; \mt{touches} \;\AR\;\N\; \PP}
  \db{\mt{the} \; \mt{girl} \; \mt{touches} \;\mt{the}\;\N\; \PP}
  \db{\mt{the} \; \mt{girl} \; \mt{touches} \;\mt{the}\;\mt{boy}\; \PP}
  \db{\mt{the} \; \mt{girl} \; \mt{touches} \;\mt{the}\;\mt{boy}\; \P\;\NP}
  \db{\mt{the} \; \mt{girl} \; \mt{touches} \;\mt{the}\;\mt{boy}\; \mt{with}\;\NP}
  \db{\mt{the} \; \mt{girl} \; \mt{touches} \;\mt{the}\;\mt{boy}\; \mt{with}\;\AR\;\N}
  \db{\mt{the} \; \mt{girl} \; \mt{touches} \;\mt{the}\;\mt{boy}\; \mt{with}\;\mt{the}\;\N}
  \db{\mt{the} \; \mt{girl} \; \mt{touches} \;\mt{the}\;\mt{boy}\; \mt{with}\;\mt{the}\;\mt{flower}}
}

\newpage
\subsubsection*{Second Derivation}
Meaning: The girl touches the boy who has the flower.

\mygrammar{  
  \dr{\SEN}{\NP \; \VP}  
  \db{\CN \; \VP}  
  \db{\AR \; \N \; \VP}  
  \db{\mt{the} \; \N \; \VP}  
  \db{\mt{the} \; \mt{girl} \; \VP}  
  \db{\mt{the} \; \mt{girl} \; \CV}
  \db{\mt{the} \; \mt{girl} \; \V\;\NP}
  \db{\mt{the} \; \mt{girl} \; \mt{touches}\;\NP}
  \db{\mt{the} \; \mt{girl} \; \mt{touches}\;\CN\;\PP}
  \db{\mt{the} \; \mt{girl} \; \mt{touches}\;\AR\;\N\;\PP}
  \db{\mt{the} \; \mt{girl} \; \mt{touches}\;\mt{the}\;\N\;\PP}
  \db{\mt{the} \; \mt{girl} \; \mt{touches}\;\mt{the}\;\mt{boy}\;\PP}
  \db{\mt{the} \; \mt{girl} \; \mt{touches}\;\mt{the}\;\mt{boy}\;\P\;\CN}
  \db{\mt{the} \; \mt{girl} \; \mt{touches}\;\mt{the}\;\mt{boy}\;\mt{with}\;\CN}
  \db{\mt{the} \; \mt{girl} \; \mt{touches}\;\mt{the}\;\mt{boy}\;\mt{with}\;\AR\;\N}
  \db{\mt{the} \; \mt{girl} \; \mt{touches}\;\mt{the}\;\mt{boy}\;\mt{with}\;\mt{the}\;\N}
  \db{\mt{the} \; \mt{girl} \; \mt{touches}\;\mt{the}\;\mt{boy}\;\mt{with}\;\mt{the}\;\mt{flower}}

}

\subsection*{Problem 6}

The following string has two derivations:

$\mt{if ~condition~ then} \; \mt{if ~condition~ then} \; \mt{a:=1} \; \mt{else}\; \mt{a:=1}$

\subsubsection*{First derivation}

\mygrammar{
\dr{\STMT}{\IT}
\db{\mt{if ~condition~ then} \; \STMT}
\db{\mt{if ~condition~ then} \; \ITE}
\db{\mt{if ~condition~ then} \; \mt{if ~condition~ then} \; \STMT \; \mt{else}\; \STMT}
\db{\mt{if ~condition~ then} \; \mt{if ~condition~ then} \; \A \; \mt{else}\; \STMT}
\db{\mt{if ~condition~ then} \; \mt{if ~condition~ then} \; \mt{a:=1} \; \mt{else}\; \STMT}
\db{\mt{if ~condition~ then} \; \mt{if ~condition~ then} \; \mt{a:=1} \; \mt{else}\; \A}
\db{\mt{if ~condition~ then} \; \mt{if ~condition~ then} \; \mt{a:=1} \; \mt{else}\; \mt{a:=1}}
}

\subsubsection*{Second derivation}

\mygrammar{
\dr{\STMT}{\ITE}
\db{\mt{if ~condition~ then} \; \STMT \; \mt{else}\; \STMT}
\db{\mt{if ~condition~ then} \; \IT\; \mt{else}\; \STMT}
\db{\mt{if ~condition~ then} \; \mt{if ~condition~ then} \; \STMT \; \mt{else}\; \STMT}
\db{\mt{if ~condition~ then} \; \mt{if ~condition~ then} \; \A \; \mt{else}\; \STMT}
\db{\mt{if ~condition~ then} \; \mt{if ~condition~ then} \; \mt{a:=1} \; \mt{else}\; \STMT}
\db{\mt{if ~condition~ then} \; \mt{if ~condition~ then} \; \mt{a:=1} \; \mt{else}\; \A}
\db{\mt{if ~condition~ then} \; \mt{if ~condition~ then} \; \mt{a:=1} \; \mt{else}\; \mt{a:=1}}
}

\subsection*{Problem 7}

An unabiguous grammar for statements.

\mygrammar{
\gr{S}{ITE \por S'}{}
\gr{S'}{IT \por A \por BE}{}
\gr{A}{\mt{a:=1}}{}
\gr{BE}{\mt{begin~} \; SL \; \mt{~end}}{}
\gr{SL}{SL \; S \por S}{}
\gr{IT}{\mt{if ~condition~} \; S \mt{~then~} S'}{}
\gr{IT}{\mt{if ~condition~} \; S \mt{~then~} S \mt{~else~} S}{}
}

\subsection*{Problem 8}

\mygrammar{
  \gr{S}{0S0 \por 1S1 \por 0S'1 \por 0S'0}{x\h y, |x| = |y|, x \neq y}
  \gr{S'}{AS'A \por \h}{x\h y, |x| = |y|}
  \gr{A}{1 \por 0}{1 ~or~ 0}

}

\subsection*{Problem 9}

\mygrammar{
  \gr{S}{0S0 \por 1S1 \por 0S'1 \por 0S'0}{xy, |x| = |y|, x \neq y}
  \gr{S'}{AS'A \por \epsilon}{xy, |x| = |y|}
  \gr{A}{1 \por 0}{1 ~or~ 0}

}

\end{document}

### 2004-09-29 classes/theory/ps4.tex
initial import

\documentclass[12pt]
{article}

\usepackage{amssymb}
\usepackage{amsmath}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{problem}[theorem]{Problem}

% General
\def\credittype{}
\def\RC{\credittype{RC}}
\def\EC{\credittype{EC}}
\def\NA{\credittype{NA}}

\newcommand\prob[1]{\noindent{\em #1 }\vspace{2mm}}

% Machines
\def\machine{\sf}
\def\mdfa{{\machine DFA}}
\def\mnfa{{\machine NFA}}
\def\mgnfa{{\machine GNFA}}
\def\mcfg{{\machine CFG}}
\def\mcfl{{\machine CFL}}
\def\mpda{{\machine PDA}}

% Grammar things
\newcommand\mygrammar[1]{
$$
\begin{array}{rcll}
#1
\end{array}
$$
}
\newcommand\gr[3]{\ensuremath{{\mit #1} &\rightarrow& #2 & ~~~~;~#3} \\}
\newcommand\gb[3]{\ensuremath{&|& #2 & ~~~~;~#3} \\}
\newcommand\dr[2]{\ensuremath{{\mit #1} &\Rightarrow& #2 &} \\}
\newcommand\db[1]{\ensuremath{&\Rightarrow& #1 &} \\}
\def\por{~|~}
\def\mt{\mathtt}

% Jay's 5 rules
\newcommand\jaarule[1]{\ensuremath{\mbox{{\sc JaysRule{#1}}}}}

% # in tt
\def\h{\ensuremath{\mt{\#}}}

% Graph stuff
\newcommand\pedge[3]{\ensuremath{#1,#2 \rightarrow #3}}
\def\pedgexxx{\pedge}
\newcommand\pedgeexx[2]{\pedge{\epsilon}{#1}{#2}}
\newcommand\pedgexex[2]{\pedge{#1}{\epsilon}{#2}}
\newcommand\pedgexxe[2]{\pedge{#1}{#2}{\epsilon}}
\newcommand\pedgexee[1]{\pedge{#1}{\epsilon}{\epsilon}}
\newcommand\pedgeexe[1]{\pedge{\epsilon}{#1}{\epsilon}}
\newcommand\pedgeeex[1]{\pedge{\epsilon}{\epsilon}{#1}}
\newcommand\pedgeeee[0]{\pedge{\epsilon}{\epsilon}{\epsilon}}

% Non-terminals
\newcommand\nterm[1]{\ensuremath{\langle\mbox{{\sc #1}}\rangle}}

\newcommand\sr[2]{\ensuremath{{\mit #1}  \rightarrow #2}}

\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\newcommand{\qed}{\nobreak \ifvmode \relax \else
      \ifdim\lastskip<1.5em \hskip-\lastskip
      \hskip1.5em plus0em minus0.5em \fi \nobreak
      \vrule height0.75em width0.5em depth0.25em\fi}

\title{CSG 713: Problem Set 4}
\author{Sam Tobin-Hochstadt}
\date{\today}

\begin{document}

\maketitle

\subsection*{Problem 1}

\prob{Prove that the context-free languages are closed under reverse.}

Consider a language $A$ that is context-free.  Then there exists a
Chomsky normal form grammar $G$, such that $L(G) = A$. Then the
language $Reverse(A)$ is generated by the grammar $G'$, defined as
follows:

For all productions of the form \sr{A}{BC}, replace them with
productions \sr{A}{CB}.  For productions of the for \sr{A}{a}, leave
them unchanged.

{\em Claim:} If a parse tree starting at non-terminal $N$ was
generated by $G$, then the mirror image of that parse tree, starting
at the same nonterminal, is generated by $G'$.

Proof by induction on the height of the parse tree.

Base Case: If the height is 1, then $w$ was generated by a single
production from the start symbol to a terminal.  By construction, that
production is in $G'$, so the parse tree can be generated by $G'$.

Inductive Step.  Consider the root of the tree.  If we are not in the
base case, then the start symbol generates exactly two trees.  By the
induction hypothesis, both the left and the right subtrees are the
mirror images of trees produced by $G$.  Therefore, since the
production at the root of the tree is of the form \sr{A}{CB}, there
must have been a production in $G$ of the form \sr{A}{BC}.  Therefore,
G must have produced a tree with the same root, and with mirrored
subtrees, but with the left and right subtrees reversed.  This is
precisely a mirror image of the current tree.  \qed

Now consider those trees which have the start symbol at the root, and
which have terminals at all the leaves.  These are the trees that
generate some string $w$, where $w \in L(G)$.  By the above claim,
the mirror image tree must be generated by $G'$.  This mirror tree
generates precisely the string $w^R$.  Therefore, if $w \in L(G) = A$,
then $w^R \in L(G')$.  So $A^R \subseteq L(G')$.

To prove that $L(G') = A^R$, consider the grammar $G''$, obtained by
applying the original construction to $G'$.  Then we have ${A^R}^R
\subseteq L(G'')$, or $A \subseteq L(G'')$. But $G'' = G$, so no new
strings have been added.  Therefore,  $L(G') = A^R$.

\subsection*{Problem 2}

\prob{Sipser 2.18 (a,c)}

\subsubsection*{Part A}

Let $A = \{0^n1^n0^n1^n~|~n\geq 0\}$.  We use the pumping lemma to
show that $A$ is not context-free.

Assume $A$ is context free.  Let $p$ be the pumping length given by
the pumping lemma.  Choose $s = 0^p1^p0^p1^p$.  Let $s = uvxyz$ and
consider the following cases.

\begin{enumerate}
  \item If either $v$ or $y$ contain more than one type of symbol,
  then $uv^2xy^2z \not\in 0^*1^*0^*1^*$.

\item Since both $v$ and $y$ contain one type of symbol, each must be
  entirely contained with in one for the four blocks in $s$.  Then
  there must be two blocks left out.  Then consider $s' = uv^0xy^0z$.  This
  will have fewer symbols in at most two of the blocks.  Therefore,
  there will be at least two blocks whose lengths do not change, and
  at least one which is shorter.  Therefore, $s' \not\in A$.  
\end{enumerate}

\subsubsection*{Part B}

Let $A = \{w\h x~|~w$ is a substring of $x \}$.  We use the pumping
lemma to show that $A$ is not context-free.

Assume $A$ is context free.  Let $p$ be the pumping length given by
the pumping lemma.  Choose $s = 0^p1^p\h 0^p1^p$.  Let $s = uvxyz$ and
consider the following cases.

\begin{enumerate}
\item If either $v$ or $y$ contains a \h, then $uv^0xy^0z$ does not,
  and is not in $A$.

\item If both $v$ and $y$ come before the \h, then $uv^2xy^2z$ has a
  longer substring before the \h than after, and thus is not in $A$.

\item If both $v$ and $y$ come after the \h, then $uv^0xy^0z$ has a
  longer substring before the \h than after, and thus is not in $A$.

\item If $v$ is before the \h, and $y$ is after, then $v = 1^k$ and $y
  = 0^l$, for some $k,l$ not both 0.  Then if $k = 0$, $uv^0xy^0z$ has
  a longer substring before the \h than after, and thus is not in
  $A$.  On the other hand, if $k \neq 0$, $uv^2xy^2z$ has more 1s
  before the \h than after, and thus $w$ cannot be a substring of
  $x$.  Therefore, it is not in $A$.  
\end{enumerate}

Since these cases are exhaustive, $A$ is not context-free.

\subsection*{Problem 3}

Let $A = \{0^n 1^m~|~ m = n^2 \}$.  We use the pumping
lemma to show that $A$ is not context-free.

Assume $A$ is context free.  Let $p$ be the pumping length given by
the pumping lemma.  Choose $s = 0^{p^2}1^p$.  Let $s = uvxyz$ and
consider the following cases.

\begin{enumerate}
\item If either $v$ or $y$ contains both 0s and 1s, then $uv^2xy^2z$
  is not of the form $0^*1^*$, and is thus not in $A$.
\item If $v$ and $y$ consist of either both 1s or both 0s, then
  $uv^2xy^2z$ is not in $A$, since $|vy| \neq 0$, and therefore $p^2
  \neq p^2 + |vy|$, and $(p+|vy|)^2 \neq p^2$.

\item If $v$ consists of 0s, and $y$ consists of 1s, then either
  $uv^2xy^2z$ or $uv^3xy^3z$ is not in $A$, since this would require
  that $p^2+|v| = (p+|y|)^2$ and $p^2+2*|v| = (p+2*|y|)^2$, which
  implies that $|v| = |y|^2$, and $|v| = 2*|y|^2$, which requires that
  $|vy| = 0$.

\end{enumerate}

Since these cases are exhaustive, $A$ is not context-free.

\subsection*{Problem 4}

Let $A = \{ww^Rw~|~ w \in \{0,1\}^* \}$.  We use the pumping
lemma to show that $A$ is not context-free.

Assume $A$ is context free.  Let $p$ be the pumping length given by
the pumping lemma.  Choose $s = 0^p1^p1^p0^p 0^p1^p$.  Let $s = uvxyz$ and
consider the following cases.

\begin{enumerate}
\item If either  $v$ or $y$ contains both 0s and 1s, then $uv^2xy^2z$
  will not be in $A$.

\item Since $|vxy| \leq p$, either $vxy$ consists of only 1s or only
  0s, or $vxy$ is wholly contained within one of the thirds of the
  original string.  If the latter is true, then $uv^2xy^2z$ will
  increase the length of only one of the sections, and the new string
  will not be in $A$.  If the former is true, then there will be more
  of the symbol in $vxy$ in whichever half of $s$ that substring appears in
  than in the other half of $s$.  Therefore, the new string will not
  have the form $ww^Rw$, and thus not be in $A$.

\end{enumerate}

Since these cases are exhaustive, $A$ is not context-free.

\subsection*{Problem 5}

Let $A = \{a^nb^nc^m~|~ m \neq n \}$.  To show that $A$ satisfies the
requirements of the strong version of the pumping lemma, we need to
show that there exists an integer $k$, such that for any string $s \in
A$, $s = uvxyz$, with $|vy| > 0$, $|vxy| \leq k$, and $\forall n
\geq 0~uv^nxy^nz \in A$.

To show this, choose an integer $k > 10$.  Then, given $s = a^nb^nc^m$,
consider two cases.
\begin{enumerate}
\item If $n > m$, choose $v$ and $y$ to be equal length strings, with
  $v$ consisting of a's, and $y$ consisting of b's so that $|n - m|
  \neq |v|$.  Then $\forall n \geq0 ~uv^nxy^nz \in A$.

\item If $m > n$, then choose $v$ to be empty, and $y$ to be a string
  of c's so that $|n - m| \neq |y|$.  Further, choose $u$ to be the
  rest of the string.  Then $uy^n \in A$ for any $n$.

\end{enumerate}

Therefore, for any string $s$, it can be broken into 5 parts so that
the conditions are met.

\end{document}

### 2004-09-29 metaclasses/paper.tex
initial import

\documentclass{acmconfbig}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}

\renewcommand{\textfraction}{0}
\renewcommand{\topfraction}{1}
\renewcommand{\bottomfraction}{1}
\renewcommand{\floatpagefraction}{.4}
\setcounter{totalnumber}{4}
\setcounter{bottomnumber}{2}

\input{semantics-macros.tex}
\input{semantics-rules.tex}
\input{semantics-proofs.tex}
\input{paper-figs.tex}

\begin{document}

\title{A Core Calculus of Metaclasses}

\author{Sam Tobin-Hochstadt \quad Eric Allen \\
\begin{tabular}{c}
\\
Sun Microsystems Laboratories \\
Burlington MA 01803 \\
{\tt \large <first>.<last>@sun.com} \\
\end{tabular}
}

\maketitle

\toappear{Submission for FOOL 2005}

\begin{abstract}
Metaclasses provide a useful method of abstraction for programmers
working with object-oriented languages, but they have not seen the
formal exploration applied to more conventional object-oriented
programming features.  In order to elucidate the structure of
metaclasses and their relationship with static typing, we present a
core calculus for a nominally-typed object-oriented language with
metaclasses and prove type soundness over this core.  This calculus is
presented as an adaptation of Featherweight GJ \cite{FJ}, and is
powerful enough to capture metaclass relationships beyond what are
expressible in common object-oriented languages, including arbitrary
metaclass hierarchies and classes as values.  We also explore the
integration of object-oriented design patterns with metaclasses, and
show how our calculus integrates several such patterns cleanly.
\end{abstract}

\section{Introduction}
One of the stated benefits of object-oriented languages is their
ability to model aspects of the world in the object hierarchy.
However, most such languages are unable to model many simple and
common relationships. For example, consider the relationship between
physical quantities, units of measurement, and physical dimensions. It
is obvious that ``length'' is a dimension, and that ``3 feet'' is a
length.  However, modeling this relationship in a conventional
language such as the Java${^{\mbox{\tiny{TM}}}}$ Programming Language
is impossible without attributing to some of these concepts properties
which do not properly belong to them.  It is natural to model {\txt 3
feet} as an instance of class {\txt Length}. But if we were to define
a class {\txt Dimension} and define {\txt Length} to be an instance of
class {\txt Dimension}, then {\txt Length} would not be a class and so
{\txt 3 feet} could not be an instance of {\txt
Length}. Alternatively, we might define {\txt Length} to be a subclass
of {\txt Dimension}. But then we still cannot define {\txt 3 feet} to
be an instance of {\txt Length} because anything that is an instance
of {\txt Length} would also be an instance of class {\txt Dimension}
and {\txt 3 feet} is obviously not a dimension.

The fundamental limitation of conventional object-oriented languages
is that they do not allow a single concept to serve both as a class
and as an instance.  In \cite{FW96}, Welty and Ferrucci lay out a
comprehensive case that this limitation prevents conventional modeling
languages from capturing many important aspects of the world.  In
their primary example, they show the difficulties in modeling a simple
ontology including the notions {\txt Species} and {\txt Eagle}, while
also capturing the relationship between {\txt Eagle} and {\txt Harry},
a particular eagle.  One solution they consider and reject is to
separate the notion of ``eagle'' into two concepts: {\txt the Eagle}
and {\txt Eagle}, with the former an instance of {\txt Species} and
the latter a class whose instances are specific birds.  This approach
is inadequate because it fails to express a relationship between the
notions {\txt the Eagle} and {\txt Eagle}.
%The languages Ferrucci and Welty discuss cannot accommodate multiple
%universes of discourse, in which a single thing may have different meanings in different universes.

In order to overcome these limitations in expressiveness, several
object-oriented languages that are not statically typed, such as
Smalltalk, Python, and Self, allow for more flexible class
relationships \cite{Smalltalk80}, \cite{ArtMOP}, \cite{PythonMeta}. In
the case of Smalltalk and Python, classes are instances of
metaclasses. Static members of a class are modeled as ordinary members
of the corresponding metaclass.  But traditional metaclass systems
have placed important limitations on expressiveness.  For example, in
Smalltalk, the metaclass hierarchy is only two levels deep.  If we
want to represent the relationships ``{\txt A} is an instance of {\txt
B}", ``{\txt B} is an instance of {\txt C}", and ``{\txt C} is an
instance of {\txt D}", we can only do this in Smalltalk if {\txt D} is
the special Smalltalk class {\txt Metaclass}.  This does not allow the
multi-level hierarchies required to model the examples discussed above
\cite{Smalltalk80}.  Self, on the other hand, is a prototype-based
object-oriented language, where there are no classes at all; object
instantiation consists of cloning an existing instance. The members of
the clone can be modified and added to at will, making static checking
difficult.

An extremely expressive system of metaclasses, very similar to ours,
was presented by Cointe in \cite{objvlisp}.  In his paper, he makes a
number of persuasive arguments for the benefits of metaclasses.
In response to complaints that the Smalltalk80 metaclass system was
too complex, he suggested that a more powerful and flexible system, as
he and we present, has the potential to be much simpler.
He claims ``The ObjVlisp model's primary advantage is uniformity.''
We agree, and believe that the treatment of metaclasses presented here
unifies both disparate metaclass systems, and different ad-hoc
mechanisms describing the properties of classes as objects in
languages without metaclasses.

Cointe also quotes from \cite{cardelli:multiple}: ``With respect to
Simula, Smalltalk also abandons ... strong typing, allowing it to
... introduce the notion of meta-classes''.  In this paper, we show
that such a compromise is not necessary, and that static typing can
coexist with the flexibility of metaclasses.

We present and formalize a statically typed calculus with nominal
subtyping for metaclasses where every class can serve as both an
instance of a class and as a class of instances, and there is no
limitation on the levels of nesting of metaclasses. Our work is
motivated by \cite{DimUnits}, which presents a system for integrating
the static checking of dimensions of physical quantities with a
conventional object-oriented type system. In that system, metaclasses
are used extensively to model the type relationships of dimensions.
The notion of metaclass in that system is more general than that of
languages such as Smalltalk, where each class has a unique metaclass
that defines its class members. In this paper, we formalize the core
of that system related to metaclasses and prove it sound.  In the
process, we discover several interesting properties about the
structure that a nominal type system for metaclasses must take.

In the interests of economy and clarity of presentation, in the
following discussion we restrict ourselves to simple examples, and do
not delve in the the complexities of object-oriented analysis, or of
dimension types.  This is certainly not because we believe that
metaclasses are most useful for the development of toy object-oriented
programs.  On the contrary, metaclasses are likely to show their value
in large systems, both in terms of modeling and in the treatment of
design patterns in section \ref{designpatterns}.

The remainder of this paper is organized as follows.  In section
\ref{MCJ} we present a core calculus for metaclasses we dub MCJ.
We introduce the language and describe its key features, as
well as some of the important design choices.  The motivating examples
from the introduction are presented in MCJ in section \ref{examples},
demonstrating the benefits of metaclasses. We discuss in section
\ref{designpatterns} how the need for many object-oriented ``design
patterns'' is obviated through the use of metaclasses. We then
describe its formal properties in section \ref{formal}.  In section
\ref{formalsem} we provide a formal semantics for the language, and in
section \ref{proof} we prove a type soundness result.  In section
\ref{related} we consider related work and in section \ref{future} we
describe conclusions and future directions.

%Another, somewhat more flexible solution to this problem can be found in
%functional languages, where sets can be represented as predicates over
%objects, and sets of sets can be represented as predicates over
%predicates.  This hierarchy can obviously be extended indefinitely,
%but such programs would quickly become unwieldy.  Specialized language
%constructs would be needed for this approach to become a serious programming
%style, just as most object-oriented programming languages
%provide constructs to support that methodology.  Also in the
%functional world, higher-order functions represent a way of
%parameterizing at multiple levels, allowing some of the desired
%behavior described here to be implemented.  But few functional
%programs nest higher-orderness more than a few levels, since
%understanding the relationships  becomes more than one programmer
%could follow.  Instead, we need a way of structuring such
%relationships explicitly.

%Formal modeling of existing languages has become common, with numerous
%formal models for languages such as the Java (TM) Programming Language.  However, formal modeling
%can also serve to aid in the design process for a new language.  In
%this paper, we describe MCJ, a nominally-typed core calculus for metaclasses,
%which is intended to be the basis for a language design. While we have focused on a formal model as a language design guide, MCJ can also help understand existing languages with metaclasses.

\section{MCJ}
\label{MCJ}

\subsection{Overview}

\label{MCJintro}

MCJ is a generically-typed object-oriented calculus based on
Featherweight GJ that includes metaclasses. An MCJ program consists of
a sequence of class definitions followed by a trailing
expression. This trailing expression is evaluated in the context of
the class table induced by the class definitions.  Before discussing
the formal specification of MCJ, we first give a high-level overview
of the nature of MCJ class definitions.

Each class is an instance of a \emph{metaclass}.  A metaclass is
either a user-defined class or class {\txt Object}.  We say that if a
class {\txt C} is an instance of a class {\txt D}, then {\txt C} is an
\emph{instance class} of {\txt D}, and that {\txt D} is the
\emph{immediate containing class}, or \emph{kind}, of {\txt C}.  If
class {\txt E} is a superclass of {\txt D} then we also say that {\txt
C} is an instance class of {\txt E} and that {\txt E} is a containing
class of {\txt C}.

There are two key distinguishing features of MCJ. First, a class (or
an instantiation of a generic class) can be used as an expression.
Second, all classes have both a superclass and a kind. Instances of a
class inherit behavior from the superclass. The class itself, when
used in an expression context, inherits behavior from its kind.

A class definition consists of a header plus a collection of fields
and methods associated with the class.  The header names the class and
specifies the type parameters, as well as the superclass and the kind.
For example, the following header declares a class {\txt C} with
superclass {\txt Object} and kind {\txt D} and no type parameters:

\begin{verbatim}
class C kind D extends Object {...}
\end{verbatim}

Each class may have both class members and instance members. Class
members define the behavior of the class when used as an
instance. Instance members define the behavior of instances of a
class.  In the abstract syntax presented in this paper, class members
are distinguished from instance members by order, and with the keyword
{\txt class} \footnote{This syntax is not ambiguous since MCJ has no
inner classes.}.  Members of a class definition are laid out as
follows: first, class fields are defined, followed by class methods,
then instance fields and finally instance methods.  For example, the
following class definition includes a class method {\txt m} and an
instance method {\txt n}:

\begin{verbatim}
class C kind Object extends Object {
   class Object m() {...}
   Object n() {...}
}
\end{verbatim}

Class fields and methods bear a resemblance to static fields and
methods, but there are important differences. For a given class {\txt
C}, instance members of {\txt C}'s kind are inherited as class
members of {\txt C}. For example, a class method {\txt m} in {\txt C}
with the same name as an instance method of {\txt C}'s kind
overrides the definition of {\txt m} in the kind (and it must have the
same signature as the overridden method). Note that if an instance
class {\txt C} of type {\txt T} is assigned to a variable {\txt v} of
type {\txt T}, references to the instance members of {\txt v} refer to
the class members of {\txt C}.  As in \cite{DimUnits}, a class is
allowed to define both a static method and an instance method of the
same name. We prevent ambiguity in member references by requiring that
all references to class members in MCJ must explicitly denote the
receiver.

Superclasses behave as in \FGJ, providing implementation inheritance
of instance behavior and subtyping. A class, when used as an
expression, is an instance of its kind.  If a method is invoked on a
class {\txt C}, first the class methods of {\txt C} are examined.  If
the method is found there, it is invoked. Otherwise, the
\emph{instance} methods of the kind {\txt D} of {\txt C} are searched,
and if it is found there, the method is invoked.  Otherwise, the
superclass hierarchy of {\txt D} is examined as it is for an ordinary
instance of {\txt D}.

For example, consider the following class definitions:

\begin{verbatim}
class A kind Object extends Object {...}
class B kind Object extends A {...}
class C kind B extends Object {...}
\end{verbatim}

To resolve the method call $\invoke \C {\mof {}}$ first the class
methods of $\C$ are examined, and if a match is found, it is invoked.
Then the \emph{instance} methods of $\B$ and then $\A$ are examined.

If the method call were instead $\invoke \bb {\mof {}}$, where {\txt
b} denotes an instance of {\txt B}, then the instance methods of first
$\B$ then $\A$ would be examined.

%For example, suppose we have the following class definitions:

%\begin{verbatim}
%class E kind Object extends Object {...}
%class D kind Object extends E {...}
%class C kind D extends Object {...}
%\end{verbatim}

%Then, for the expression:

%\begin{verbatim}
%C.m()
%\end{verbatim}

%Class {\txt C} would first be examined to see if it contained a class method named {\txt m}. If so, this method would be invoked. Otherwise, class {\txt D} would be checked to see if it contained an instance method named {\txt m}. If not, class {\txt E} would then be checked for an instance method named {\txt m}.

The generic type system of MCJ is similar to the generic type system
in \FGJ. However, as in \cite{NextGen, DimUnits}, we allow for type
variables to occur in type-dependent contexts such as casts,
preventing the use of type erasure as an implementation technique. In
addition, the receiver of a reference to a class member may be a type
variable. Unlike the system presented in \cite{MixGen}, MCJ does not
support first-class genericity because a naked type variable must not
appear in the {\txt extends} clause or {\txt kind} clause of a class
definition and {\txt new} expressions on naked type variables are not
supported. Also, polymorphic methods (which are orthogonal to the
features we explore) are not supported.

Because classes can be used as expressions, we need a bound on the
behavior of classes when used in expression contexts.  Therefore, we
place two bounds on every type variable {\txt T} in the header of the
class definition. The first bound on {\txt T} is a bound on the kind
of an instantiation of {\txt T}; the second bound is a bound on the
superclass of {\txt T}. For example, the following class header
declares a class {\txt C} with one type parameter {\txt T} that must
be instantiated with a type of kind {\txt D} that is a subtype of
{\txt E}:

\begin{verbatim}
class C<T extends (D,E)>
   kind Object
   extends Object {...}
\end{verbatim}

\subsection{Key Design Points}

\paragraph{Private Constructors}
\label{privconst}

One benefit of a metaclass system is that constructors no longer need
to play such a central role in the language. Class instances exist at
the beginning of a program execution; they need not be
constructed. Non-class instances are naturally constructed with class
(factory) methods \cite{GOF}; in this way a class instance can be
passed as a type parameter and factory methods can be called on the
class instance to construct new instances without stipulating the
class of the constructed instance. This flexibility is similar to that
provided with first-class genericity \cite{NextGen, MixGen}, but
obviates the need for {\txt with} clauses because the bound on the
kind of a type variable stipulates what factory methods can be called
on it.  A class can define class methods with arbitrary signatures
that return new instances of the class, but each such method must
ultimately includes a {\txt new} expression, which can occur only
within the scope of the class whose instance it returns. A {\txt new}
expression looks like a call to method named {\txt new}, that takes a
single argument for each type parameter and each instance field of the
syntactically enclosing class. The result of a {\txt new} expression
is a new instance of the enclosing class whose type parameters are
instantiated with, and whose fields are initialized with, the given
arguments. In our formal semantics, {\txt new} expressions are
annotated with the name of the enclosing class. These annotations need
not be added by a programmer; they could be added easily by a
straightforward syntactic preprocessing over the program, since they
can be determined merely by the lexical scopes of the {\txt new}
expressions.

All uses of constructors in \FGJ\ \cite{FJ} are macro-expressible
\cite{FellExpress} in MCJ, that is, they can be expressed via local
transformations.  Of course, not all \FGJ\ programs can be expressed
in MCJ, because of the differences in the type system. Namely, MCJ
does not include polymorphic methods.

Metaclasses allow other kinds of flexibility in object creation, as
discussed in section \ref{designpatterns}.

\paragraph{Fields and Initialization}

To ensure type safety, we must have an initial value for every field,
or prevent fields from being used before they are initialized.  For
instance fields of objects created with a constructor, this is achieved
by requiring that the constructor have an argument for every instance
field.  Combined with the solution for flexibility in object creation
outlined above, this allows us the simplicity in semantics of FGJ, and
the flexibility of more general constructors.

However, classes which are themselves instances of other classes have
fields as well, and these fields must also be initialized before they
are used.  These fields include those instance fields of the kind
which become class fields of the new class.  Our solution is to
require that \emph{all class fields, including those obtained by
inheritance, must be redeclared} and provided with values.

% In a {\txt new} expression, all instance fields must be initialized at
% creation time. This requirement eliminates the problem of
% uninitialized instance fields. However, when a class is the kind of
% another class, the instance fields of the kind become class fields of
% the instance class.  Since the instance class can then be used as a
% value with no further initialization, these variables must be
% initialized somewhere. Our solution is to require that every class
% field inherited from a container class be included in a new class
% field definition and that every class field in a class definition
% include an initializing expression.

\paragraph{typeOf}

Given that class references can be used as expressions, it is natural
to ask: what is the type of a class reference?  In the world of the
$\lambda$-calculus, such types are known as kinds.  In MCJ, however,
the kind of a class does not capture all of the properties of the
class as a value.  For example, a class may add new class fields or
class methods which are not present in the kind.  Therefore, each
class freely generates a new type, which is the type of the class
considered as a value.  We represent this type with the {\txt typeOf}
operator, which is produces a type that is not also a class.  Since
this type is not a class, it cannot be the superclass or kind of
another class, and it cannot serve as the instantiation of a generic
type parameter.

These new types can create complex relationships between classes and types.  For
example, the following class headers:

\begin{verbatim}
class A() kind Object extends Object {...}
class B() kind A extends Object {...}
\end{verbatim}

induce the following type relationships:
\begin{displaymath}
\begin{array}{c}
\typ \B {\typeof \B} \\
\subtype {\typeof \B} \A \\
\B~ {\txt instanceof}\ \A
\end{array}
\end{displaymath}

The name {\txt typeOf} has been used in other contexts to refer to a
runtime operation that determines the dynamic type of a object.
However, despite the name similarity, these functions bear no
relationship to the operator described here.  In MCJ, an application
of {\txt typeOf} is a static type reference.

\paragraph{Non-transitivity}

The standard subclassing relationship, like subtyping, is transitive.
That is, if $\B$ is subclass of $\C$ and $\C$ is a subclass of $\D$,
then $\B$ is a subclass of $\D$.  However, this relationship does not
hold for {\txt instanceof} relationships.  Therefore, if if $\B$ is an
instance of $\C$ and $\C$ is an instance of $\D$, then no judgment
about the relationship of $\B$ to $\D$ can be inferred.

\paragraph{Cyclic class hierarchies}
One property of metaclasses that we have not yet discussed is that
cycles can occur among instance class relationships. The simplest such
cycle is:

\begin{displaymath}
\typeFront {\typeArg \C{}} {\typeArg \C {}} \Object \braces{}
\end{displaymath}

Although this pathology appears to be dangerously close to Russell's
paradox, it does not lead to inconsistencies.  Class \C\ is an
instance of itself. It can be used in any context where one of its
instances can be used.  \C\ can be thought of as a self-replicating
value. This class hierarchy causes no problems for method lookup
because method resolution never proceeds through more than a single
containing class to an instance class. Therefore, we see no reason to
disallow it.

\subsection{Examples, Reprised}

\label{examples}

Having seen the outlines of MCJ, we return to the motivational
examples discussed earlier.  First, the relationships between
dimensions is easy to capture. \footnote{Of course, dimensions have
many subtleties, not captured here.  All we present is the essence of
the type relationships.} We define

\begin{verbatim}
class Dimension kind Object extends Object { ... }
class Length kind Dimension extends Object { ... }
class Meter kind Length extends Object { ... }
\end{verbatim}

Here {\txt Meter} is a singleton class (there is only one Meter, in
the Platonic sense).  We have easily expressed the desired type
relationships, and can statically check program invariants that rely
on dimensional relationships.

The example from \cite{FW96} is also easily expressed:

\begin{verbatim}
class Species kind Object extends Object { ... }
class Eagle kind Species extends Object {
      class Eagle make(String name) { ... }
}
Eagle harry = Eagle.make("Harry")
\end{verbatim}

Here we simply create an object to represent our concrete instance of
an Eagle.  The naive implementation of the above code in a
conventional OO language would look something like this:

\begin{verbatim}
class Species extends Object { ... }
class TheEagle extends Species { ... }
class Eagle extends Object { ... }
Eagle harry = new Eagle("Harry")
\end{verbatim}

There are at least two substantial problems with this code.  First,
{\txt TheEagle} and {\txt Eagle} have no relationship to each other in
the type system.  One is a singleton class, representing a particular
species, and one is a name for a set of objects, being the members of
that species.  The fundamental nature of species has a two-level
containment relationship, which the type system is failing to express.
This means that our type checker cannot determine what we want, and
cannot help us avoid mistakes.

Another problem, which is in some ways just a symptom of the first
problem, but which bears special attention, is the use of generic
types.  Imagine the following method, where T is a type parameter in
scope:\footnote{Generic methods would be straightforward to add to
MCJ, but we excluded them to simplify the presentation of the type
system, and since they do not add complexity to the proofs.}

\begin{verbatim}
foo(T x) { ... }
\end{verbatim}

If we want to perform the call {\txt foo(harry)}, then the bound on
{\txt T} must be either {\txt Object}, in which case {\txt foo} must
have no knowledge of the object, or Eagle, in which case the function
cannot handle multiple {\txt Species}.  The ``solution'' in
conventional languages is to use a bound of {\txt Object} and insert a
downcast, which fails at runtime if the wrong argument is passed.  In
MCJ {\txt T} can be bound by kind {\txt Species}, and instantiated
with {\txt Eagle}, allowing the original call to be type-checked.
This gives the programmer both safety and expressiveness.

\section{Design Patterns as Language Features}
\label{designpatterns}

Design Patterns, as exemplified in \cite{GOF}, have had a very
substantial impact on the world of object-oriented programming and
beyond.  Design patterns allow programmers to increase the flexibility
and abstraction of their software designs.  However, few design
patterns have been integrated into programming languages.

One of the advantages of our metaclass framework is that it allows
clean expression of several patterns in the language, rather than
adding additional abstraction techniques on top of the language.  For
example, a number of the classic design patterns deal with object
creation, as discussed above in section \ref{privconst}.  It is clear
that the concept of a Factory method is easily (and necessarily)
expressed in MCJ.  However, a number of other patterns also find easy
expression.  The Abstract Factory pattern is a simple application of
inheritance in the metaclass hierarchy.  The Prototype pattern is, as
mentioned in the Design Patterns book, supported in languages with
metaclasses.  Finally, there is no need for the Singleton pattern,
since the class itself can serve as a singleton.  While this is
possible in the Java Programming Language and in \Cpp, it is
undesirable since the static behavior that would need to be used
cannot inherit from another class.  In fact, the inflexibility of
class or static operations is one of the rationales cited for the
Singleton pattern.

Seeing that a number of patterns can be expressed quite simply in MCJ,
the question arises: is the simple expression of these patterns a
benefit?  We argue that it is.  The existence of design patterns is a
sign of shortcomings in language design.  Fundamentally, a design
pattern is an abstraction that cannot be expressed in the language.
For example, the Singleton and Visitor patterns are abstractions that
have obvious invariants, but they cannot be expressed directly and
they require complex cooperation from many parts of the system.

Sophisticated macro systems, such as those found in Common Lisp
\cite{steele-90} or Scheme \cite{SyntaxCase} allow for expression of
many forms of abstraction, as do extremely flexible object systems
such as CLOS. However, these solutions add significantly to what a
programmer must understand in order to use the language productively.

An alternative solution explored in this paper is to provide additional
flexibility in the language, which does not allow for arbitrary
additional abstraction, but instead makes the abstractions commonly
needed easy to express.  One does not need the full power of macros to
encode the Factory pattern.  In MCJ, we have presented a language
where several kinds of patterns are easy to express, without adding an
additional language on top of the original.

\section{Formal Specification of MCJ}

\label{formal}

Having outlined the motivation for metaclasses, and the basics of
their use, we now turn to a formal exposition of the syntax and
semantics of MCJ, followed by an outline of the proof of soundness for
the calculus.

\subsection{Syntax}
\label{Syntax}

\figSyntax

The syntax of MCJ is given in Figure \ref{fig:syntax}. When describing
the formal semantics of MCJ, we use the following metavariables:

\metavars

As in \FJ\ and many other works, $\ox$\ stands for a possibly-empty
sequence of \x.  $\SforX$ denotes a substitution of the $\oS$ for the
$\oX$, which can be applied to either an expression or a type, and
which can substitute either type or expression variables.

A number of symbols are used to abbreviate keywords: $\superSym$
stands for {\txt extends} and $\kindSym$ stands for {\txt kind}. Also
$@$ represents concatenation of sequences of syntactic constructs.
Finally, $CT(\T)$ is a lookup in the class table for the definition of
$\T$.

A number of restrictions on MCJ programs are implicit in the formal
rules.  First, we assume that all sequences of methods and fields are
free of duplicates.  Second, there is an implicit well-formedness
constraint on programs that no class be a superclass of itself, either
directly or indirectly (however, as discussed below, cycles in the
kind hierarchy are allowed).  Third, we assume that {\txt this} is
never used as the name of a variable, method or field.  We also take
\Object\ to be a distinguished member of the hierarchy, with no
specific definition, which does not have a superclass or a kind.

For example, the simplest MCJ class is:

\vspace{.1cm}
$\typeFront {\typeArg \C {}}  {\Object} \Object \braces{}$
\vspace{.1cm}

This is a class named $\C$, with no type arguments, with kind
$\Object$ and superclass $\Object$, which has no fields or methods
whatsoever.  In examples that follow, we omit empty {\txt <>}.  Note
that both the kind and the superclass are $\Object$.  MCJ does not
have a distinguished class $\Class$.  Interestingly, we determined
that a class $\Class$ would need no consideration from the type
system.  Since both $\Class$ and $\Object$ would sit atop the
hierarchy with no contents, we determined there was no need to include
both of them.  In a more substantial language, where $\Object$ might
have methods or fields, there might be a use for a distinguished
$\Class$ class.  However, it would not need any special treatment by
the type system.

A more complicated MCJ class is:\footnote{Here we use {\txt extends}
for $\superSym$ and {\txt kind} for $\kindSym$.}

\begin{verbatim}
class Pair<A extends (Object, Object),
           B extends (Object, Object)>
   kind Object extends Object
{
 class Pair<A,B> make(A a, B b){ new(A,B) (a,b) } ;
 A fst;
 B snd;
 Pair<B,B> setfst(B b){
     return new <B,B> (b, this.snd);
 };
}
\end{verbatim}

This class specifies a polymorphic pair data structure.  It also
demonstrates a number of important MCJ features.  First, we see two
different kinds of methods.  The {\txt make} method creates a new
instance of {\txt Pair}.  This method contains a {\txt new}
expression, which initializes the instance fields of the class
positionally, so that the first argument to {\txt make} becomes {\txt
 fst}, and the second becomes {\txt snd}.  Also, note that we must
mention the type parameters of the newly created instance.

Second, we have an instance method, {\txt setfst}, which creates a new
pair with the old second element, and a new first element which has
the same type as the second element.

Finally, adding the following to the above definition of {\txt Pair}
gives a full MCJ program:

\begin{verbatim}
class O kind Object extends Object
{
 class O make(){return new () ()};
}

Pair<O,O>.make(O.make(), O.make()).fst
\end{verbatim}

This program, which creates two instances of {\txt O}, then creates
a {\txt Pair} to hold them both, and finally selects one of the two to
become the value of the program, demonstrates three of the expression
forms in MCJ.  {\txt Pair<O,O>} is a type, which, by itself, can
be a value.  {\txt O.make()} is a method invocation (here with a
class, not an instance, as the receiver).  The entire expression is a
field access, of the {\txt fst} instance field.

Other kinds of expressions are seen in the body of the {\txt Pair}
class.  {\txt new<A,B> (a,b)} is a new expression, which creates an
instance of the enclosing class (here {\txt Pair}).  In the body of
{\txt make} is a reference to the variable {\txt b}, which has the
obvious value.  The final directly expressible expression is the cast,
which works as follows: {\txt (Object)O.make()} is an expression of
type \Object that evaluates to an instance of {\txt O}.

With an understanding of expressions, we can examine the rest of the
{\txt Pair} class.  There are two methods: a class method ({\txt
make}) and an instance method ({\txt setfst}).  There are also two
fields, both of which are instance fields and initialized the {\txt
new} expressions. In MCJ, a {\txt new} expression is different from
those in \FGJ; each {\txt new} expression evaluates to an instance of
the syntactically enclosing class.\footnote{In the grammar of figure
\ref{fig:syntax}, {\txt new} expressions are annotated with this
enclosing class.  In the examples, this redundant information is
elided.}  Had there been additional fields in the superclass of {\txt
Pair}, it would have been necessary to initialize them in the {\txt
new} expression as well. {\txt new} expressions are explained in
detail in section \ref{privconst}.

%\footnote{There is a strong
%resemblance between the {\txt new} expressions here and the
%constructors of \FJ.  However, in MCJ they are a part of the
%language, whereas in \FJ\ they are a restriction on a more general form.}

\subsection{Semantics}
\label{formalsem}

There are two forms that a value can take in MCJ: that of an instance
class, and that of a conventional (non-class) value. If we were to
distinguish these two forms of value, we would significantly increase
the number of rules necessary to describe our semantics because every
rule referring to a value would have to be written twice. To avoid
this complexity, we introduce a special {\em mapping construct}
(similar to a record value) to denote the results of
computations.\footnote{This simplification was suggested by Jan-Willem
Maessen in response to an earlier draft of the MCJ semantics.}  A
mapping takes field names to expressions, and is annotated with a
ground type. Mappings from a sequence of fields $\of$ to a sequence of
expressions $\oe$ with type $\GG$ are written $\mapping \of \oe \GG$.
A mapping denoting an instance class {\txt C} consists of a sequence
of class fields mapped to a sequence of expressions and is annotated
with the type {\txt typeOf(C)}. Note that the right hand sides of such
maps need not always be values, and thus computation can take place
inside of a mappings.  Mappings are not available to the programmer,
and thus can only be created by operation of the reduction rules.
Therefore, unlike \FJ, our reductions do not operate entirely in the
expressible syntax of the language.  We use the metavariable $\Phi$ to
range over mappings.  When we need to refer to the type annotation of
a mapping explicitly, the metavariable is written $\Phi_{\GG}$.

The
semantics are given in figures \ref{fig:subtype}, \ref{fig:exprty}
,\ref{fig:wf}, and \ref{fig:comp} with auxiliary functions given in
figure \ref{fig:aux}.

\subsubsection{Typing}
Rules governing the typing of MCJ programs are given in figures
\ref{fig:subtype}, \ref{fig:exprty} and \ref{fig:wf}.  The
$bound_{\Delta}$ function is defined as follows:
\begin{displaymath}
\begin{array}{c}
\boundVar \\
\boundTypeof \\
\boundClass
\end{array}
\end{displaymath}

This function maps type variables and {\txt typeOf} applied to type
variables to their bounds, and leaves others unchanged.

The metavariables $\Delta$ and $\Gamma$ range over bounds
environments, written $\oX \superSym \oS$ and type environments,
written $\typ \ox \oT$ respectively.  A bounds environment contains
two bounds for each type variable, so that if $\Delta = \X \subtypeSym
(\A,\B)$, then $\Delta(\X) = \A$ and $\Delta({\typeof \X}) = \B$.

The notation $\A \subtypeSym \B$ means $\A$ is a subtype of $\B$.
Subtyping judgments are made in the context of a bounds environment,
which relates a type to the declared bound of that type from the class
header.

Type judgments are of the form $\dgprove {\typ \e \T}$, which states
that in bounds environment $\Delta$ and type environment $\Gamma$,
expression $\e$ has type $\T$.  Type judgments are not transitive in
the way that {\txt instanceof} relationships are with respect to
subtyping: if $\dgprove {\typ \e \T}$, and $\dproves {\subtype \T
\SSS}$, it is not necessarily the case that $\dgprove {\typ \e \SSS}$.
This is important, since our proofs depend on having unique
derivations of a given typing judgment. We represent empty
environments with $\emptyset$ and we abbreviate judgments of the form
$\eeproves {\typ \e \T}$ and $\eproves {\subtype \T \SSS}$ as ${\typ
\e \T}$ and ${\subtype \T \SSS}$ respectively.

We now discuss several non-obvious aspects of our type system resulting from the need to statically check uses of metaclasses.

\paragraph{Casts and Mappings}
We follow \cite{MixGen} in typing all casts as statically correct, so as to
avoid the complications of ``stupid casts''.  Mappings, which are not
expressible in the source language, are typed merely by a
well-formedness constraint.

\subsubsection{Well-Formedness}

There are three kinds of well-formedness constraints on MCJ
programs.  First, type well-formedness, written ``$\dproves \T \ok$'',
states that type $\T$ refers to a defined class, and that if it is a type
application, the arguments satisfy the bounds.

More important are the class and method well-formedness constraints,
which together determine if a class table is well-formed, and thus
part of a legal MCJ program.  Method well-formedness is a
judgment of the form ``$\M \okin\ \GG$'', where $\M$ is a method.  We make
use of the latter portion ($\GG$) of this judgment as a second argument,
allowing us to use this rule to check both instance and class methods.
Method well-formedness consists mostly in fitting the body to the
return type, and checking for invalid overrides.

Class well-formedness involves all of the above checks.
All methods must be well-formed, and all class fields must
have correct initialization expressions.  Further, class fields
must contain the instance fields of the kind.  Finally, all
{\txt new} expressions must have the correct class name annotation.

\figExprty

\figWF

\subsubsection{Evaluation}

The evaluation rules for MCJ are given in figure \ref{fig:comp}.
The evaluation relation, written $\trans \e \ep$, states that $\e$
transitions to $\ep$ in one step.  The reflexive transitive closure of
this relation is written $\e \transarrow \ep$.

%\paragraph{Method Dispatch}
%
%Users of most object-oriented languages are familiar with the concept
%of dynamic dispatch, which resolves method access to the run-time type
%of the receiver, and looks up the class hierarchy for the closest
%matching method definition.

%The notion of classes as values has several repercussions on the semantics of dynamic dispatch.
%First, a receiver can be a type variable, meaning that the actual instantiation type of the receiver
%must be determined at runtime.
%Second, when a
%class in a receiver context does not have a matching method definition, the
%system must search upward, not to the superclass, but to the instance
%methods of the kind.  This is because although the receiver
%is a class, it is also an instance of the containing class.

\paragraph{Bad Casts}
There is one way that evaluation in MCJ can get stuck: the bad cast.
This problem occurs for all languages that allow static downcasts.  A
expression $\e$ is a bad cast iff $\e = \cast \SSS {\Phi_{\T}}$ where
\SSS\ is not a supertype of \T.

The evaluation relation given here is non-deterministic. For example,
no order is prescribed for evaluating the arguments to a {\txt new}
expression or method call, or for evaluating the initializers for
class fields.  There are also a number of places where either
congruence or reduction rules can be applied.  Therefore, in the
presence of non-termination or bad casts, the results may differ
depending on evaluation order.  However, confluence can be regained
simply by requiring that $\rulename{C-Mapping}$ is applied whenever
possible.  This restriction, combined with the requirement in the
premise of $\rulename{R-New}$ that the arguments be values, ensures
that every program with an error or non-termination will either cause
an error, or fail to terminate.

\figComp

\subsection{Type Soundness}
\label{proof}

With the above definitions, we are now able to turn to a proof of type
soundness. Given the simplicity we are able to achieve in the
definitions, the proof is not significantly more complicated than the
soundness proof given for Featherweight GJ\ \cite{FJ}.  However, there
are a number of significant lemmas, of which we list the most
important:

\fieldsPreservedLemma

\fieldsPreservedProof

\methodTypingLemma

\methodTypingProof

\substitutionPreservesTyping
%\substitutionPreservesTypingProof
\myProof With the above two lemmas, this one follows by
straightforward structural induction over the derivation of $\dgprove
{\typ \e \T}$.

\vspace{.3cm}

Given the above lemmas, the following subject reduction proof is a
simple structural induction with a case analysis on the typing rule
used to derive $\eept \e \SSS$.

\preservation

\preservationProof

The proof of progress presents no additional complications, and
requires only one new lemma.

\fieldsAgreeLemma

\progress

\progressProof

From these, we can conclude the desired type safety result.

\soundness

\soundnessProofShort

\section{Related Work}
\label{related}

A number of object-oriented languages have included some form of
metaclass system. Most notable among these is Smalltalk
\cite{Smalltalk80}, but others include Common Lisp with CLOS
\cite{ArtMOP}.

All of these systems share a common architecture of the metaclass system
in which each class has its own freely generated metaclass, defined by
the class methods and fields of the class.  In contrast, MCJ
provides a hierarchy for structuring metaclass relationships, which
provides significantly more modeling and abstraction flexibility.

The metaclass system present in Python \cite{PythonMeta} is quite
similar to that provided here, with inheritance of instance methods as
class methods from arbitrary other classes as metaclasses.  However,
Python is dynamically-typed, and many of the uses to which Python
metaclasses are put are not possible in a statically typed language.
The work on static type systems for Python has not included
metaclasses \cite{Starkiller}.

Several type systems have also been proposed for languages with
metaclasses, including the Strongtalk language \cite{Strongtalk},
which turns Smalltalk into a structurally-typed language with
extensive static checking.  However, because the Smalltalk metaclass
system is so different from the one in MCJ, many of the interesting
aspects of the type system do not carry over.  Furthermore, the
Strongtalk papers do not provide a formal semantics and analysis of
the system.  A formal analysis of inheritance in Smalltalk is provided
in \cite{cook:thesis} but this again does not consider the hierarchy
of metaclasses presented here.

Graver and Johnson \cite{graver:90} present another type system for
Smalltalk, with a formalism and sketch of a safety proof.  Again, the
metaclass type system is substantially different, reflecting the
underlying Smalltalk system.  Additionally, the paper is concerned
primarily with optimization as opposed to static checking.

Cointe \cite{objvlisp} presents a model of metaclasses extremely
similar to that presented here.  In it, he provides several
overlapping motivations to our own.  One is to regularize the
metaclass system of Smalltalk, and another is to enable additional
programming flexibility.  To this we add modeling freedom, and a
relation to static methods in more recent OO languages.  Cointe's
work, however, does not provide a formal model, so it is difficult to
determine the exact relationship between the systems.  Additionally,
his work is in a untyped setting (Lisp and Smalltalk) and thus the
safety theorems proved here are not possible.

When viewed as ``instance generators'', our metaclasses are similar to
prototypes in untyped languages such as Self.  Prototypes generate new
instances, which themselves can generate new instances.  Our language
is more restrictive than prototype-based languages in the sense that
all metaclasses and instance classes must be declared statically
(i.e., by writing down class definitions).  But our language is more
expressive in the sense that we include classes and subclassing
relationships.  Also, unlike typical prototype-based languages, our
language is statically typed.

Formalized calculi for object-oriented languages are abundant in the
programming languages community today, including \FJ\ \cite{FJ}, upon
which MCJ is based.  However, none of them have considered static
methods, which are the closest analogue of the metaclass functionality
in MCJ.

Finally, the motivation for this work comes from the language MetaGen,
introduced in \cite{DimUnits}.  MetaGen can be seen as an extension of
MCJ, which provides numerous other advanced type features.  Here we
restrict ourselves to metaclasses and analyze the properties of the
system formally.

\section{Conclusions and Future Work}
\label{future}
With MCJ, we have devised a core calculus for metaclasses that is more
flexible than that available in more traditional metaclass systems
such as Smalltalk and that allows clean expression of many common
design patterns.  In doing so, we have demonstrated that metaclasses
can be added to a nominally-typed, statically-checked language without
either significant complication of the semantics or difficulty in the
proof of soundness.

In addition to this contribution, we have elucidated several other
important points about the integration of metaclasses into an
object-oriented system.  The {\txt typeOf} type operator is to our
knowledge unique, and plays a key role in the soundness of the
system.  The discovery that the {\txt Class} class played no special
role, and that thus the class and metaclass hierarchies could both be
rooted at {\txt Object} is also novel.  Finally, we have shown how an
expressive framework of metaclasses has positive effects in other
areas of the language, such as mechanisms for object construction.

A natural extension of the work in this paper is to expand MCJ to
include more of the features presented in \cite{DimUnits}, so as to
allow inclusion of the system for checking dimensions of physical
quantities. Such an extension would allow for a proof of ``dimensional
soundness'' in the resulting system.  Further, any realistic system
will have imperative features, and while such features do not seem
likely to interact badly with metaclasses, no sound system is built on
such assumptions.

Another interesting extension would be to expand our calculus to
include either multiple inheritance (as does Python) or some
alternative such as mixins or traits, as there may be interesting
interactions between metaclasses and these features that have yet to
be discovered.

\section*{Acknowledgments}

We would like to thank Jan-Willem Maessen for his extremely helpful
feedback on the formal rules presented in this paper.  We also thank
David Chase for not believing in us unless we were right, Victor
Luchangco for many valuable discussions, and Guy Steele for his many
helpful comments.

\bibliographystyle{plain}

\bibliography{paper}

\end{document}

### 2004-09-29 metaclasses/semantics-proofs.tex
initial import

\newcommand{\metavars}
{
\begin{itemize}
 \item Expressions or mappings: \e, \dd, \rr
 \item Field names: \f, \g
 \item Variables: \x
 \item Method names: \m
 \item Values: \vv
 \item Method declarations: \M, \N
 \item Type names: \C, \D
 \item Types: \I, \J, \K, \T, \U, \V, \W
 \item Non-{\txt typeOf} Types: \OO, \PP, \Q, \R, \SSS
 \item Types that are either {\txt Object} or a type application: \A, \B
 \item Ground types (contain no type variables): \GG
 \item Type variables: \X, \Y, \Z
 \item Mappings (field name $\mapsto$ value): $\Phi$
\end{itemize}
}

\newcommand{\unproved}
{
\myProof
Unproved.
}

\newcommand{\preservation}{
\begin{theorem}[Subject Reduction]
If \eeproves{\typ{\e}{\SSS}} and \trans{\e}{\ep}
then \eeproves{\typ{\ep}{\T}} where {\proves \emptyset {\subtype{\T}{\SSS}}}.
\end{theorem}
}
\newcommand{\progress}
{
\begin{theorem}[Progress]
If \eeproves{\typ{\e}{\SSS}} then one of the following holds:
\begin{itemize}
\item $\e = \mapping{\of}{\ov \vv}{\GG}$
\item \trans \e \ep
\item $\e = \cast{\SSS}{\ep}$ and $\eeproves{\typ{\ep}{\T}}$ and $\eproves{\notsubtype \T \SSS}$.
\end{itemize}
\end{theorem}
}
\newcommand{\soundness}
{
\begin{theorem}[Soundness]
If \typ{\e}{\SSS} then either
\begin{itemize}
\item $\e \transarrow  \mapping{\of}{\ov \vv}{\GG}$
\item $\e \transarrow \ep$ where $\ep$ is an invalid cast
\item $\e$ reduces infinitely.
\end{itemize}
\end{theorem}
}

\newcommand{\preservationProof}
{
\myProof
We prove this by structural induction on the derivation of \trans{\e}{\ep}.

\begin{description}

\proofcase{\rulename {R-Object}}

Immediate.

\proofcase{\rulename {R-New}}

Immediate from the premises of \rulename{T-New} and \rulename{R-New}.

\proofcase{\rulename {R-Class}}

Immediate from the premises of \rulename{WF-ClassDef} and \rulename{R-Class}.

\proofcase{\rulename {R-Cast}}

By \rulename{T-Cast}, $\e = {\cast{\T}{\Phi_{\GG}}}$ must have type
$\T$.  By \rulename{T-Mapping}, $\ep = {\Phi_{\GG}}$ must have type
$\GG$.  By hypothesis of the reduction rule, $\eproves {\subtype \GG
\T}$.

\proofcase{\rulename{R-Field}}

We know that $\e = \invoke {\mapping \of \od \GG} \f_i$ and that $\ep
= \dd_i$.  Since $\e$ must have been typed by \rulename{T-Field}, we
know that ${\fields \GG} = \oT\ \of$ and $\eeproves {\typ \e \T_i}$.
Further, since ${\mapping \of \od \GG}$ must have been typed by
\rulename{T-Mapping}, we know that ${\eeproves {\typ {\dd_i} \SSS}}$
where $\eproves {\subtype \SSS \T}$.

\proofcase{\rulename {R-Invk}}

We know that $\e = \invoke {\Phi_{\GG}} {\mof \dd}$ and that $\ep =
{\substituteTwoArg \ox \od \this {\Phi_{\GG}}} \e_0$. Further, $\e$
was typed by \rulename{T-Invk} to have type $\U$ where $\mtype \m \GG
= \arrType \oT \U$ and $\eeproves {\typ \od \oTp}$ and $\eproves {\subtype \oTp
\oT}$.  As a premise of \rulename{R-Invk}, we know that $\mbody \m \GG
= (\ox, \e_0')$.

By the lemma $mtype$ and $mbody$ agree, $\proves
{\emptyset;{\typ{\ox}{\oT}},{\typ{\this}{\GG}}} {\typ{\e_0'}{\Up}}$
where $\eproves {\subtype \Up \U}$.  Then by the lemma Substitution
Preserves Typing, $\eeproves {\typ {\ep = {\substituteTwoArg \ox \od
\this {\Phi_{\GG}}} \e_0} \Upp}$ where $\eproves {\subtype \Upp \Up}$.

\proofcase{\rulename {C-Cast}}

Trivial, since $\eproves{\typ{\cast{\T}{\e}}{\T}}$ for any $\e$.

\proofcase{\rulename {C-Map}}

Immediate from the induction hypothesis and the transitivity of subtyping.

\proofcase{\rulename {C-New}}

Immediate from the induction hypothesis and the transitivity of subtyping.

\proofcase{\rulename {C-Arg}}

Immediate from the induction hypothesis and the transitivity of subtyping.

\proofcase{\rulename {C-Rcvr}}

We know that ${\e = {\invoke {\e_0} {\mof \dd}}}$ and $\ep = \invoke {\e_0'}
{\mof \dd}$. Further, $\e$ must have been typed by \rulename{T-Invk},
which means that $\eept {\e_0} \W$ for some ground type $\W$, and that
$\eept \od \oV$ and $\esub \oV \oU$, and also $\mtype \m \W = \arrType
\oU \T$.  By the induction hypothesis, $\eept {\e_0'} \Wp$ where $\esub
\Wp \W$.  Therefore, by lemma Subtyping Preserves Method Typing,
$\mtype \m \Wp = \arrType \oU \T$ and thus $\eept {\invoke {\e_0'}
 {\mof \dd}} \T$ by \rulename{T-Invk}.

\proofcase{\rulename {C-Field}}

If $\trans {\invoke \e \f_i} {\invoke \ep \f_i}$, then $\trans \e
\ep$.  Further, ${\invoke \e \f_i}$ must have been typed by
\rulename{T-Field} to have type $\T_i$.  Therefore, by the induction
hypothesis, $\eept \e \SSS$ and $\eept \ep \Sp$ where $\esub \SSS
\Sp$.  Then, by the lemma Fields are Preserved by Subtypes, $\fields
\Sp = \fields {\SSS}\ @\ \oF$ for some $\oF$, and by
\rulename{T-Field}, $\eept {\invoke \ep \f_i} \T_i$.

\end{description}
}

\newcommand{\progressProof}
{
\myProof  By induction over the derivation of $\eept \e \SSS$.

\begin{description}

\proofcase{\rulename{T-Var}} This is a contradiction, since $\e$ is ground.

\proofcase{\rulename{T-Class}} In this case $\e = \CSp$ where $\CSp$ is ground.  Then by the lemma  Agreement of {\it field-vals} and {\it fields}, $\typeparams{\CSp} = \oX \implSym \parens {\oI, \oJ}$ for some     Further, $\fieldVals{\typeof \CSp}  = \Tfe$ for some $\oT$, $\of$, and $\oe$.  Therefore, \rulename{R-Class} applies and $\trans \SSS {\mapping \of {\SpforX \CSp} {\typeof \CSp}}$.

\proofcase{\rulename{T-Mapping}} Either $\e$ is already a value, or $\e = \mapping \of \oe \GG$ where not all of the $\oe$ are values.  Then by the induction hypothesis, there is some $i$ such that either $\trans {\e_i} {\ep_i}$, in which case \rulename{C-Mapping} applies, or $\e_i$ contains a bad cast, and the case is complete.

\proofcase{\rulename{T-Cast}}  Here there are three cases:
\begin{itemize}
\item $\e = \cast \SSS \ep$ where $\ep$ is not a mapping.  Then \rulename{C-Cast} applies.
\item $\e = \cast \SSS {\Phi_{\T}}$ where ${\esub \T \SSS}$.  Then \rulename{R-Cast} applies.
\item $\e = \cast \SSS {\Phi_{\T}}$  where ${\proves \empty {\notsubtype \T \SSS}}$.  Then $\e$ is a bad cast.
\end{itemize}

\proofcase{\rulename{T-New}} From the antecedent of \rulename{T-New} the premise of \rulename{R-New} applies.

\proofcase{\rulename{T-Invk}}  Here there are two cases.
\begin{itemize}
\item $\e = \invoke \rr {\mof \dd}$ where $\rr$ is not a mapping.  Then by the induction hypothesis, either $\rr$ contains a bad cast or $\trans \rr \rr'$ and \rulename{C-Rcvr} applies.
\item $\e = \invoke {\Phi_{\T}} {\mof \dd}$  We know from the antecedent that $\mtype \m {\bound \T} = \arrType \oU \V$ and therefore $\mtype \m \T = \arrType \oU \V$ since $\T$ is ground.  Therefore, since {\it mbody} is defined everywhere {\it mtype} is defined, $\mbody \m \T = (\ox, \e_0)$ for some $\ox$ and $\e_0$.  Thus \rulename{R-Invk} applies.
\end{itemize}

\proofcase{\rulename{T-Field}} Here there are two cases, either the reciever is a mapping or not.  In the first, by the antecedent of the typing rule, we can lookup the field successfully and apply \rulename{R-Field}.  Otherwise, we can apply \rulename{C-Field}.

\end{description}

}

\newcommand{\substitutionPreservesTyping}{
\begin{lemma}[Substitution Preserves Typing]
If $\dgprove {\typ \e \T}$ and $\Gamma = \typ{\ox}{\oS}$ and
${\proves {\Delta;\emptyset} {\typ {\od} {\oU}}}$ and ${\dproves {\subtype \oU \oS}}$
then ${\proves {\Delta;\emptyset} {\typ {{\substitute \ox \od} \e} \Tp}}$ where ${\dproves {\subtype{\Tp}{\T}}}$.
\end{lemma}
}

\newcommand{\substitutionPreservesTypingProof}{\unproved}

\newcommand{\methodTypingLemma}
{
\begin{lemma}[Subtyping Preserves Method Typing]
If $\mtype \m \U = \arrType \oT \SSS$ and $\proves \emptyset {\subtype \V \U}$ then
$\mtype \m \V = \arrType \oT \SSS$.
\end{lemma}
}

\newcommand{\methodTypingProof}
{
\myProof By induction over the derivation of $\esub \V \U$.

\begin{description}
\proofcase{\rulename{S-Reflex}}

Trivial.

\proofcase{\rulename{S-Bound}}

Not applicable.

\proofcase{\rulename{S-Super}}

In this case, $\V = \CR$, where $CT(\C) = {\typeFront {\typeArgsBounds \C \X \I \J} \B \A}
\braces{... \oM}$ and $\U = \RforX \A$.  We need to show that $\mtype
\m \CR = \arrType \oT \SSS$.   By lemma $methods$ is well-defined, it
suffices to show that $\simpleMethod \in \methods \CR$.

We know that $\simpleMethod \in \methods \U$.  So, we proceed by
induction on the derivation of $\methods \U$.

\subcase{$\U = \Object$}
Impossible, since $\Object$ has no methods.

\subcase{$\U = \typeArgs \D \W$}
Then, $\U = \RforX \A$.  By \rulename{MethodsClass}, $\methods \V =
\methods \CR = (\RforX \oM)\ \cup\ \methods {\RforX \A} = (\RforX
\oM)\ \cup\ \methods \U$.  Therefore, any method in  $\methods \U$
must be in $\methods \V$.

\proofcase{\rulename{S-Typeof}}

Analagous to \rulename{S-Super}.

\proofcase{\rulename{S-Trans}}  Let the intermediate type be
$\T$. Then $\esub \V \T$ and $\esub \T \U$.  Thus, by the induction
hypothesis, $\mtype \m \V = \mtype \m \T = \mtype \m \U$.

\end{description}
}

\newcommand{\fieldsPreservedLemma}
{
\begin{lemma}[Fields are Preserved by Subtypes]
If $\fields \U = \oF$ and ${\proves {\emptyset} {\subtype \V \U}}$ then $\exists\ \oG$ where $\fields \V = \oF\ @\ \oG$.
\end{lemma}
}

\newcommand{\fieldsPreservedProof}
{

\myProof
We prove this by induction over the derivation for $\fields \V$.

\begin{description}
\proofcase{$\V = \Object$}

Then $\U = \Object$ and $\emptyset = \emptyset\ @\ \emptyset$.

\proofcase{$\V = \CR$}

Then $CT(\C) = {\typeFront {\typeArgsBounds \C \X \I \J} \B \A}
\braces{... {\oH} ...}$

Continue by induction on the derivation of $\esub \V \U$.  The only
interesting cases are \rulename{S-Super} and \rulename{S-Trans}.

\subcase{\rulename {S-Super}}

$\RforX \A = \U$. Thus $\oG = \oH$.

\subcase{\rulename {S-Trans}}

Let $\T$ be the intermediate type. Then $\fields \V = {\fields \T}\ @\
{\ov {\HH'}}$ and $\fields \V = {\fields \T}\ @\ {\ov {\HH''}}$ by the
induction hypothesis.  Thus $\oG = {\ov {\HH'}}\ @\ {\ov {\HH''}}$.

\proofcase{$\V = \typeof{\CS}$}

Then $CT(\C) = {\typeFront {\typeArgsBounds \C \X \I \J} \B \A}
\braces{{\oW\ \oh\ \oe} ...}$.  We continue by induction over the
derivation of $\esub \V \U$.  The only complex cases are
\rulename{S-Kind} and \rulename{S-Trans}.

\subcase{\rulename{S-Trans}}
As above.

\subcase{\rulename {S-Kind}}

$\SforX \A = \U$ Then $\fields \V = \SforX \oW\ \oh$.  By
well-formedness of $\C$, we know that $\fields \B \subseteq \oW\ \oh$.
Then $\SforX \fields \B \subseteq \SforX \oW\ \oh$, and $\fields {\SforX \B} \subseteq
\SforX \oW\ \oh$ by lemma Substitution Distributes over $fields$.
\end{description}

}

\newcommand{\soundnessProofShort}
{
\myProof

Immediate from Subject Reduction and Progress.
}

\newcommand{\fieldsAgreeLemma}{
\begin{lemma}[Agreement of $fields$ and {\it field-vals}]
If $\fields \GG = \Tf$ and $\fieldVals \GG = \oTp\ \ofp\ \oe$ then
$\oT = \oTp$, $\of = \ofp$ and $\eept \oe \oS$ where $\esub \oS \oT$.
\end{lemma}
}
\newcommand{\fieldsAgreeProof}{
\myProof The proof is exactly analagous to the proof of lemma {\it mtype} and {\it mbody} Agree.
}

\newcommand{\noBadCastLemma}
{
\begin{lemma}[Reduction Preserves Safety]
If $\eept \e \T$ and $\trans \e \ep$ then $\e$ contains a bad cast iff $\ep$ does.
\end{lemma}
}

\newcommand{\typeSubstSubtypeLemma}
{
\begin{lemma}[Type Substitution Preserves Subtyping]
If $\Delta = \oX \superSym \oS$ and $\dproves {\subtype \U \T}$
and $\esub \oSp {\SpforX \oS}$
then ${\esub{\SpforX \U} {\SpforX \T}}$.
\end{lemma}
}

\newcommand{\typeSubstSubtypeProof}
{
\myProof By induction over the derivation of $\dproves {\subtype \U \T}$.
\begin{description}

\proofcase{\rulename{S-Reflex}} Immediate.

\proofcase{\rulename{S-Trans}} Let $\V$ be the intermediate type.  By
the induction hypothesis, ${\esub{\SpforX \U} {\SpforX \V}}$ and
${\esub{\SpforX \V} {\SpforX \T}}$ and application of
\rulename{S-Trans} completes the case

\newcommand{\RforY}{\substitute {\overline {\Y}} {\overline {\R}}}

\proofcase{\rulename{S-Super}} In this case $\U = \CR$ and $CT(\C) =
{\typeDefShort {\typeArgsBounds \C \Y \I \J} \B \A}$ and $\RforY \A =
\T$.  We must show that $\esub {\SpforX \CR} {\SpforX {\RforY \A}}$.
But note that:

$\SpforX \CR = \typeArg \C {\SpforX {\ov \R}} = \typeArg \C {\SpforX {\RforY {\ov \Y}}} =
\typeArg \C {{\substitute {\ov \Y} {\SpforX {\ov \R}}} {\ov \Y}}$.

Also, ${\SpforX {\RforY \A}} = {\substitute {\ov \Y} {\SpforX {\ov \R}}} \A$.  So,
$\esub {\typeArg \C {\SpforX {\ov \R}}} {\substitute {\ov \Y} {\SpforX {\ov \R}}} \A$
by \rulename{S-Super}, completing the case.

\proofcase{\rulename{S-Kind}} Analagous to \rulename{S-Super}.

\proofcase{\rulename{S-Bound}} In this case, $\U = \X_i$ and $\T =
\SSS_i$.  Then ${\SpforX \U} = \Sp_i$ and $\SpforX \T = \SpforX
\SSS_i$.  We are given $\esub {\Sp_i} {\SpforX \SSS_i}$, completing the case.

\end{description}
}

\newcommand{\typeSubstTypeLemma}
{
\begin{lemma}[Type Substitution Preserves Typing]
If $\Delta = \oX \superSym \oS$ and $\dgprove {\typ \e \T}$ and $\esub
\oSp {\SpforX \oS}$ then ${\eept {\SpforX \e} {\SpforX \T}}$.

\end{lemma}
}

\newcommand{\typeSubstTypeProof}
{
\unproved
}

\newcommand{\substmbodyLemma}
{
\begin{lemma}[Substitution Preserves $mbody$ Soundness]
If $~\simpleMethod \okin\ \GG$ and $\mbody{\m}{\GG} = (\ox,\e_0)$ then
${\proves {\emptyset;{\typ {\this} {\GG}},{\typ{\ox}{\oT}}} {\typ{\e_0}{\Up}}}$ where ${\proves {\emptyset} {\subtype \Up {\SforX \U}}}$.
\end{lemma}
}
\newcommand{\substmbodyProof}{\unproved}

\newcommand{\extensionMethodOKLemma}{
\begin{lemma}[Extension Preserves Method OKness]
If $\M \okin\ \GG$ and $\super \Gp = \GG$ then $\M \okin\ \Gp$.
\end{lemma}
}

\newcommand{\methodsDefinedLemma}{
\begin{lemma}[$methods$ is well-defined]
If ${\methodDef \U \m \T \x {\e}} \in \methods \GG$ and ${\methodDef \Up \m \Tp \x {\e}} \in \methods \GG$
then $\T = \Tp$ and $\U = \Up$.
\end{lemma}
}

\newcommand{\methodsDefinedProof}
{
\myProof Immediate from the side constraint on class tables that no two methods share the same name.
}

\newcommand{\extensionMethodOKProof}{\unproved}

\newcommand{\mtypembodyLemma}
{
\label{lem:agree}
\begin{lemma}[$mtype$ and $mbody$ agree]
If ${\mbody \m \GG} = (\ox, \e)$ and ${\mtype \m \GG} = \arrType \oT \U$ then
$\proves {\emptyset;{\typ \this \GG}, {\typ{\oX}{\oT}}} {\typ \e \Up}$ where $\proves \emptyset {\subtype \Up \U}$.
\end{lemma}
}

\newcommand{\substOKLemma}
{
\begin{lemma}[Type Substitution Preserves Type OKness]
If $\proves {\oX \superSym \oG} \T \ok$ and $\esub \oH \oG$ and $\proves \emptyset {\oH \ok}$
then $\proves \emptyset {{{\substitute \oX \oH} \T} \ok}$.
\end{lemma}
}

\newcommand{\substOKProof}
{
\myProof By induction over the derivation of $\proves {\oX \superSym \oG} \T \ok$.

\begin{description}

\proofcase{\rulename{WF-Object}} Immediate.

\proofcase{\rulename{WF-TypeOf}} Immediate from the induction
hypothesis.

\proofcase{\rulename{WF-Var}} In this case, $\T = \X_i$ and
${\substitute \oX \oH} \X_i = \HH_i$.  And we are given that $\proves
\emptyset {\oH \ok}$.

\proofcase{\rulename{WF-Class}} Immediate from lemma Type
Substitution Preserves Subtyping and the induction hypothesis.

\end{description}
}

\newcommand{\substDistFieldsLemma}
{
\begin{lemma}[Substitution Distributes over $fields$]
$\RforX {\fields \GG} = \fields {\RforX \GG}$
\end{lemma}
}

\newcommand{\substDistFieldsProof}
{
\myProof By induction over the derivation of $\fields \GG$.

\begin{description}
\proofcase{\rulename{F-Object}}

Trivial.

\proofcase{\rulename{F-Class}}

\newcommand{\SforY}{\substitute {\overline {\Y}} {\overline {\SSS}}}

Let $\GG = \CS$  Then $CT(\C) = {\typeFront {\typeArgsBounds \C \Y \I \J} \B \A}
\braces{... \oW\ \og}$, $\fields {\SforY A} = \oU\ \oh$ and $\fields \CS
= \oU\ \oh\ @\ {\SforY {\oW\ \og}}$.

Thus $\RforX \fields \CS = \RforX (\oU\ \oh\ @\ \SforY \oW\ \og) =
(\RforX \oU\ \oh)\ @\ (\RforX \SforY \oW\ \og) =
\fields {\RforX \CS}$.

\proofcase{\rulename{F-Typeof}}

In this case, $\GG = \typeof \CS$.  Then $CT(\C) = {\typeFront
{\typeArgsBounds \C \Y \I \J} \B \A} \braces{{\Tfe} ...}$ and $\fields
{\typeof \CS} = {\SforY}\Tf$. Therefore, $\RforX \fields {\typeof \CS}
= \RforX {\SforY}\Tf$.  But note that $\fields {\RforX {\typeof \CS}}
= \fields {\typeof {\typeArgs \C {\RforX \oS}}} = {\substitute \oY
{\RforX \oS}}\Tf = \RforX {\SforY}\Tf$.

\end{description}

}

### 2004-09-29 classes/711/mp10/test.py

@@ -1,0 +1,29 @@
+x = 100
+z = 0
+u = 0
+v = 0
+w = 1
+d = 2
+dd = x
+while (dd > 0):
+    w = w+w+w+w
+    d = d+d
+    dd = dd/2
+
+while d > 0:
+    if ((u + v + w) == x):
+        z = z+d/2
+        d = 0
+    elif (u + v + w > x):
+        d = d/2
+        u = u
+        v = v/2
+        w = w/4
+    else:
+        z = z+d/2
+        d = d/2
+        u = u + v + w
+        v = v/2 + w
+        w = w/4
+
+print z

### 2004-09-29 classes/hopl/logic/test.pl

@@ -1,0 +1,11 @@
+member2(X, [X | _]).
+member2(X, [_ | Xs]) :- member(X, Xs).
+sublist2(X,Y) :- append(_,Z,Y), append(X,_,Z).
+
+:- op(1100,yfx,arrow).
+
+type(E, var(X), T) :- member([X,T], E).
+type(E, apply(M,N), T) :- type(E,M,(S arrow T)), type(E,N,S).
+type(E, lambda(X,M), (S arrow T)) :- type([[X,S] | E],M,T).
+
+

### 2004-09-29 larceny/larceny.sh

@@ -1,0 +1,44 @@
+#!/bin/sh
+# Copyright 1998 Lars T Hansen.
+#
+# $Id: larceny,v 1.3 1998/12/18 22:31:05 lth Exp $
+#
+# Shell script to run larceny.
+#
+# Usage: larceny [larceny-options ...] [program-options ...]
+# where larceny-options are valid options for larceny, and program-
+# options are anything else.  If you want to pass an option
+# that is a valid larceny-option as a program-option, then use -args
+# to separate program options and larceny options.
+#
+# Additionally, the larceny option -small selects the interpreter-only heap.
+#
+# For example,
+#  larceny -np -foo         ; -np is a larc.opt., -foo a prog.opt.
+#  larceny -np -args -load  ; -np is a larc.opt., -args -load a prog.opt.
+
+# EDIT THIS PATH TO SUIT YOUR SYSTEM.
+LARCENY_PATH=/home/samth/work/larceny-pnk
+unset LARGS
+unset PARGS
+SMALL=0
+while [ "$1" != "" ]; do 
+  case $1 in
+    -annoy-user | -annoy-user-greatly | -gen | -help | -nobreak | -nostatic | \
+    -np | -quiet | -reorganize-and-dump | -stats | -step | -stopcopy)
+      LARGS="$LARGS $1"; shift 
+      ;;
+    -areas | -load | -max | -rhash | -size* | -ssb | -steps | -stepsize | \
+    -ticks)
+      LARGS="$LARGS $1 $2"; shift; shift 
+      ;;
+    -small) SMALL=1; shift ;;
+    -args) PARGS="$@"; break ;;
+    *) PARGS="-args $@"; break ;;
+  esac
+done
+if [ $SMALL = 0 ]; then
+  exec $LARCENY_PATH/larceny.bin $LARGS $LARCENY_PATH/larceny.heap $PARGS
+else
+  exec $LARCENY_PATH/larceny.bin $LARGS $LARCENY_PATH/r5rs.heap $PARGS
+fi

### 2004-09-29 larceny/twobit.sh

@@ -1,0 +1,37 @@
+#!/bin/sh
+# Copyright 1998 Lars T Hansen.
+#
+# $Id: twobit,v 1.3 1998/12/18 22:31:07 lth Exp $
+#
+# Shell script to run twobit.
+#
+# Usage: twobit [larceny-options ...] [program-options ...]
+# where larceny-options are valid options for larceny, and program-
+# options are anything else.  If you want to pass an option
+# that is a valid larceny-option as a program-option, then use -args
+# to separate program options and larceny options.
+#
+# For example,
+#  larceny -np -foo         ; -np is a larc.opt., -foo a prog.opt.
+#  larceny -np -args -load  ; -np is a larc.opt., -args -load a prog.opt.
+
+# EDIT THIS PATH TO SUIT YOUR SYSTEM.
+LARCENY_PATH=/home/samth/work/larceny-pnk
+unset LARGS
+unset PARGS
+SMALL=0
+while [ "$1" != "" ]; do 
+  case $1 in
+    -annoy-user | -annoy-user-greatly | -gen | -help | -nobreak | -nostatic | \
+    -np | -quiet | -reorganize-and-dump | -stats | -step | -stopcopy)
+      LARGS="$LARGS $1"; shift 
+      ;;
+    -areas | -load | -max | -rhash | -size* | -ssb | -steps | -stepsize | \
+    -ticks)
+      LARGS="$LARGS $1 $2"; shift; shift 
+      ;;
+    -args) PARGS="$@"; break ;;
+    *) PARGS="-args $@"; break ;;
+  esac
+done
+exec $LARCENY_PATH/larceny.bin $LARGS $LARCENY_PATH/twobit.heap $PARGS

### 2004-09-29 larceny/larceny.sh
fixed up paths

@@ -18,7 +18,7 @@
 #  larceny -np -args -load  ; -np is a larc.opt., -args -load a prog.opt.
 
 # EDIT THIS PATH TO SUIT YOUR SYSTEM.
-LARCENY_PATH=/home/samth/work/larceny-pnk
+LARCENY_PATH=/scratch/samth/larceny
 unset LARGS
 unset PARGS
 SMALL=0

### 2004-09-29 larceny/twobit.sh

@@ -16,7 +16,7 @@
 #  larceny -np -args -load  ; -np is a larc.opt., -args -load a prog.opt.
 
 # EDIT THIS PATH TO SUIT YOUR SYSTEM.
-LARCENY_PATH=/home/samth/work/larceny-pnk
+LARCENY_PATH=/scratch/samth/larceny
 unset LARGS
 unset PARGS
 SMALL=0

### 2004-12-10 metaclasses/paper.tex
Eric's edits to FOOL paper

@@
-\documentclass{acmconfbig}
+\documentclass{acm-sigplan}
 
@@
 
-\author{Sam Tobin-Hochstadt \quad Eric Allen \\
-\begin{tabular}{c}
-\\
-Sun Microsystems Laboratories \\
-Burlington MA 01803 \\
-{\tt \large <first>.<last>@sun.com} \\
-\end{tabular}
-}
+\authorinfo{Sam Tobin-Hochstadt}{Northeastern University}{samth@ccs.neu.edu}
+\authorinfo{Eric Allen}{Sun Microsystems Laboratories}{eric.allen@sun.com}
+
+%\authorinfo{Sam Tobin-Hochstadt \quad Eric Allen \\
+%\begin{tabular}{c}
+%\\
+%Sun Microsystems Laboratories \\
+%Burlington MA 01803 \\
+%{\tt \large <first>.<last>@sun.com} \\
+%\end{tabular}
+%}
 
@@
 \begin{abstract}
-Metaclasses provide a useful method of abstraction for programmers
-working with object-oriented languages, but they have not seen the
-formal exploration applied to more conventional object-oriented
-programming features.  In order to elucidate the structure of
-metaclasses and their relationship with static typing, we present a
+Metaclasses provide a useful mechanism for abstraction in
+object-oriented languages.  But most languages supporting metaclasses
+have imposed severe restrictions on their use. Typically, a metaclass
+is allowed to have only a single instance and all metaclasses are
+required to share a common superclass \cite{objvlisp}. In addition,
+few languages supporting metaclasses have included a static type
+system, and none have included a type system with nominal subtyping
+(i.e., subtyping as defined in mainstream languages such as the
+Java${^{\mbox{\tiny{TM}}}}$ Programming Language or C\#).
+
+In order to elucidate the structure of
+metaclasses and their relationship with static types, we present a
 core calculus for a nominally-typed object-oriented language with
-metaclasses and prove type soundness over this core.  This calculus is
+metaclasses and prove type soundness over this core. 
+To our knowledge, no previous formalization of metaclasses in a 
+language with nominal subtyping exists.
+This calculus is
 presented as an adaptation of Featherweight GJ \cite{FJ}, and is
@@
 length.  However, modeling this relationship in a conventional
-language such as the Java${^{\mbox{\tiny{TM}}}}$ Programming Language
-is impossible without attributing to some of these concepts properties
-which do not properly belong to them.  It is natural to model {\txt 3
-feet} as an instance of class {\txt Length}. But if we were to define
-a class {\txt Dimension} and define {\txt Length} to be an instance of
-class {\txt Dimension}, then {\txt Length} would not be a class and so
-{\txt 3 feet} could not be an instance of {\txt
-Length}. Alternatively, we might define {\txt Length} to be a subclass
-of {\txt Dimension}. But then we still cannot define {\txt 3 feet} to
-be an instance of {\txt Length} because anything that is an instance
-of {\txt Length} would also be an instance of class {\txt Dimension}
-and {\txt 3 feet} is obviously not a dimension.
-
-The fundamental limitation of conventional object-oriented languages
-is that they do not allow a single concept to serve both as a class
-and as an instance.  In \cite{FW96}, Welty and Ferrucci lay out a
-comprehensive case that this limitation prevents conventional modeling
-languages from capturing many important aspects of the world.  In
-their primary example, they show the difficulties in modeling a simple
+language such as the Java Programming Language is difficult without
+attributing to some of these concepts properties which do not properly
+belong to them.  It is natural to model {\txt 3 feet} as an instance
+of a class {\txt Length}. But if we were to define a class {\txt
+Dimension} and define {\txt Length} to be an instance of class {\txt
+Dimension}, then {\txt Length} could not be a class and so {\txt 3
+feet} could not be an instance of {\txt Length}. Alternatively, we
+might define {\txt Length} to be a subclass of {\txt Dimension}. But
+then we still cannot define {\txt 3 feet} to be an instance of {\txt
+Length} because anything that is an instance of {\txt Length} would
+also be an instance of class {\txt Dimension} and {\txt 3 feet} is
+obviously not a dimension. The key limitation preventing us from
+modeling this domain properly is that we cannot define a single
+concept to serve both as a class and as an instance of a class.
+
+Another example is provided by Welty and Ferrucci in \cite{FW96}, where
+they show the difficulties in modeling a simple
 ontology including the notions {\txt Species} and {\txt Eagle}, while
@@
 notions {\txt the Eagle} and {\txt Eagle}.
+Again, the limitation is that we cannot define a class as an instance of another class.
+%Welty and Ferrucci proceed to explain how this limitation prevents conventional modeling
+%languages from capturing many important aspects of the world.  
 %The languages Ferrucci and Welty discuss cannot accommodate multiple
@@
 
-In order to overcome these limitations in expressiveness, several
+In order to overcome this limitation in expressiveness, several
 object-oriented languages that are not statically typed, such as
-Smalltalk, Python, and Self, allow for more flexible class
+Smalltalk, Python, and Self, allow for more flexible object
 relationships \cite{Smalltalk80}, \cite{ArtMOP}, \cite{PythonMeta}. In
@@
 metaclasses. Static members of a class are modeled as ordinary members
-of the corresponding metaclass.  But traditional metaclass systems
-have placed important limitations on expressiveness.  For example, in
+of the corresponding metaclass.  But in both of these languages there
+are important limitations on expressiveness.  For example, in
 Smalltalk, the metaclass hierarchy is only two levels deep.  If we
@@
 instantiation consists of cloning an existing instance. The members of
-the clone can be modified and added to at will, making static checking
-difficult.
-
-An extremely expressive system of metaclasses, very similar to ours,
-was presented by Cointe in \cite{objvlisp}.  In his paper, he makes a
-number of persuasive arguments for the benefits of metaclasses.
-In response to complaints that the Smalltalk80 metaclass system was
-too complex, he suggested that a more powerful and flexible system, as
-he and we present, has the potential to be much simpler.
-He claims ``The ObjVlisp model's primary advantage is uniformity.''
-We agree, and believe that the treatment of metaclasses presented here
-unifies both disparate metaclass systems, and different ad-hoc
-mechanisms describing the properties of classes as objects in
-languages without metaclasses.
-
-Cointe also quotes from \cite{cardelli:multiple}: ``With respect to
-Simula, Smalltalk also abandons ... strong typing, allowing it to
-... introduce the notion of meta-classes''.  In this paper, we show
-that such a compromise is not necessary, and that static typing can
-coexist with the flexibility of metaclasses.
-
-We present and formalize a statically typed calculus with nominal
-subtyping for metaclasses where every class can serve as both an
-instance of a class and as a class of instances, and there is no
-limitation on the levels of nesting of metaclasses. Our work is
-motivated by \cite{DimUnits}, which presents a system for integrating
-the static checking of dimensions of physical quantities with a
-conventional object-oriented type system. In that system, metaclasses
-are used extensively to model the type relationships of dimensions.
-The notion of metaclass in that system is more general than that of
-languages such as Smalltalk, where each class has a unique metaclass
-that defines its class members. In this paper, we formalize the core
-of that system related to metaclasses and prove it sound.  In the
-process, we discover several interesting properties about the
-structure that a nominal type system for metaclasses must take.
+the clone can be modified and added to at will, dispensing with class
+relationships entirely and making static checking difficult.
+
+A more expressive metaclass system in ObjVlisp, described informally
+by Conte in \cite{objvlisp}.  In ObjVlisp, there is no restriction on
+the number of instances of a metaclass, and metaclasses are not
+required to share a common superclass. The instance methods of a
+metaclass are inherited as the static members of its instances.  Conte
+makes a number of arguments for the benefits of generalized
+metaclasses.  In response to complaints that the Smalltalk80 metaclass
+system is too complex, he suggests that a more powerful and flexible
+system has the potential to be much simpler.
+%, arguing that ``The ObjVlisp model's primary advantage is uniformity.''
+However, ObjVlisp does not include a static type system. 
+%Cointe quotes from \cite{cardelli:multiple}: ``With respect to
+%Simula, Smalltalk also abandons ... strong typing, allowing it to
+%... introduce the notion of meta-classes''.   
+In this paper, we show that this limitation is not necessary. We
+present a semantics for a statically typed calculus of metaclasses
+where, as in ObjVlisp, every class can serve as both an instance of a
+class and as a class of instances, and there is no limitation on the
+levels of nesting of metaclasses. In the process, we also reveal
+several interesting properties about the structure that a nominal type
+system for metaclasses must take. We then present a formal operational
+semantics for this calculus, as well as a proof of type soundness.
+
+Our work is motivated by \cite{DimUnits}, which presents a system for
+integrating the static checking of dimensions of physical quantities
+in MetaGen, an extension of the Java Programming Language with support
+for metaclasses. In MetaGen, metaclasses are used to model the type
+relationships of dimensions.  As with ObjVlisp, the notion of a
+metaclass in MetaGen is more general than that of languages such as
+Smalltalk, but, unlike ObjVlisp, MetaGen includes a static type system
+\cite{DimUnits}. The calculus in this paper can be viewed as a
+formalization of a purely functional core of MetaGen.
 
@@
 following discussion we restrict ourselves to simple examples, and do
-not delve in the the complexities of object-oriented analysis, or of
-dimension types.  This is certainly not because we believe that
+not delve into the complexities of object-oriented analysis with metaclasses.  
+This is certainly not because we believe that
 metaclasses are most useful for the development of toy object-oriented
-programs.  On the contrary, metaclasses are likely to show their value
+programs.  On the contrary, metaclasses are most likely to show their value
 in large systems, both in terms of modeling and in the treatment of
@@
 patterns'' is obviated through the use of metaclasses. We then
-describe its formal properties in section \ref{formal}.  In section
+describe the formal properties of MCJ in section \ref{formal}.  In section
 \ref{formalsem} we provide a formal semantics for the language, and in
@@
 are distinguished from instance members by order, and with the keyword
-{\txt class} \footnote{This syntax is not ambiguous since MCJ has no
+{\txt class} \footnote{This syntax is not ambiguous because MCJ has no
 inner classes.}.  Members of a class definition are laid out as
@@
 If the method call were instead $\invoke \bb {\mof {}}$, where {\txt
-b} denotes an instance of {\txt B}, then the instance methods of first
-$\B$ then $\A$ would be examined.
+b} denotes an ordinary (non-class) instance of {\txt B}, then the
+instance methods of first $\B$ then $\A$ would be examined.
 
@@
 that return new instances of the class, but each such method must
-ultimately includes a {\txt new} expression, which can occur only
+ultimately include a {\txt new} expression, which can occur only
 within the scope of the class whose instance it returns. A {\txt new}
@@
 Given that class references can be used as expressions, it is natural
-to ask: what is the type of a class reference?  In the world of the
-$\lambda$-calculus, such types are known as kinds.  In MCJ, however,
-the kind of a class does not capture all of the properties of the
-class as a value.  For example, a class may add new class fields or
-class methods which are not present in the kind.  Therefore, each
-class freely generates a new type, which is the type of the class
-considered as a value.  We represent this type with the {\txt typeOf}
-operator, which is produces a type that is not also a class.  Since
-this type is not a class, it cannot be the superclass or kind of
-another class, and it cannot serve as the instantiation of a generic
-type parameter.
+to ask: what is the type of a class reference?  
+%In the world of the
+%$\lambda$-calculus, such types are known as kinds.  
+In MCJ, however, the kind of a class does not capture all of the
+properties of the class as a value.  For example, a class may add new
+class fields or class methods which are not present in the kind.
+Therefore, each class freely generates a new type, which is the type
+of the class considered as a value.  We represent this type with the
+{\txt typeOf} operator, which takes a class instantiation and produces
+a type that is not also a class.  Because this type is not a class, it
+cannot be the superclass or kind of another class, and it cannot serve
+as the instantiation of a generic type parameter.
 
@@
 
+\hspace{-.43cm}where we use the convention that {\txt A:B} denotes
+that {\txt A} has type {\txt B}, and {\txt A<:B} denotes that {\txt A}
+is a subtype of {\txt B}.
+
 The name {\txt typeOf} has been used in other contexts to refer to a
 runtime operation that determines the dynamic type of a object.
-However, despite the name similarity, these functions bear no
+However, despite the name similarity, this function bears no
 relationship to the operator described here.  In MCJ, an application
@@
 The standard subclassing relationship, like subtyping, is transitive.
-That is, if $\B$ is subclass of $\C$ and $\C$ is a subclass of $\D$,
+That is, if $\B$ is a subclass of $\C$ and $\C$ is a subclass of $\D$,
 then $\B$ is a subclass of $\D$.  However, this relationship does not
-hold for {\txt instanceof} relationships.  Therefore, if if $\B$ is an
-instance of $\C$ and $\C$ is an instance of $\D$, then no judgment
+hold for {\txt instanceof} relationships.  In general, if $\B$ is an
+instance of $\C$ and $\C$ is an instance of $\D$, no judgment
 about the relationship of $\B$ to $\D$ can be inferred.
 
-\paragraph{Cyclic class hierarchies}
-One property of metaclasses that we have not yet discussed is that
-cycles can occur among instance class relationships. The simplest such
-cycle is:
+Another difference between subclassing and {\txt instanceof}
+relationships is that cycles can occur among instance class
+relationships. The simplest such cycle is:
 
@@
 
-Having seen the outlines of MCJ, we return to the motivational
+Having seen this outline of MCJ, we return to the motivational
 examples discussed earlier.  First, the relationships between
@@
 that species.  The fundamental nature of species has a two-level
-containment relationship, which the type system is failing to express.
-This means that our type checker cannot determine what we want, and
+containment relationship, which the type system fails to express.
+This means that our type checker cannot determine what we want and
 cannot help us avoid mistakes.
@@
 Another problem, which is in some ways just a symptom of the first
-problem, but which bears special attention, is the use of generic
+problem but which bears special attention, is the use of generic
 types.  Imagine the following method, where T is a type parameter in
 scope:\footnote{Generic methods would be straightforward to add to
-MCJ, but we excluded them to simplify the presentation of the type
-system, and since they do not add complexity to the proofs.}
+MCJ, but we exclude them to simplify the presentation of the type
+system and to focus on the important interactions of metaclasses with
+other language features.}
 
@@
 
-Design Patterns, as exemplified in \cite{GOF}, have had a very
+Design Patterns, as exemplified in \cite{GOF}, have had a
 substantial impact on the world of object-oriented programming and
@@
 One of the advantages of our metaclass framework is that it allows
-clean expression of several patterns in the language, rather than
-adding additional abstraction techniques on top of the language.  For
-example, a number of the classic design patterns deal with object
-creation, as discussed above in section \ref{privconst}.  It is clear
-that the concept of a Factory method is easily (and necessarily)
-expressed in MCJ.  However, a number of other patterns also find easy
-expression.  The Abstract Factory pattern is a simple application of
-inheritance in the metaclass hierarchy.  The Prototype pattern is, as
-mentioned in the Design Patterns book, supported in languages with
-metaclasses.  Finally, there is no need for the Singleton pattern,
-since the class itself can serve as a singleton.  While this is
-possible in the Java Programming Language and in \Cpp, it is
-undesirable since the static behavior that would need to be used
-cannot inherit from another class.  In fact, the inflexibility of
-class or static operations is one of the rationales cited for the
-Singleton pattern.
+clean expression of several common design patterns in the language,
+rather than requiring that they be expressed with abstraction
+techniques on top of the language.  For example, a number of the
+classic design patterns deal with object creation, as discussed above
+in section \ref{privconst}.  It is clear that the concept of a Factory
+method is easily (and necessarily) expressed in MCJ.  However, a
+number of other patterns also find easy expression.  The Abstract
+Factory pattern is a simple application of inheritance in the
+metaclass hierarchy.  The Prototype pattern is, as mentioned in the
+Design Patterns book, supported in languages with metaclasses.
+Finally, there is no need for the Singleton pattern; a class itself
+can serve as a singleton.  While classes can be used as limited
+singletons in the Java Programming Language and in \Cpp, doing so is
+undesirable because the static behavior of a class cannot inherit from
+another class.  In fact, the inflexibility of class or static
+operations is one of the rationales cited for the Singleton pattern.
 
@@
 programmer must understand in order to use the language productively.
-
-An alternative solution explored in this paper is to provide additional
-flexibility in the language, which does not allow for arbitrary
+In contrast, the inclusion of metaclasses does not allow for arbitrary
 additional abstraction, but instead makes the abstractions commonly
@@
 that a class $\Class$ would need no consideration from the type
-system.  Since both $\Class$ and $\Object$ would sit atop the
-hierarchy with no contents, we determined there was no need to include
+system.  Because both $\Class$ and $\Object$ would sit atop the
+hierarchy with no contents, there is no need to include
 both of them.  In a more substantial language, where $\Object$ might
 have methods or fields, there might be a use for a distinguished
-$\Class$ class.  However, it would not need any special treatment by
+$\Class$ class, but it would not need any special treatment by
 the type system.
@@
 {\txt make} is a reference to the variable {\txt b}, which has the
-obvious value.  The final directly expressible expression is the cast,
-which works as follows: {\txt (Object)O.make()} is an expression of
-type \Object that evaluates to an instance of {\txt O}.
+obvious value.  The final form of expression writable by the
+programmer is the cast, with the usual semantics. For example, {\txt
+(Object)O.make()}\ is an expression of type \Object that evaluates to
+an instance of {\txt O}.
 
@@
 
-We now discuss several non-obvious aspects of our type system resulting from the need to statically check uses of metaclasses.
+We now discuss several non-obvious aspects of our type system
+resulting from the need to statically check uses of metaclasses.
 
@@
 soundness. Given the simplicity we are able to achieve in the
-definitions, the proof is not significantly more complicated than the
+definitions, the proof is not significantly more complex than the
 soundness proof given for Featherweight GJ\ \cite{FJ}.  However, there
 are a number of significant lemmas, of which we list the most
-important:
+important. The full proof is available at:
+
+\vspace{.1cm}
+\verb|http://research.sun.com/projects/plrg/|
 
@@
 
-The metaclass system present in Python \cite{PythonMeta} is quite
-similar to that provided here, with inheritance of instance methods as
-class methods from arbitrary other classes as metaclasses.  However,
+The metaclass system present in Python \cite{PythonMeta} is
+similar to that provided here, where classes inheritance the instance methods
+of their metaclasses as class methods.  However,
 Python is dynamically-typed, and many of the uses to which Python
@@
 which turns Smalltalk into a structurally-typed language with
-extensive static checking.  However, because the Smalltalk metaclass
+static checking.  However, because the Smalltalk metaclass
 system is so different from the one in MCJ, many of the interesting
@@
 
-Cointe \cite{objvlisp} presents a model of metaclasses extremely
-similar to that presented here.  In it, he provides several
+Cointe \cite{objvlisp} presents a model of metaclasses
+similar to that presented here, but for a dynamically typed object-oriented
+language based on Lisp.  In it, he provides several
 overlapping motivations to our own.  One is to regularize the
@@
 work, however, does not provide a formal model, so it is difficult to
-determine the exact relationship between the systems.  Additionally,
+determine the exact relationship between the two systems.  Additionally,
 his work is in a untyped setting (Lisp and Smalltalk) and thus the
@@
 
-We would like to thank Jan-Willem Maessen for his extremely helpful
+We would like to thank Jan-Willem Maessen for his 
 feedback on the formal rules presented in this paper.  We also thank

### 2004-12-12 metaclasses/paper.tex
lots of paper updates

@@
 However, most such languages are unable to model many simple and
-common relationships. For example, consider the relationship between
+common relationships.
+
+An excellent example is provided by Welty and Ferrucci in \cite{FW96}, where
+they show the difficulties in modeling a simple
+ontology including the notions {\txt{Species}} and {\txt{Eagle}}, while
+also capturing the relationship between {\txt{Eagle}} and {\txt{Harry}},
+a particular eagle.
+
+To see the problem, consider a system with a class \txt{Eagle} and an
+instance \txt{Harry}.  The static type system can ensure that any
+messages that \txt{Harry} responds to are also understood by any other
+instance of \txt{Eagle}.
+
+However, even in a language where classes themselves can have methods,
+such relationships cannot be expressed.  For example, in a language
+with static methods, we can write the expression
+\txt{Eagle.isEndangered()}
+
+However, if \txt{Salmon} is a species, our type system provides us no
+way of requiring that it also have an \txt{isEndangered} method.  The
+solution that Ferruci and Welty suggest, and that we develop in this
+paper, is to allow a consistent treatment of both instances such as
+\txt{Harry} and classes such as \txt{Eagle}.
+
+One solution they consider and reject is to separate the notion of
+``eagle'' into two concepts: {\txt{the Eagle}} and {\txt{Eagle}}, with
+the former an instance of {\txt{Species}} and the latter a class whose
+instances are specific birds.  This approach is inadequate because it
+fails to express a relationship between the notions {\txt{the Eagle}}
+and {\txt{Eagle}}.  Therefore, if we create another species such as
+\txt{Salmon} in our universe, we must define one new class and one new
+instance.  This adds new invariants which the programmer must
+maintain, with no help from the type system.  Instead, we propose that
+the type system should allow the programmer to express the desired
+relationship directly, and enforce the invariants naturally.
+
+For example, consider the relationship between
 physical quantities, units of measurement, and physical dimensions. It
@@
 attributing to some of these concepts properties which do not properly
-belong to them.  It is natural to model {\txt 3 feet} as an instance
-of a class {\txt Length}. But if we were to define a class {\txt
-Dimension} and define {\txt Length} to be an instance of class {\txt
-Dimension}, then {\txt Length} could not be a class and so {\txt 3
-feet} could not be an instance of {\txt Length}. Alternatively, we
-might define {\txt Length} to be a subclass of {\txt Dimension}. But
+belong to them.  It is natural to model {\txt{3 feet}} as an instance
+of a class {\txt{Length}}. But if we were to define a class {\txt
+Dimension} and define {\txt{Length}} to be an instance of class {\txt
+Dimension}, then {\txt{Length}} could not be a class and so {\txt 3
+feet} could not be an instance of {\txt{Length}}. Alternatively, we
+might define {\txt{Length}} to be a subclass of {\txt{Dimension}}. But
 then we still cannot define {\txt 3 feet} to be an instance of {\txt
-Length} because anything that is an instance of {\txt Length} would
-also be an instance of class {\txt Dimension} and {\txt 3 feet} is
+Length} because anything that is an instance of {\txt{Length}} would
+also be an instance of class {\txt{Dimension}} and {\txt 3 feet} is
 obviously not a dimension. The key limitation preventing us from
@@
 
-Another example is provided by Welty and Ferrucci in \cite{FW96}, where
-they show the difficulties in modeling a simple
-ontology including the notions {\txt Species} and {\txt Eagle}, while
-also capturing the relationship between {\txt Eagle} and {\txt Harry},
-a particular eagle.  One solution they consider and reject is to
-separate the notion of ``eagle'' into two concepts: {\txt the Eagle}
-and {\txt Eagle}, with the former an instance of {\txt Species} and
-the latter a class whose instances are specific birds.  This approach
-is inadequate because it fails to express a relationship between the
-notions {\txt the Eagle} and {\txt Eagle}.
 Again, the limitation is that we cannot define a class as an instance of another class.
@@
 A more expressive metaclass system in ObjVlisp, described informally
-by Conte in \cite{objvlisp}.  In ObjVlisp, there is no restriction on
+by Cointe in \cite{objvlisp}.  In ObjVlisp, there is no restriction on
 the number of instances of a metaclass, and metaclasses are not
@@
 
-MCJ is a generically-typed object-oriented calculus based on
-Featherweight GJ that includes metaclasses. An MCJ program consists of
+MCJ is a object-oriented calculus with generic types based on
+Featherweight GJ which includes metaclasses. An MCJ program consists of
 a sequence of class definitions followed by a trailing
@@
 Each class is an instance of a \emph{metaclass}.  A metaclass is
-either a user-defined class or class {\txt Object}.  We say that if a
+either a user-defined class or class {\txt{Object}}.  We say that if a
 class {\txt C} is an instance of a class {\txt D}, then {\txt C} is an
@@
 For example, the following header declares a class {\txt C} with
-superclass {\txt Object} and kind {\txt D} and no type parameters:
+superclass {\txt{Object}} and kind {\txt D} and no type parameters:
 
@@
 \end{verbatim}
+\footnote{In other examples, we will leave out the \kind\ and \extends\
+declaration when they refer to \Object.}
 
@@
 are distinguished from instance members by order, and with the keyword
-{\txt class} \footnote{This syntax is not ambiguous because MCJ has no
+\class \footnote{This syntax is not ambiguous because MCJ has no
 inner classes.}.  Members of a class definition are laid out as
@@
 \begin{verbatim}
-class C kind Object extends Object {
+class C {
    class Object m() {...}
@@
 the class members of {\txt C}.  As in \cite{DimUnits}, a class is
-allowed to define both a static method and an instance method of the
+allowed to define both a class method and an instance method of the
 same name. We prevent ambiguity in member references by requiring that
@@
 \begin{verbatim}
-class A kind Object extends Object {...}
-class B kind Object extends A {...}
-class C kind B extends Object {...}
+class A {...}
+class B extends A {...}
+class C kind B {...}
 \end{verbatim}
@@
 support first-class genericity because a naked type variable must not
-appear in the {\txt extends} clause or {\txt kind} clause of a class
-definition and {\txt new} expressions on naked type variables are not
-supported. Also, polymorphic methods (which are orthogonal to the
-features we explore) are not supported.
+appear in the \extends\ clause or \kind\ clause of a class definition
+and {\txt{new}} expressions on naked type variables are not
+supported.\footnote{First-class genericity is difficult to integrate
+with our presentation of constructors.} Also, polymorphic methods
+(which are orthogonal to the features we explore) are not supported.
 
@@
 \begin{verbatim}
-class C<T extends (D,E)>
-   kind Object
-   extends Object {...}
+class C<T extends D kind E> {...}
 \end{verbatim}
@@
 provided with first-class genericity \cite{NextGen, MixGen}, but
-obviates the need for {\txt with} clauses because the bound on the
+obviates the need for {\txt{with}} clauses because the bound on the
 kind of a type variable stipulates what factory methods can be called
-on it.  A class can define class methods with arbitrary signatures
-that return new instances of the class, but each such method must
-ultimately include a {\txt new} expression, which can occur only
-within the scope of the class whose instance it returns. A {\txt new}
-expression looks like a call to method named {\txt new}, that takes a
-single argument for each type parameter and each instance field of the
-syntactically enclosing class. The result of a {\txt new} expression
+on it.
+
+For example, in MixGen the following program is legal:
+
+\begin{verbatim}
+class C<T extends S with T()> {
+  T foo() { return new T(); }
+}
+\end{verbatim}
+
+However, to make this program compile, a \txt{with} clause is required
+on the type parameter \txt{T}.  This is because the type associated
+with a class does not describe its constructors.
+
+In MCJ, this program can be naturally rewritten with a factory method.
+
+\begin{verbatim}
+class C<T extends S> {
+  T foo() { return T.make(); }
+}
+\end{verbatim}
+
+Here, no with clause is required.
+
+A class can define class methods with arbitrary signatures that return
+new instances of the class, but each such method must ultimately
+include a {\txt{new}} expression, which can occur only within the
+scope of the class whose instance it returns. A {\txt{new}} expression
+looks like a call to method named {\txt{new}}, that takes a single
+argument for each type parameter and each instance field of the
+syntactically enclosing class. The result of a {\txt{new}} expression
 is a new instance of the enclosing class whose type parameters are
 instantiated with, and whose fields are initialized with, the given
-arguments. In our formal semantics, {\txt new} expressions are
+arguments. In our formal semantics, {\txt{new}} expressions are
 annotated with the name of the enclosing class. These annotations need
-not be added by a programmer; they could be added easily by a
+not be added by a programmer; they can be added easily by a
 straightforward syntactic preprocessing over the program, since they
-can be determined merely by the lexical scopes of the {\txt new}
+can be determined merely by the lexical scopes of the {\txt{new}}
 expressions.
@@
 
-% In a {\txt new} expression, all instance fields must be initialized at
+% In a {\txt{new}} expression, all instance fields must be initialized at
 % creation time. This requirement eliminates the problem of
@@
 of the class considered as a value.  We represent this type with the
-{\txt typeOf} operator, which takes a class instantiation and produces
+{\txt{typeOf}} operator, which takes a class instantiation and produces
 a type that is not also a class.  Because this type is not a class, it
 cannot be the superclass or kind of another class, and it cannot serve
-as the instantiation of a generic type parameter.
+as the instantiation of a generic type parameter.  We refer to this as
+{\txt{typeOf}} because it is the type of the expression that is its
+argument.
 
@@
 \begin{verbatim}
-class A() kind Object extends Object {...}
-class B() kind A extends Object {...}
+class A {...}
+class B kind A {...}
 \end{verbatim}
@@
 \subtype {\typeof \B} \A \\
-\B~ {\txt instanceof}\ \A
+\B~ {\mathtt {instanceof}}\ \A
 \end{array}
@@
 
-\hspace{-.43cm}where we use the convention that {\txt A:B} denotes
-that {\txt A} has type {\txt B}, and {\txt A<:B} denotes that {\txt A}
+\hspace{-.43cm}where we use the convention that {\txt{A:B}} denotes
+that {\txt A} has type {\txt B}, and {\txt{A<:B}} denotes that {\txt A}
 is a subtype of {\txt B}.
 
-The name {\txt typeOf} has been used in other contexts to refer to a
+The name {\txt{typeOf}} has been used in other contexts to refer to a
 runtime operation that determines the dynamic type of a object.
@@
 relationship to the operator described here.  In MCJ, an application
-of {\txt typeOf} is a static type reference.
+of {\txt{typeOf}} is a static type reference.
 
@@
 then $\B$ is a subclass of $\D$.  However, this relationship does not
-hold for {\txt instanceof} relationships.  In general, if $\B$ is an
+hold for {\txt{instanceof}} relationships.  In general, if $\B$ is an
 instance of $\C$ and $\C$ is an instance of $\D$, no judgment
@@
 
-Another difference between subclassing and {\txt instanceof}
+Another difference between subclassing and {\txt{instanceof}}
 relationships is that cycles can occur among instance class
@@
 
-\begin{displaymath}
-\typeFront {\typeArg \C{}} {\typeArg \C {}} \Object \braces{}
-\end{displaymath}
+\begin{verbatim}
+class C kind C {}
+\end{verbatim}
 
@@
 because method resolution never proceeds through more than a single
-containing class to an instance class. Therefore, we see no reason to
-disallow it.
+containing class to an instance class. If the programmer uses a field
+of \C\ to initialize a class field of \C\, the evaulation of the
+expression \C\ may fail to terminate.  But this is no different than
+the possibility of non-termination from any other self-reference.
+Therefore, we see no reason to disallow it.
 
@@
 \begin{verbatim}
-class Dimension kind Object extends Object { ... }
-class Length kind Dimension extends Object { ... }
-class Meter kind Length extends Object { ... }
-\end{verbatim}
-
-Here {\txt Meter} is a singleton class (there is only one Meter, in
+class Dimension { ... }
+class Length kind Dimension { ... }
+class Meter kind Length { ... }
+\end{verbatim}
+
+Here {\txt{Meter}} is a singleton class (there is only one Meter, in
 the Platonic sense).  We have easily expressed the desired type
@@
 \begin{verbatim}
-class Species kind Object extends Object { ... }
-class Eagle kind Species extends Object {
+class Species { ... }
+class Eagle kind Species {
       class Eagle make(String name) { ... }
@@
 \begin{verbatim}
-class Species extends Object { ... }
+class Species { ... }
 class TheEagle extends Species { ... }
-class Eagle extends Object { ... }
+class Eagle { ... }
 Eagle harry = new Eagle("Harry")
@@
 There are at least two substantial problems with this code.  First,
-{\txt TheEagle} and {\txt Eagle} have no relationship to each other in
+{\txt{TheEagle}} and {\txt{Eagle}} have no relationship to each other in
 the type system.  One is a singleton class, representing a particular
@@
 problem but which bears special attention, is the use of generic
-types.  Imagine the following method, where T is a type parameter in
-scope:\footnote{Generic methods would be straightforward to add to
-MCJ, but we exclude them to simplify the presentation of the type
-system and to focus on the important interactions of metaclasses with
-other language features.}
-
-\begin{verbatim}
-foo(T x) { ... }
-\end{verbatim}
-
-If we want to perform the call {\txt foo(harry)}, then the bound on
-{\txt T} must be either {\txt Object}, in which case {\txt foo} must
-have no knowledge of the object, or Eagle, in which case the function
-cannot handle multiple {\txt Species}.  The ``solution'' in
-conventional languages is to use a bound of {\txt Object} and insert a
-downcast, which fails at runtime if the wrong argument is passed.  In
-MCJ {\txt T} can be bound by kind {\txt Species}, and instantiated
-with {\txt Eagle}, allowing the original call to be type-checked.
+types.  Consider the following class definition, added to the
+definitions of \txt{Species}, \txt{theEagle}, etc. above.
+
+\begin{verbatim}
+class C<T extends X> {
+   foo(T x) { return T.isEndangered() }
+}
+\end{verbatim}
+
+If we want to perform the call {\txt{foo(harry)}}, then we have two
+choices for the bound \txt{X} on {\txt T}.  It must be either \Object,
+in which case {\txt{m}} must be a method that \Object\
+understands. Otherwise, it must be \txt{Eagle}, in which case the
+function cannot handle multiple {\txt{Species}}.  The ``solution'' in
+conventional languages is to use a bound of \Object\ and insert a
+downcast, which fails at runtime if the wrong argument is passed.
+This is the problem that generics were intended to alleviate.  In
+MCJ {\txt T} can be bound by kind {\txt{Species}}, and instantiated
+with {\txt{Eagle}}, allowing the original call to be type-checked.
 This gives the programmer both safety and expressiveness.
+
+\paragraph{Metaclasses and Nested Classes}
+Some of the problems are alleviated in a language with nested, or
+inner classes.  In the language Scala\cite{Scala} the following
+program expresses some of the desired invariants:
+
+\begin{verbatim}
+class Species {
+  class Member {}
+}
+object Eagles extends Species;
+class Eagle() extends Eagles.Member;
+val harry = new Eagle;
+\end{verbatim}
+\footnote{This example, and this discussion, was suggested by an
+anonymous reviewer.}
+
+This program automates many of the invariants required in a language
+without metaclasses.  However, two seperate classes must still be
+constructed per \txt{Species}.  Further, the fundamental problem, of
+not being able to group classes by functionality in the same way as
+instances, still remains.  Inner classes are not constrained by
+invariants imposed by their container.  In MCJ, the invariants of the
+kind are enforced by the type system.
 
@@
 
-Design Patterns, as exemplified in \cite{GOF}, have had a
-substantial impact on the world of object-oriented programming and
-beyond.  Design patterns allow programmers to increase the flexibility
-and abstraction of their software designs.  However, few design
-patterns have been integrated into programming languages.
+Design Patterns, as exemplified in \cite{GOF}, have had a substantial
+impact on the world of object-oriented programming and beyond.  Design
+patterns allow programmers to increase the flexibility and abstraction
+of their software designs.  However, few design patterns have been
+integrated into programming languages.
 
@@
 Factory pattern is a simple application of inheritance in the
-metaclass hierarchy.  The Prototype pattern is, as mentioned in the
-Design Patterns book, supported in languages with metaclasses.
-Finally, there is no need for the Singleton pattern; a class itself
-can serve as a singleton.  While classes can be used as limited
-singletons in the Java Programming Language and in \Cpp, doing so is
-undesirable because the static behavior of a class cannot inherit from
-another class.  In fact, the inflexibility of class or static
-operations is one of the rationales cited for the Singleton pattern.
+metaclass hierarchy.  In this, several classes could share a kind,
+giving them all a single interface for construction. The Prototype
+pattern is, as mentioned in the Design Patterns book, supported in
+languages with metaclasses.  A class is in many ways a clone of its
+metaclass.  Finally, there is no need for the Singleton pattern; a
+class itself can serve as a singleton. If no factory methods are
+provided, no instances can be created at runtime.  While classes can
+be used as limited singletons in the Java Programming Language and in
+\Cpp, doing so is undesirable because the static behavior of a class
+cannot inherit from another class.  In fact, the inflexibility of
+class or static operations is one of the rationales cited for the
+Singleton pattern.
 
@@
 A number of symbols are used to abbreviate keywords: $\superSym$
-stands for {\txt extends} and $\kindSym$ stands for {\txt kind}. Also
+stands for \extends\ and $\kindSym$ stands for \kind. Also
 $@$ represents concatenation of sequences of syntactic constructs.
@@
 directly or indirectly (however, as discussed below, cycles in the
-kind hierarchy are allowed).  Third, we assume that {\txt this} is
+kind hierarchy are allowed).  Third, we assume that \this is
 never used as the name of a variable, method or field.  We also take
@@
 
-This is a class named $\C$, with no type arguments, with kind
-$\Object$ and superclass $\Object$, which has no fields or methods
-whatsoever.  In examples that follow, we omit empty {\txt <>}.  Note
-that both the kind and the superclass are $\Object$.  MCJ does not
-have a distinguished class $\Class$.  Interestingly, we determined
-that a class $\Class$ would need no consideration from the type
-system.  Because both $\Class$ and $\Object$ would sit atop the
-hierarchy with no contents, there is no need to include
-both of them.  In a more substantial language, where $\Object$ might
-have methods or fields, there might be a use for a distinguished
-$\Class$ class, but it would not need any special treatment by
-the type system.
-
-A more complicated MCJ class is:\footnote{Here we use {\txt extends}
-for $\superSym$ and {\txt kind} for $\kindSym$.}
-
-\begin{verbatim}
-class Pair<A extends (Object, Object),
-           B extends (Object, Object)>
-   kind Object extends Object
+This is a class named $\C$, with no type arguments, with kind \Object\
+and superclass \Object\, which has no fields or methods whatsoever.
+In examples that follow, we omit empty type parameter lists.  Note
+that both the kind and the superclass are $\Object$.  Unlike Smalltalk
+or ObjVLisp, MCJ does not have a distinguished class
+$\Class$.\footnote{This is not the same as the \Class\ class used in
+reflection.}  Interestingly, we determined that a class $\Class$ would
+need no consideration from the type system.  Because both $\Class$ and
+$\Object$ would sit atop the hierarchy with no contents, there is no
+need to include both of them.  In a more substantial language, where
+$\Object$ might have methods or fields, there might be a use for a
+distinguished $\Class$ class, but it would not need any special
+treatment by the type system.
+
+A more complicated MCJ class is:
+
+\begin{verbatim}
+class Pair<A extends Object kind Object,
+           B extends Object kind Object>
 {
- class Pair<A,B> make(A a, B b){ new(A,B) (a,b) } ;
+ class Pair<A,B> make(A a, B b){ 
+    return new <A,B> (a,b) 
+} ;
  A fst;
@@
 demonstrates a number of important MCJ features.  First, we see two
-different kinds of methods.  The {\txt make} method creates a new
-instance of {\txt Pair}.  This method contains a {\txt new}
+different kinds of methods.  The {\txt{make}} method creates a new
+instance of {\txt{Pair}}.  This method contains a {\txt{new}}
 expression, which initializes the instance fields of the class
-positionally, so that the first argument to {\txt make} becomes {\txt
- fst}, and the second becomes {\txt snd}.  Also, note that we must
-mention the type parameters of the newly created instance.
-
-Second, we have an instance method, {\txt setfst}, which creates a new
+positionally, so that the first argument to {\txt{make}} becomes
+{\txt{fst}}, and the second becomes {\txt{snd}}.  Also, note that we
+must mention the type parameters of the newly created instance.
+
+Second, we have an instance method, {\txt{setfst}}, which creates a new
 pair with the old second element, and a new first element which has
@@
 
-Finally, adding the following to the above definition of {\txt Pair}
+Finally, adding the following to the above definition of {\txt{Pair}}
 gives a full MCJ program:
@@
 \begin{verbatim}
-class O kind Object extends Object
+class O
 {
@@
 This program, which creates two instances of {\txt O}, then creates
-a {\txt Pair} to hold them both, and finally selects one of the two to
+a {\txt{Pair}} to hold them both, and finally selects one of the two to
 become the value of the program, demonstrates three of the expression
-forms in MCJ.  {\txt Pair<O,O>} is a type, which, by itself, can
-be a value.  {\txt O.make()} is a method invocation (here with a
+forms in MCJ.  {\txt{Pair<O,O>}} is a type, which, by itself, can
+be a value.  {\txt{O.make()}} is a method invocation (here with a
 class, not an instance, as the receiver).  The entire expression is a
-field access, of the {\txt fst} instance field.
-
-Other kinds of expressions are seen in the body of the {\txt Pair}
-class.  {\txt new<A,B> (a,b)} is a new expression, which creates an
-instance of the enclosing class (here {\txt Pair}).  In the body of
-{\txt make} is a reference to the variable {\txt b}, which has the
-obvious value.  The final form of expression writable by the
-programmer is the cast, with the usual semantics. For example, {\txt
-(Object)O.make()}\ is an expression of type \Object that evaluates to
-an instance of {\txt O}.
+field access, of the {\txt{fst}} instance field.
+
+Other kinds of expressions are seen in the body of the {\txt{Pair}}
+class.  {\txt{new<A,B> (a,b)}} is a new expression, which creates an
+instance of the enclosing class (here {\txt{Pair}}).  In the body of
+{\txt{make}} is a variable reference to {\txt b}.  The final form of
+expression writable by the programmer is the cast, with the usual
+semantics. For example, {\txt{(Object)O.make()}}\ is an expression of
+type \Object\ that evaluates to an instance of {\txt O}.
 
 With an understanding of expressions, we can examine the rest of the
-{\txt Pair} class.  There are two methods: a class method ({\txt
-make}) and an instance method ({\txt setfst}).  There are also two
-fields, both of which are instance fields and initialized the {\txt
-new} expressions. In MCJ, a {\txt new} expression is different from
-those in \FGJ; each {\txt new} expression evaluates to an instance of
-the syntactically enclosing class.\footnote{In the grammar of figure
-\ref{fig:syntax}, {\txt new} expressions are annotated with this
-enclosing class.  In the examples, this redundant information is
-elided.}  Had there been additional fields in the superclass of {\txt
-Pair}, it would have been necessary to initialize them in the {\txt
-new} expression as well. {\txt new} expressions are explained in
-detail in section \ref{privconst}.
+{\txt{Pair}} class.  There are two methods: a class method
+({\txt{make}}) and an instance method ({\txt{setfst}}).  There are
+also two fields, both of which are instance fields and initialized the
+{\txt{new}} expressions. In MCJ, a {\txt{new}} expression is different
+from those in \FGJ; each {\txt{new}} expression evaluates to an
+instance of the syntactically enclosing class.\footnote{In the grammar
+of figure \ref{fig:syntax}, {\txt{new}} expressions are annotated with
+this enclosing class.  In the examples, this redundant information is
+elided.}  Had there been additional fields in the superclass of
+{\txt{Pair}}, it would have been necessary to initialize them in the
+{\txt{new}} expression as well. {\txt{new}} expressions are explained
+in detail in section \ref{privconst}.
 
 %\footnote{There is a strong
-%resemblance between the {\txt new} expressions here and the
+%resemblance between the {\txt{new}} expressions here and the
 %constructors of \FJ.  However, in MCJ they are a part of the
@@
 of class fields mapped to a sequence of expressions and is annotated
-with the type {\txt typeOf(C)}. Note that the right hand sides of such
+with the type \typeof{\C}. Note that the right hand sides of such
 maps need not always be values, and thus computation can take place
@@
 Therefore, unlike \FJ, our reductions do not operate entirely in the
-expressible syntax of the language.  We use the metavariable $\Phi$ to
+user-level syntax of the language.  We use the metavariable $\Phi$ to
 range over mappings.  When we need to refer to the type annotation of
-a mapping explicitly, the metavariable is written $\Phi_{\GG}$.
-
-The
-semantics are given in figures \ref{fig:subtype}, \ref{fig:exprty}
-,\ref{fig:wf}, and \ref{fig:comp} with auxiliary functions given in
-figure \ref{fig:aux}.
+a mapping explicitly, the metavariable is written $\Phi_{\GG}$.  A
+mapping whose right-hand sides are all values is itself value, and
+will be written $\Phi_{\GG}^v$.
+
+The semantics are given in figures \ref{fig:subtype},
+\ref{fig:exprty}, \ref{fig:wf}, and \ref{fig:comp} with auxiliary
+functions given in figure \ref{fig:aux}.
 
@@
 
-This function maps type variables and {\txt typeOf} applied to type
+This function maps type variables and {\txt{typeOf}} applied to type
 variables to their bounds, and leaves others unchanged.
@@
 expression $\e$ has type $\T$.  Type judgments are not transitive in
-the way that {\txt instanceof} relationships are with respect to
+the way that {\txt{instanceof}} relationships are with respect to
 subtyping: if $\dgprove {\typ \e \T}$, and $\dproves {\subtype \T
@@
 must contain the instance fields of the kind.  Finally, all
-{\txt new} expressions must have the correct class name annotation.
+{\txt{new}} expressions must have the correct class name annotation.
 
@@
 The evaluation relation given here is non-deterministic. For example,
-no order is prescribed for evaluating the arguments to a {\txt new}
+no order is prescribed for evaluating the arguments to a {\txt{new}}
 expression or method call, or for evaluating the initializers for
@@
 important points about the integration of metaclasses into an
-object-oriented system.  The {\txt typeOf} type operator is to our
+object-oriented system.  The {\txt{typeOf}} type operator is to our
 knowledge unique, and plays a key role in the soundness of the
-system.  The discovery that the {\txt Class} class played no special
+system.  The discovery that the {\txt{Class}} class played no special
 role, and that thus the class and metaclass hierarchies could both be
-rooted at {\txt Object} is also novel.  Finally, we have shown how an
+rooted at {\txt{Object}} is also novel.  Finally, we have shown how an
 expressive framework of metaclasses has positive effects in other
@@
 quantities. Such an extension would allow for a proof of  ``dimensional
-soundness'' in the resulting system.  Further, any realistic system
-will have imperative features, and while such features do not seem
-likely to interact badly with metaclasses, no sound system is built on
-such assumptions.
+soundness'' in the resulting system.  Further, the calculus presented
+here is sufficent for demonstrating the techincal detail, but not for
+practical programming.  Many extensions would be neccessary for MCJ to
+become a usable language in practice.
+
+One of the most important extensions is a mechanism for side effects.
+Imperative features present a some difficulties for this calculus as
+formulated.  The most important problem is that the initializers for
+class fields are re-evaluated every time the class is used as an
+expression.  If those expressions had side effects, this prospect
+would make class fields unusable.  One potential solution is to
+remember the previously evaluated results, but much more work remains
+in this area.
 
@@
 
+Finally, the language has not yet been implemented.  An implementation
+might suggest changes to the language, or make the differences with
+other languages clearer.  Further, an implementation on the Java
+Virtual Machine would be a useful are of exploration.
+
 \section*{Acknowledgments}

### 2004-12-12 classes/algorithms/final.tex
work on final exam

\documentclass[twoside]{amsart}
\usepackage{geometry}
\usepackage{stmaryrd}
\usepackage{cancel}
\usepackage{clrscode}
\usepackage{textcomp}
\geometry{tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
\usepackage[curve]{xypic}

\newsavebox{\fmbox}
\newenvironment{fmpage}[1]
{\begin{lrbox}{\fmbox}
   \begin{minipage}{#1}}
{\end{minipage}
\end{lrbox}\fbox{\usebox{\fmbox}}}

\newenvironment{directive}{\begin{fmpage}{5.5in}}{\end{fmpage}}
\newenvironment{problem}[1]{\newpage\section{#1}\begin{fmpage}{5.5in}}{\end{fmpage}}

\newtheorem{prop}{Proposition}
\newtheorem{claim}[prop]{Claim}
\newtheorem{lemma}[prop]{Lemma}

\begin{document}

{\raggedleft
Sam Tobin-Hochstadt \\
CSG 713 Final Exam \\
\today \\
{\tt samth@ccs.neu.edu} \\
}

%\newcommand{\min}[#1]{\textrm{min}(

\section*{Problem 1}

First, initialize min($v_1$) to $v_1$.  Then, for each successive
vertex $v_k$, perform BFS from that vertex.  If a previously finished
vertex $v_j$ is reached, then terminate the BFS and min($v_k$) =
min($v_j$). This algorithm visits every vertex and every edge at most
once, and thus takes $O(V + E)$ time.

\section*{Problem 2}

\subsection*{Part i}
This conjecture is false.

\xymatrix{
  v_1 \ar@{-}[d] \ar@{-}[r] \ar@{-}[dr] & v_2 \ar@{-}[d] \\
  v_3 \ar@{-}[r] & v_4
}

\subsection*{Part ii}
This conjecture is false.

\xymatrix{
  v_1 \ar@{-}[rr] \ar@{-}[ddr] \ar@{-}[dr]  & \txt{} & v_2 \ar@{-}[ddl]
  \ar@{-}[dl]    \\
  \txt{} & v_3 \ar@{-}[d]   & \txt{} \\
  \txt{} & v_4    & \txt{} 
}

\section*{Problem 3}

\subsection*{Part i}

This conjecture is true.

\subsection*{Part ii}

This conjecture is false.

\xymatrix{
  v1 \ar@{-}[r] \ar@{-}[d] & v2 \\
  v3 \ar@{-}[r] & v4 \ar@{-}[u]
}

\section*{Problem 4}

The Laksurk method successfully finds a minimum spanning tree for any
graph.

\section*{Problem 5}

\section*{Problem 6}

\subsection*{Part i}

\subsection*{Part ii}

Create a vertex for every number in the sequence.  Then create an edge
from every vertex to

\section*{Problem 7}

\section*{Problem 8}

To see that an approximation ratio of 2 is tight, consider the
following graph:

\xymatrix{
  v_1 \ar@{-}[r] & v_3 \ar@{-}[r] & v_2
}

Then $A$ and $B$ will be initialized, and there will only be one
remaining vertex.  Since $d(v_3,\{v_1\}) = d(v_3,\{v_2\}) = 1$, $v_3$
will be allocated to $B$.  This produces a cut of size 1.  However,
the maximal cut is $(\{v_1,v_2\},\{v_3\})$, which has size 2.

\end{document}

### 2004-12-12 ffi/cairo.h
add test files

@@ -1,0 +1,859 @@
+/*
+ * Copyright © 2002 University of Southern California
+ *
+ * Permission to use, copy, modify, distribute, and sell this software
+ * and its documentation for any purpose is hereby granted without
+ * fee, provided that the above copyright notice appear in all copies
+ * and that both that copyright notice and this permission notice
+ * appear in supporting documentation, and that the name of the
+ * University of Southern California not be used in advertising or
+ * publicity pertaining to distribution of the software without
+ * specific, written prior permission. The University of Southern
+ * California makes no representations about the suitability of this
+ * software for any purpose.  It is provided "as is" without express
+ * or implied warranty.
+ *
+ * THE UNIVERSITY OF SOUTHERN CALIFORNIA DISCLAIMS ALL WARRANTIES WITH
+ * REGARD TO THIS SOFTWARE, INCLUDING ALL IMPLIED WARRANTIES OF
+ * MERCHANTABILITY AND FITNESS, IN NO EVENT SHALL THE UNIVERSITY OF
+ * SOUTHERN CALIFORNIA BE LIABLE FOR ANY SPECIAL, INDIRECT OR
+ * CONSEQUENTIAL DAMAGES OR ANY DAMAGES WHATSOEVER RESULTING FROM LOSS
+ * OF USE, DATA OR PROFITS, WHETHER IN AN ACTION OF CONTRACT,
+ * NEGLIGENCE OR OTHER TORTIOUS ACTION, ARISING OUT OF OR IN
+ * CONNECTION WITH THE USE OR PERFORMANCE OF THIS SOFTWARE.
+ *
+ * Author: Carl D. Worth <cworth@isi.edu>
+ */
+
+#ifndef _CAIRO_H_
+#define _CAIRO_H_
+
+#include <cairo-features.h>
+
+#include <pixman.h>
+#include <stdio.h>
+
+typedef struct cairo cairo_t;
+typedef struct cairo_surface cairo_surface_t;
+typedef struct cairo_matrix cairo_matrix_t;
+typedef struct cairo_pattern cairo_pattern_t;
+
+#ifdef __cplusplus
+extern "C" {
+#endif
+
+/* Functions for manipulating state objects */
+cairo_t *
+cairo_create (void);
+
+void
+cairo_reference (cairo_t *cr);
+
+void
[…]

### 2004-12-13 classes/algorithms/final.tex
finished 1.5 problems

@@
 \subsection*{Part i}
-This conjecture is false.
+This conjecture is false.  No execution of Breadth-First search can
+give the minimal spanning tree of this graph.
 
 \xymatrix{
-  v_1 \ar@{-}[d] \ar@{-}[r] \ar@{-}[dr] & v_2 \ar@{-}[d] \\
-  v_3 \ar@{-}[r] & v_4
+  v_1 \ar@{-}[d]_{100} \ar@{-}[r]^1 \ar@{-}[dr]^1 & v_2 \ar@{-}[d]^{100} \\
+  v_3 \ar@{-}[r]^1 & v_4
 }
 
+Any execution of BFS must begin at one of the four vertices.  It must
+then visit all of the edges incident on that vertex.  For any vertex,
+that includes one of the edges weighted 100.  Thus, that execution
+cannot produce the minimal spanning tree. \qed
+
 \subsection*{Part ii}
-This conjecture is false.
+This conjecture is false. No execution of Depth-First search can
+produce the minimal spanning tree of this graph.
 
 \xymatrix{
-  v_1 \ar@{-}[rr] \ar@{-}[ddr] \ar@{-}[dr]  & \txt{} & v_2 \ar@{-}[ddl]
-  \ar@{-}[dl]    \\
-  \txt{} & v_3 \ar@{-}[d]   & \txt{} \\
+  v_1 \ar@{-}[rr]^{100} \ar@{-}[ddr]_{100} \ar@{-}[dr]^1  & \txt{} & v_2 \ar@{-}[ddl]^{100}
+  \ar@{-}[dl]_1    \\
+  \txt{} & v_3 \ar@{-}[d]^1   & \txt{} \\
   \txt{} & v_4    & \txt{} 
 }
+
+There are two cases.  One, the DFS starts at the center ($v_3$).  Then it
+moves to one of the other vertices (WLOG $v_1$), and it encounters an
+edge weighted 100 to $v_2$, which has not yet been visited.
+Therefore, that edge will be included in the final spanning tree, and
+the tree will not be minimal.
+
+In the other case, the DFS starts on one of the edge vertices (WLOG
+$v_1$).  Either it begins with one of the outer edges, in which case
+the tree cannot be minimal.  Otherwise, it moves to the middle, and
+then to one of the other outer vertices (WLOG $v_2$).  Then there is
+an edge to $v_4$, which is not yet explored, so that edge is taken.
+Therefore, the tree will not be minimal. \qed
 
@@
 
-This conjecture is false.
+This conjecture is false. This graph
 
 \xymatrix{
-  v1 \ar@{-}[r] \ar@{-}[d] & v2 \\
-  v3 \ar@{-}[r] & v4 \ar@{-}[u]
+  v1 \ar@{-}[r]^2 \ar@{-}[d]_2 & v2 \\
+  v3 \ar@{-}[r]_1 & v4 \ar@{-}[u]_1
 }
+
+has the following MMST:
+
+\xymatrix{
+  v1 \ar@{-}[r]^2 \ar@{-}[d]_2 & v2 \\
+  v3  & v4 \ar@{-}[u]_1
+}
+
+but this is an MST with less total weight.
+
+\xymatrix{
+  v1 \ar@{-}[r]^2  & v2 \\
+  v3 \ar@{-}[r]_1 & v4 \ar@{-}[u]_1
+}
+
+\qed
 

### 2004-12-15 classes/algorithms/final.tex
lots and lots of work on final

@@
 vertex $v_k$, perform BFS from that vertex.  If a previously finished
-vertex $v_j$ is reached, then terminate the BFS and min($v_k$) =
-min($v_j$). This algorithm visits every vertex and every edge at most
-once, and thus takes $O(V + E)$ time.
+vertex $v_j$ is reached, then do not follow any vertices from that
+point, and tenatively set min($v_k$) = min($v_j$). Continue the BFS
+until either the graph is exhausted, or all unvisited vertices have
+been previously finished.  Each time a new finished vertex $v_l$ is
+encountered, make min($v_l$) the new minimum for $v_k$ if it is lower
+than the current minimum.  Therefore, min($v_k$) will be the smallest
+index previously visited that is reachable from $v_l$. After
+performing this for all vertices, each vertex will have the correct
+minimum associated with it.
+
+This algorithm visits every vertex at least once, obviously.  It also
+never visits a previously-visited edge, since it does not pursue BFS
+past the point where it finds a previously-visited vertex, and a
+previously visited edge must connected two previously-unvisited
+vertexes.  Therefore, it cannot visit any more than $E$ edges.
+Therefore, the total time must be $O(V+E)$. \qed
 
@@
 
-\xymatrix{
+\[\xymatrix{
   v_1 \ar@{-}[d]_{100} \ar@{-}[r]^1 \ar@{-}[dr]^1 & v_2 \ar@{-}[d]^{100} \\
   v_3 \ar@{-}[r]^1 & v_4
-}
+}\]
 
@@
 
-\xymatrix{
+\[\xymatrix{
   v_1 \ar@{-}[rr]^{100} \ar@{-}[ddr]_{100} \ar@{-}[dr]^1  & \txt{} & v_2 \ar@{-}[ddl]^{100}
@@
   \txt{} & v_4    & \txt{} 
-}
+}\]
 
@@
 $v_1$).  Either it begins with one of the outer edges, in which case
-the tree cannot be minimal.  Otherwise, it moves to the middle, and
+the tree cannot be minimal.  Otherwise, it moves to $v_3$, and
 then to one of the other outer vertices (WLOG $v_2$).  Then there is
@@
 
-This conjecture is true.
+All MSTs are MMSTs.  Consider the heaviest edge in the MST.  Then
+removing that edge divides the tree into two trees.  Now, the edge
+must be the lightest edge connecting those two trees.  Consider the
+other edges in the graph that connect those trees.  If they exist,
+they must be heavier than the edge in the MST.  Therefore, the MST
+cannot have a lighter heaviest edge.  Otherwise, there is no other
+edge connecting those two trees.  Therefore, that edge must be in any
+spanning tree, in particular the MMST.  \qed
 
@@
 
-This conjecture is false. This graph
+Not all MMSTs are MSTs.  This graph:
 
-\xymatrix{
+\[\xymatrix{
   v1 \ar@{-}[r]^2 \ar@{-}[d]_2 & v2 \\
   v3 \ar@{-}[r]_1 & v4 \ar@{-}[u]_1
-}
+}\]
 
@@
 
-\xymatrix{
+\[\xymatrix{
   v1 \ar@{-}[r]^2 \ar@{-}[d]_2 & v2 \\
   v3  & v4 \ar@{-}[u]_1
-}
+}\]
 
@@
 
-\xymatrix{
+\[\xymatrix{
   v1 \ar@{-}[r]^2  & v2 \\
   v3 \ar@{-}[r]_1 & v4 \ar@{-}[u]_1
-}
+}\]
 
@@
 The Laksurk method successfully finds a minimum spanning tree for any
-graph.
+graph.  To show this, assume that there is some graph for which this
+is not true.  Then there must be some edge in the graph produced which
+is not in the MST.  Call this edge $e$.  Removing $e$ separates the
+tree into two parts.  Therefore, the MST must have an edge between
+these two trees.  But if there was a lighter edge between those two
+trees, it would have replaced $e$ in the course of the Laksurk
+method.  Therefore, $e$ cannot exist.  Thus the Laksurk method
+produces the minimal spanning tree. \qed
 
 \section*{Problem 5}
+
+First, compute all pairs shortest paths for the graph.  By repeated
+application of Dijkstra's algorithm, this can be accomplished in
+$O(V^3)$ time.  Then, for each vertex $v_i$, iterate through all
+vertices $v_j$ and compute the sum of the shortest path $v_i \leadsto v_j +
+v_j \leadsto v_i$.  Find the minimum such sum, and store it for that
+vertex.  Then find the mimumum of these sums over all vertices in the
+graph.  This is the minimum weight cycle for the entire graph.
+
+This process involves $O(V^3)$ preprocessing, and then $O(V)$ work for
+each of the $V$ nodes in the graph, so it runs in $O(V^3 + V^2) =
+O(V^3)$ time.  \qed
 
@@
 
+Create a graph with a start vertex, an end vertex, and vertex
+$v_{x_i}$ for every number $x_i$.  Add an edge from the start vertex
+to every vertex except the end vertex.  Add a 0-weight edge from every
+vertex except the start vertex to the end vertex.  Then weight the
+edge from the start vertex to each $v_{x_i}$ with weight $-x_i$.  Then
+add an edge $(v_{x_i}, v_{x_{i+1}}$ with weight $-x_{i+1}$ for vertices
+$v_{x_1}$ to $v_{x_{n-1}}$.
+
+This creates a graph with $n+2$ vertices and $n + n + n-1 = 3n-1$
+edges.
+
+Then we simply use DAG-SP to compute the shortest path from the start
+to then end vertex.  This takes time $O(V+E) = O(n)$.  This shortest
+path will have weight $\sigma_{k=i}^j -x_k$ for all the $v_{x_k}$ that
+it passes through.  But minimizing this sum corresponds exactly to
+maximizing  $\sigma_{k=i}^j x_k$.  Therefore, the shortest path is the
+sequence of numbers that maximizes $S(i,j)$.
+
 \subsection*{Part ii}
 
-Create a vertex for every number in the sequence.  Then create an edge
-from every vertex to
+Create a vertex for every number in the sequence.  Create a single end
+vertex. Then create an edge from every vertex to the end vertex.  Then
+create a start vertex with and edge to every other vertex.
+
+Then create edges from a number to every number in the sequence that
+is later in the sequence and larger.  This requires iterating through
+the sequence once for each element, which requires $O(n^2)$ time.
+
+Then we merely need to find the longest path from the start vertex to
+the end vertex.  While this problem is NP-complete in general, our
+graph is very simple.
+
+First, compute the inverse of the graph.  Then begin at the end
+vertex.  Perform BFS.  When each vertex is reached, record the number
+of links followed thus far to get to that vertex.  If the number is
+greater than the value currently stored at that vertex, then update
+the value to the larger one, and continue with BFS, even though the
+vertex was already discovered.  Also record the vertex from which this
+vertex was reached.  When we complete, we will have a record of the
+longest path in the graph.  To compute the answer, begin at the start
+vertex.  Then add the number represented by the vertex stored at the
+start vertex.  Then add the vertex stored at that vertex, and so on.
+Eventually the end vertex will be reached, and we will have output the
+entire longest list.
+
+This requires processing of every edge and every vertex, and thus take
+$O(V+E) = O(n+n^2) = O(n^2)$ time.
+
+\qed
 
@@
 
+\subsection*{Part i}
+
+Given the max flow in $G$, determine the residual network.  This is
+easily computed by subtracting the flow on each edge from the capacity
+of that edge.  This takes time $O(E)$.
+
+This produces a disconnected graph, by the Max-Flow/Min-Cut theorem.
+As stated in the problem, we can assume that every edge in this new
+residual network has integral capacites, which means that all the
+remaining edges have capacity 1.
+
+Then add the new edge $e'$ to this residual network.  If there is a
+path from the source to the sink with this addition, then there must
+an augmenting path of capacity 1 in the residual network, and the new
+flow is $|f| + 1$. We can then update the flow with the new augmenting
+path in $O(E)$ time.  Otherwise, there is no new augmenting path, and
+the new flow is the same as the old flow. Since we can compute
+reachability in $O(V+E)$ time, the entire algorithm runs in $O(V+E)$
+time. \qed
+
+\subsection*{Part ii}
+
+Given the max flow in $G$, determine the residual network.  This is
+easily computed by subtracting the flow on each edge from the capacity
+of that edge.  This takes time $O(E)$.
+
+This produces a disconnected graph, by the Max-Flow/Min-Cut theorem.
+As stated in the problem, we can assume that every edge in this new
+residual network has integral capacites, which means that all the
+remaining edges have capacity 1.
+
+Then, determine if the edge $e$ is in the residual network.  If it is,
+then it must not be used in the maximal flow, and removing it does not
+change the flow.  If it is not, then it must be used in the maximal
+flow.
+
 \section*{Problem 8}
+
+The optimal cut has size at most $|E|$.  There are two kinds of
+edges in the result produced by the algorithm: edges within $A$ or
+within $B$ and edges between $A$ and $B$.  Call the size of the former
+set $x$, and the size of the latter set $y$.  On each iteration of the
+algorithm, we have a new $x'$ and $y'$.  But because of the test in
+the algorithm, $x' - x <= y' - y$.  Therefore, since $x$ starts of
+smaller than $y$ (since there are no self-loops), $y$ must always be
+greater than $x$.  Therefore, at least half of the edges must be
+between $A$ and $B$.  Therefore, the cut must be at least half as big
+as optimal.  \qed
 
@@
 
-\xymatrix{
+\[\xymatrix{
   v_1 \ar@{-}[r] & v_3 \ar@{-}[r] & v_2
-}
+}\]
 

### 2004-12-16 classes/algorithms/final.tex
final algo exam

@@
 maximizing  $\sigma_{k=i}^j x_k$.  Therefore, the shortest path is the
-sequence of numbers that maximizes $S(i,j)$.
+sequence of numbers that maximizes $S(i,j)$. \qed
 
@@
 
+If edge $e$ does participate in the maximal flow, then consider the
+flow network for graph $G$ given by making the capacity of each edge
+the quanity of flow on that edge in $f$. Compute some path from the
+source to the sink using edge $e$.  This can be done in time $O(V+E)$
+by simply searching forwards from the ending vertex of $e$, and
+backwards from the starting vertex of $e$.  Then add the flow on this
+path to the residual network left after computing the original flow
+$f$.  Then, remove edge $e$ from this augmented flow network.  If
+there is still a path from the source to the sink, then it is possible
+to route around the deletion, and the new flow is the original flow
+with the first augmenting path removed, and the new augmenting path
+added.  Otherwise, the new flow is simply the old flow with the
+augmenting path removed, and the new flow has size $|f|-1$.
+\qed
+
 \section*{Problem 8}

### 2004-12-16 metaclasses/paper.tex
submitted version of paper

@@
-\documentclass{acm-sigplan}
+\documentclass[nocopyrightspace,10pt]{acm-sigplan}
 
@@
 Metaclasses provide a useful mechanism for abstraction in
-object-oriented languages.  But most languages supporting metaclasses
-have imposed severe restrictions on their use. Typically, a metaclass
+object-oriented languages.  But most languages that support metaclasses
+ impose severe restrictions on their use. Typically, a metaclass
 is allowed to have only a single instance and all metaclasses are
 required to share a common superclass \cite{objvlisp}. In addition,
-few languages supporting metaclasses have included a static type
-system, and none have included a type system with nominal subtyping
-(i.e., subtyping as defined in mainstream languages such as the
+few languages that support metaclasses include a static type
+system, and none include a type system with nominal subtyping
+(i.e., subtyping as defined in languages such as the
 Java${^{\mbox{\tiny{TM}}}}$ Programming Language or C\#).
 
-In order to elucidate the structure of
-metaclasses and their relationship with static types, we present a
-core calculus for a nominally-typed object-oriented language with
-metaclasses and prove type soundness over this core. 
-To our knowledge, no previous formalization of metaclasses in a 
-language with nominal subtyping exists.
-This calculus is
-presented as an adaptation of Featherweight GJ \cite{FJ}, and is
-powerful enough to capture metaclass relationships beyond what are
+To elucidate the structure of metaclasses and their relationship with
+static types, we present a core calculus for a nominally typed
+object-oriented language with metaclasses and prove type soundness
+over this core.  To our knowledge, no previous formalization of
+metaclasses in a language with nominal subtyping exists.  This
+calculus is presented as an adaptation of Featherweight GJ \cite{FJ},
+and is powerful enough to capture metaclass relationships beyond those
 expressible in common object-oriented languages, including arbitrary
-metaclass hierarchies and classes as values.  We also explore the
-integration of object-oriented design patterns with metaclasses, and
-show how our calculus integrates several such patterns cleanly.
+metaclass hierarchies and classes as values.  We also describe how the
+addition of metaclasses allow for integrated and natural expression of
+several common design patterns.
 \end{abstract}
@@
 \section{Introduction}
+
+\subsection{A Problem of Modeling}
 One of the stated benefits of object-oriented languages is their
@@
 However, most such languages are unable to model many simple and
-common relationships.
-
-An excellent example is provided by Welty and Ferrucci in \cite{FW96}, where
-they show the difficulties in modeling a simple
-ontology including the notions {\txt{Species}} and {\txt{Eagle}}, while
-also capturing the relationship between {\txt{Eagle}} and {\txt{Harry}},
-a particular eagle.
-
-To see the problem, consider a system with a class \txt{Eagle} and an
-instance \txt{Harry}.  The static type system can ensure that any
-messages that \txt{Harry} responds to are also understood by any other
-instance of \txt{Eagle}.
-
-However, even in a language where classes themselves can have methods,
-such relationships cannot be expressed.  For example, in a language
-with static methods, we can write the expression
-\txt{Eagle.isEndangered()}
-
-However, if \txt{Salmon} is a species, our type system provides us no
-way of requiring that it also have an \txt{isEndangered} method.  The
-solution that Ferruci and Welty suggest, and that we develop in this
-paper, is to allow a consistent treatment of both instances such as
-\txt{Harry} and classes such as \txt{Eagle}.
+common relationships.  An excellent example is provided by Welty and
+Ferrucci \cite{FW96}, where they show the difficulties in modeling
+a simple ontology including the notions {\txt{Species}} and
+{\txt{Eagle}}, while also capturing the relationship between
+{\txt{Eagle}} and {\txt{Harry}}, a particular eagle.  To see the
+problem, consider a system with a class \txt{Eagle} and an instance
+\txt{Harry}.  The static type system can ensure that any messages that
+\txt{Harry} responds to are also understood by any other instance of
+\txt{Eagle}.
+
+However, even in a language where classes can have static methods, the
+desired relationships cannot be expressed.  For example, in such a
+language, we can write the expression \linebreak[4]\txt{Eagle.isEndangered()}
+provided that \txt{Eagle} has the appropriate static method.  Such a
+method is appropriate for any class that is a species.  However, if
+\txt{Salmon} is another class in our system, our type system provides
+us no way of requiring that it also have an \txt{isEndangered}
+method.  This is because we cannot classify both \txt{Salmon} and
+\txt{Eagle}, \emph{when treated as receivers of methods}, into a
+larger group, such as species.
 
@@
 instance.  This adds new invariants which the programmer must
-maintain, with no help from the type system.  Instead, we propose that
-the type system should allow the programmer to express the desired
-relationship directly, and enforce the invariants naturally.
-
-For example, consider the relationship between
-physical quantities, units of measurement, and physical dimensions. It
-is obvious that ``length'' is a dimension, and that ``3 feet'' is a
-length.  However, modeling this relationship in a conventional
-language such as the Java Programming Language is difficult without
-attributing to some of these concepts properties which do not properly
-belong to them.  It is natural to model {\txt{3 feet}} as an instance
-of a class {\txt{Length}}. But if we were to define a class {\txt
-Dimension} and define {\txt{Length}} to be an instance of class {\txt
-Dimension}, then {\txt{Length}} could not be a class and so {\txt 3
-feet} could not be an instance of {\txt{Length}}. Alternatively, we
-might define {\txt{Length}} to be a subclass of {\txt{Dimension}}. But
-then we still cannot define {\txt 3 feet} to be an instance of {\txt
-Length} because anything that is an instance of {\txt{Length}} would
-also be an instance of class {\txt{Dimension}} and {\txt 3 feet} is
-obviously not a dimension. The key limitation preventing us from
-modeling this domain properly is that we cannot define a single
-concept to serve both as a class and as an instance of a class.
-
-Again, the limitation is that we cannot define a class as an instance of another class.
-%Welty and Ferrucci proceed to explain how this limitation prevents conventional modeling
-%languages from capturing many important aspects of the world.  
-%The languages Ferrucci and Welty discuss cannot accommodate multiple
-%universes of discourse, in which a single thing may have different meanings in different universes.
-
-In order to overcome this limitation in expressiveness, several
-object-oriented languages that are not statically typed, such as
-Smalltalk, Python, and Self, allow for more flexible object
-relationships \cite{Smalltalk80}, \cite{ArtMOP}, \cite{PythonMeta}. In
-the case of Smalltalk and Python, classes are instances of
-metaclasses. Static members of a class are modeled as ordinary members
-of the corresponding metaclass.  But in both of these languages there
-are important limitations on expressiveness.  For example, in
-Smalltalk, the metaclass hierarchy is only two levels deep.  If we
-want to represent the relationships ``{\txt A} is an instance of {\txt
-B}", ``{\txt B} is an instance of {\txt C}", and ``{\txt C} is an
-instance of {\txt D}", we can only do this in Smalltalk if {\txt D} is
-the special Smalltalk class {\txt Metaclass}.  This does not allow the
-multi-level hierarchies required to model the examples discussed above
-\cite{Smalltalk80}.  Self, on the other hand, is a prototype-based
-object-oriented language, where there are no classes at all; object
-instantiation consists of cloning an existing instance. The members of
-the clone can be modified and added to at will, dispensing with class
-relationships entirely and making static checking difficult.
-
-A more expressive metaclass system in ObjVlisp, described informally
-by Cointe in \cite{objvlisp}.  In ObjVlisp, there is no restriction on
+maintain, with no help from the type system.
+
+These problems are not unique to the modeling of biology.  Consider
+the relationship between physical quantities, units of measurement,
+and physical dimensions. It is obvious that ``length'' is a dimension,
+and that ``3 feet'' is a length.  However, modeling this relationship
+in a conventional language such as the Java Programming Language is
+difficult without attributing to some of these concepts properties
+which do not properly belong to them.  It is natural to model {\txt{3
+    feet}} as an instance of a class {\txt{Length}}. But if we were to
+define a class {\txt Dimension} and define {\txt{Length}} to be an
+instance of class {\txt Dimension}, then {\txt{Length}} could not be a
+class and so {\txt 3 feet} could not be an instance of {\txt{Length}}.
+Alternatively, we might define {\txt{Length}} to be a subclass of
+{\txt{Dimension}}. But then we still cannot define {\txt 3 feet} to be
+an instance of {\txt Length} because anything that is an instance of
+{\txt{Length}} would also be an instance of class {\txt{Dimension}}
+and {\txt 3 feet} is obviously not a dimension.
+
+The fundamental problem that both these examples illustrate is that we
+are unable to describe multi-level \txt{instanceof} hierarchies.
+Everything must be either a class or an instance.  Classes cannot be
+instances of classes, and instances cannot have instances.  The
+solution to this problem is to remove these divisions, and allow
+the relationships we want to have a natural expression in our
+language.
+
+\subsection{Some Previous Solutions}
+In order to overcome this limitation in expressiveness described
+above, several object-oriented languages that are not statically
+typed, such as Smalltalk, Python, and Self, allow for more flexible
+object relationships \cite{Smalltalk80}, \cite{ArtMOP},
+\cite{PythonMeta}. In the case of Smalltalk and Python, classes are
+instances of metaclasses. Static members of a class are modeled as
+ordinary members of the corresponding metaclass.  But in both of these
+languages there are important limitations on expressiveness.  For
+example, in Smalltalk, the metaclass hierarchy is only two levels
+deep.  If we want to represent the relationships ``{\txt A} is an
+instance of {\txt B}'', ``{\txt B} is an instance of {\txt C}'', and
+``{\txt C} is an instance of {\txt D}'', we can only do this in
+Smalltalk if {\txt D} is the special Smalltalk class {\txt Metaclass}.
+This does not allow the multi-level hierarchies required to model the
+examples discussed above \cite{Smalltalk80}.  Self, on the other hand,
+is a prototype-based object-oriented language, where there are no
+classes at all; object instantiation consists of cloning an existing
+instance.  The members of the clone can be modified and added to at
+will, dispensing with class relationships entirely and making static
+checking difficult.
+
+A more expressive metaclass system is ObjVLisp, described informally
+by Cointe \cite{objvlisp}.  In ObjVLisp, there is no restriction on
 the number of instances of a metaclass, and metaclasses are not
 required to share a common superclass. The instance methods of a
-metaclass are inherited as the static members of its instances.  Conte
+metaclass are inherited as the static members of its instances.  Cointe
 makes a number of arguments for the benefits of generalized
-metaclasses.  In response to complaints that the Smalltalk80 metaclass
+metaclasses.  In response to complaints that the Smalltalk-80 metaclass
 system is too complex, he suggests that a more powerful and flexible
 system has the potential to be much simpler.
-%, arguing that ``The ObjVlisp model's primary advantage is uniformity.''
-However, ObjVlisp does not include a static type system. 
-%Cointe quotes from \cite{cardelli:multiple}: ``With respect to
-%Simula, Smalltalk also abandons ... strong typing, allowing it to
-%... introduce the notion of meta-classes''.   
-In this paper, we show that this limitation is not necessary. We
-present a semantics for a statically typed calculus of metaclasses
-where, as in ObjVlisp, every class can serve as both an instance of a
-class and as a class of instances, and there is no limitation on the
-levels of nesting of metaclasses. In the process, we also reveal
-several interesting properties about the structure that a nominal type
-system for metaclasses must take. We then present a formal operational
-semantics for this calculus, as well as a proof of type soundness.
+However, ObjVLisp does not include a static type system.  As Cardelli
+points out in the context of Smalltalk \cite{cardelli:multiple}, ``With
+respect to Simula, Smalltalk also abandons ... strong typing, allowing
+it to ... introduce the notion of meta-classes''.
+
+\subsection{A More Expressive Solution}
+
+In this paper, we show that the limitations of previous metaclass
+systems are not inherent. We present a semantics for a calculus with
+metaclasses in which, as in ObjVLisp, every class can serve as both an
+instance of a class and as a class of instances, and there is no
+limitation on the levels of nesting of metaclasses. But, unlike
+ObjVLisp, we present a static type system for this semantics, and use
+it to allow more invariants to be checked statically.  In the process,
+we also reveal several interesting properties about the structure that
+a nominal type system for metaclasses must take. We then present a
+formal operational semantics for this calculus, as well as a proof of
+type soundness.
 
@@
 for metaclasses. In MetaGen, metaclasses are used to model the type
-relationships of dimensions.  As with ObjVlisp, the notion of a
+relationships of dimensions.  As with ObjVLisp, the notion of a
 metaclass in MetaGen is more general than that of languages such as
-Smalltalk, but, unlike ObjVlisp, MetaGen includes a static type system
+Smalltalk, but, unlike ObjVLisp, MetaGen includes a static type system
 \cite{DimUnits}. The calculus in this paper can be viewed as a
@@
 following discussion we restrict ourselves to simple examples, and do
-not delve into the complexities of object-oriented analysis with metaclasses.  
-This is certainly not because we believe that
+not delve into the complexities of object-oriented analysis with
+metaclasses.  This is certainly not because we believe that
 metaclasses are most useful for the development of toy object-oriented
-programs.  On the contrary, metaclasses are most likely to show their value
-in large systems, both in terms of modeling and in the treatment of
-design patterns in section \ref{designpatterns}.
+programs.  On the contrary, metaclasses are most likely to show their
+value in large systems, both in terms of modeling and in the treatment
+of design patterns in section \ref{designpatterns}.
 
@@
 
-%Another, somewhat more flexible solution to this problem can be found in
-%functional languages, where sets can be represented as predicates over
-%objects, and sets of sets can be represented as predicates over
-%predicates.  This hierarchy can obviously be extended indefinitely,
-%but such programs would quickly become unwieldy.  Specialized language
-%constructs would be needed for this approach to become a serious programming
-%style, just as most object-oriented programming languages
-%provide constructs to support that methodology.  Also in the
-%functional world, higher-order functions represent a way of
-%parameterizing at multiple levels, allowing some of the desired
-%behavior described here to be implemented.  But few functional
-%programs nest higher-orderness more than a few levels, since
-%understanding the relationships  becomes more than one programmer
-%could follow.  Instead, we need a way of structuring such
-%relationships explicitly.
-
-%Formal modeling of existing languages has become common, with numerous
-%formal models for languages such as the Java (TM) Programming Language.  However, formal modeling
-%can also serve to aid in the design process for a new language.  In
-%this paper, we describe MCJ, a nominally-typed core calculus for metaclasses,
-%which is intended to be the basis for a language design. While we have focused on a formal model as a language design guide, MCJ can also help understand existing languages with metaclasses.
-
 \section{MCJ}
@@
 MCJ is a object-oriented calculus with generic types based on
-Featherweight GJ which includes metaclasses. An MCJ program consists of
+Featherweight GJ that includes metaclasses. An MCJ program consists of
 a sequence of class definitions followed by a trailing
@@
 \end{verbatim}
-\footnote{In other examples, we will leave out the \kind\ and \extends\
-declaration when they refer to \Object.}
-
-Each class may have both class members and instance members. Class
-members define the behavior of the class when used as an
-instance. Instance members define the behavior of instances of a
-class.  In the abstract syntax presented in this paper, class members
-are distinguished from instance members by order, and with the keyword
-\class \footnote{This syntax is not ambiguous because MCJ has no
-inner classes.}.  Members of a class definition are laid out as
-follows: first, class fields are defined, followed by class methods,
-then instance fields and finally instance methods.  For example, the
-following class definition includes a class method {\txt m} and an
-instance method {\txt n}:
+
+Each class may have both static members and instance members. Static
+members define the behavior of the class when used as an instance.
+Instance members define the behavior of instances of a class.  In the
+abstract syntax presented in this paper, static members are
+distinguished from instance members by order, and with the keyword
+\static.  Members of a class definition are laid out as follows:
+first, static fields are defined, followed by static methods, then
+instance fields and finally instance methods.  For example, the
+following class definition includes a static method {\txt m} and an
+instance method {\txt n}.  Beginning with this example, we will leave
+out the \kind\ and \extends\ declaration when they refer to \Object.
 
@@
 class C {
-   class Object m() {...}
+   static Object m() {...}
    Object n() {...}
@@
 
-Class fields and methods bear a resemblance to static fields and
-methods, but there are important differences. For a given class {\txt
-C}, instance members of {\txt C}'s kind are inherited as class
-members of {\txt C}. For example, a class method {\txt m} in {\txt C}
-with the same name as an instance method of {\txt C}'s kind
+Static fields and methods in MCJ bear a strong resemblance to static
+fields and methods, but there are important differences. For a given
+class {\txt C}, instance members of {\txt C}'s kind are inherited as
+class members of {\txt C}. For example, a static method {\txt m} in
+{\txt C} with the same name as an instance method of {\txt C}'s kind
 overrides the definition of {\txt m} in the kind (and it must have the
@@
 type {\txt T}, references to the instance members of {\txt v} refer to
-the class members of {\txt C}.  As in \cite{DimUnits}, a class is
-allowed to define both a class method and an instance method of the
+the static members of {\txt C}.  As in \cite{DimUnits}, a class is
+allowed to define both a static method and an instance method of the
 same name. We prevent ambiguity in member references by requiring that
-all references to class members in MCJ must explicitly denote the
+all references to static members in MCJ must explicitly denote the
 receiver.
@@
 expression, is an instance of its kind.  If a method is invoked on a
-class {\txt C}, first the class methods of {\txt C} are examined.  If
+class {\txt C}, first the static methods of {\txt C} are examined.  If
 the method is found there, it is invoked. Otherwise, the
@@
 
-To resolve the method call $\invoke \C {\mof {}}$ first the class
+To resolve the method call $\invoke \C {\mof {}}$, first the class
 methods of $\C$ are examined, and if a match is found, it is invoked.
@@
 
-%Class {\txt C} would first be examined to see if it contained a class method named {\txt m}.
+%Class {\txt C} would first be examined to see if it contained a static method named {\txt m}.
 If so, this method would be invoked.
@@
 preventing the use of type erasure as an implementation technique. In
-addition, the receiver of a reference to a class member may be a type
+addition, the receiver of a reference to a static member may be a type
 variable. Unlike the system presented in \cite{MixGen}, MCJ does not
@@
 declares a class {\txt C} with one type parameter {\txt T} that must
-be instantiated with a type of kind {\txt D} that is a subtype of
-{\txt E}:
+be instantiated with a type that is a subtype of {\txt D} and is of
+kind {\txt E} :
 
@@
 to play such a central role in the language. Class instances exist at
-the beginning of a program execution; they need not be
-constructed. Non-class instances are naturally constructed with class
-(factory) methods \cite{GOF}; in this way a class instance can be
-passed as a type parameter and factory methods can be called on the
-class instance to construct new instances without stipulating the
-class of the constructed instance. This flexibility is similar to that
-provided with first-class genericity \cite{NextGen, MixGen}, but
-obviates the need for {\txt{with}} clauses because the bound on the
-kind of a type variable stipulates what factory methods can be called
-on it.
-
-For example, in MixGen the following program is legal:
+the beginning of a program execution; they need not be constructed.
+Non-class instances are naturally constructed with class (factory)
+methods \cite{GOF}; in this way a class instance can be passed as a
+type parameter and factory methods can be called on the class instance
+to construct new instances without stipulating the class of the
+constructed instance.  In a language with first-class genericity, type
+parameters may used in arbitrary contexts, including in \txt{new}
+expressions \cite{NextGen, MixGen}, however additional bounds, called
+\txt{with} clauses, are required on the type parameters.  MCJ provides
+similar flexibility, but obviates the need for {\txt{with}} clauses
+because the bound on the kind of a type variable stipulates what
+factory methods can be called on it.
+
+For example, in MixGen \cite{MixGen} the following program is legal:
 
@@
 
-Here, no with clause is required.
-
-A class can define class methods with arbitrary signatures that return
+Here, no \txt{with} clause is required.
+
+A class can define static methods with arbitrary signatures that return
 new instances of the class, but each such method must ultimately
@@
 All uses of constructors in \FGJ\ \cite{FJ} are macro-expressible
-\cite{FellExpress} in MCJ, that is, they can be expressed via local
+\cite{FellExpress} in MCJ; that is, they can be expressed via local
 transformations.  Of course, not all \FGJ\ programs can be expressed
-in MCJ, because of the differences in the type system. Namely, MCJ
-does not include polymorphic methods.
-
-Metaclasses allow other kinds of flexibility in object creation, as
-discussed in section \ref{designpatterns}.
+in MCJ, because of the differences in the type system. Specifically,
+MCJ does not include polymorphic methods.
+
+While these choices may seem constricting, the architecture of MCJ
+allows for considerable flexibility in object creation.  We discuss
+this point further in section \ref{designpatterns}.
 
@@
 
-However, classes which are themselves instances of other classes have
-fields as well, and these fields must also be initialized before they
-are used.  These fields include those instance fields of the kind
-which become class fields of the new class.  Our solution is to
-require that \emph{all class fields, including those obtained by
-inheritance, must be redeclared} and provided with values.
+Classes that are themselves instances of other classes have fields as
+well, and these fields must also be initialized before they are used.
+These fields include instance fields of the kind, which become static
+fields of the new class.  Our solution is to require that \emph{all
+  static fields, including those obtained by inheritance, must be
+  redeclared} and provided with values.
 
@@
 % uninitialized instance fields. However, when a class is the kind of
-% another class, the instance fields of the kind become class fields of
+% another class, the instance fields of the kind become static fields of
 % the instance class.  Since the instance class can then be used as a
@@
 % field inherited from a container class be included in a new class
-% field definition and that every class field in a class definition
+% field definition and that every static field in a class definition
 % include an initializing expression.
@@
 Given that class references can be used as expressions, it is natural
-to ask: what is the type of a class reference?  
-%In the world of the
-%$\lambda$-calculus, such types are known as kinds.  
-In MCJ, however, the kind of a class does not capture all of the
-properties of the class as a value.  For example, a class may add new
-class fields or class methods which are not present in the kind.
-Therefore, each class freely generates a new type, which is the type
-of the class considered as a value.  We represent this type with the
-{\txt{typeOf}} operator, which takes a class instantiation and produces
-a type that is not also a class.  Because this type is not a class, it
+to ask: what is the type of a class reference?  In MCJ, however, the
+kind of a class does not capture all of the properties of the class as
+a value.  For example, a class may add new static fields or static
+methods that are not present in the kind.  Therefore, each class
+freely generates a new type, which is the type of the class considered
+as a value.  We represent these types with a compile time
+{\txt{typeOf}} operator that takes a class instantiation and produces
+a new type that is not also a class.  Because this type is not a class, it
 cannot be the superclass or kind of another class, and it cannot serve
-as the instantiation of a generic type parameter.  We refer to this as
-{\txt{typeOf}} because it is the type of the expression that is its
+as the instantiation of a generic type parameter.  We refer to it as
+{\txt{typeOf}} because it returns the type of the expression that is its
 argument.
@@
 
-Another difference between subclassing and {\txt{instanceof}}
+Another difference between subclassing and instance class
 relationships is that cycles can occur among instance class
@@
 containing class to an instance class. If the programmer uses a field
-of \C\ to initialize a class field of \C\, the evaulation of the
+of \C\ to initialize a static field of \C\, the evaluation of the
 expression \C\ may fail to terminate.  But this is no different than
-the possibility of non-termination from any other self-reference.
+the possibility of nontermination from any other self-reference.
 Therefore, we see no reason to disallow it.
@@
 examples discussed earlier.  First, the relationships between
-dimensions is easy to capture. \footnote{Of course, dimensions have
+dimensions is easy to capture.\footnote{Of course, dimensions have
 many subtleties, not captured here.  All we present is the essence of
@@
 Here {\txt{Meter}} is a singleton class (there is only one Meter, in
-the Platonic sense).  We have easily expressed the desired type
+the platonic sense).  We have easily expressed the desired type
 relationships, and can statically check program invariants that rely
@@
 class Eagle kind Species {
-      class Eagle make(String name) { ... }
+      static Eagle make(String name) { ... }
 }
@@
 Another problem, which is in some ways just a symptom of the first
-problem but which bears special attention, is the use of generic
+problem but that bears special attention, is the use of generic
 types.  Consider the following class definition, added to the
-definitions of \txt{Species}, \txt{theEagle}, etc. above.
-
-\begin{verbatim}
-class C<T extends X> {
-   foo(T x) { return T.isEndangered() }
-}
-\end{verbatim}
-
-If we want to perform the call {\txt{foo(harry)}}, then we have two
+definitions of \txt{Species}, \txt{theEagle}, etc. above, where
+\txt{?} is a placeholder for the bound.
+
+\begin{verbatim}
+class C<T extends ?> {
+   T foo(T x) { return T.isEndangered() }
+}
+\end{verbatim}
+
+If we want to perform the call {\txt{c.foo(harry)}}, then we have two
 choices for the bound \txt{X} on {\txt T}.  It must be either \Object,
-in which case {\txt{m}} must be a method that \Object\
+in which case {\txt{isEndangered}} must be a method that \Object\
 understands. Otherwise, it must be \txt{Eagle}, in which case the
@@
 This is the problem that generics were intended to alleviate.  In
-MCJ {\txt T} can be bound by kind {\txt{Species}}, and instantiated
+MCJ, {\txt T} can be bound by kind {\txt{Species}}, and instantiated
 with {\txt{Eagle}}, allowing the original call to be type-checked.
@@
 
-\paragraph{Metaclasses and Nested Classes}
-Some of the problems are alleviated in a language with nested, or
-inner classes.  In the language Scala\cite{Scala} the following
-program expresses some of the desired invariants:
-
-\begin{verbatim}
-class Species {
-  class Member {}
-}
-object Eagles extends Species;
-class Eagle() extends Eagles.Member;
-val harry = new Eagle;
-\end{verbatim}
-\footnote{This example, and this discussion, was suggested by an
-anonymous reviewer.}
-
-This program automates many of the invariants required in a language
-without metaclasses.  However, two seperate classes must still be
-constructed per \txt{Species}.  Further, the fundamental problem, of
-not being able to group classes by functionality in the same way as
-instances, still remains.  Inner classes are not constrained by
-invariants imposed by their container.  In MCJ, the invariants of the
-kind are enforced by the type system.
-
 \section{Design Patterns as Language Features}
@@
 
-Design Patterns, as exemplified in \cite{GOF}, have had a substantial
+Design patterns, as exemplified in \cite{GOF}, have had a substantial
 impact on the world of object-oriented programming and beyond.  Design
@@
 classic design patterns deal with object creation, as discussed above
-in section \ref{privconst}.  It is clear that the concept of a Factory
-method is easily (and necessarily) expressed in MCJ.  However, a
-number of other patterns also find easy expression.  The Abstract
-Factory pattern is a simple application of inheritance in the
-metaclass hierarchy.  In this, several classes could share a kind,
-giving them all a single interface for construction. The Prototype
-pattern is, as mentioned in the Design Patterns book, supported in
-languages with metaclasses.  A class is in many ways a clone of its
-metaclass.  Finally, there is no need for the Singleton pattern; a
-class itself can serve as a singleton. If no factory methods are
+in section \ref{privconst}.  Several of these are naturally expressed
+in MCJ.
+
+\paragraph{Factory}
+The requirement that all constructors be private enforces the factory
+pattern for all MCJ programs.
+\begin{verbatim}
+class MyClass {
+    static MyClass make() { return new(); }
+}
+\end{verbatim}
+Since \txt{make()} is a method, any computation can be performed,
+without the problems of constructors.
+
+\paragraph{Abstract Factory}
+ The Abstract Factory pattern is a simple application of
+inheritance in the metaclass hierarchy.  In this, several classes
+could share a kind, giving them all a single interface for
+construction. Again, the example is very simple:
+\begin{verbatim}
+class AbsFac {
+    AbsFac make() { ... }
+}
+class Derived kind AbsFac {
+    AbsFac make() { return new(); }
+}
+\end{verbatim}
+
+\paragraph{Prototype}
+As mentioned in \cite{GOF}, languages with metaclasses naturally
+support the prototype pattern.  A class is in many ways a clone of its
+metaclass.  Thus the following example uses a prototype:
+\begin{verbatim}
+class Proto {
+    Object x;
+}
+class Clone kind Proto {
+    Object x = ...;
+}
+\end{verbatim}
+
+\paragraph{Singleton}
+Finally, there is no need for the Singleton pattern; a
+class itself can serve as a singleton. 
+\begin{verbatim}
+class AClass { ... }
+class AClassSingleton kind AClass { ... }
+\end{verbatim}
+If no factory methods are
 provided, no instances can be created at runtime.  While classes can
@@
 Sophisticated macro systems, such as those found in Common Lisp
-\cite{steele-90} or Scheme \cite{SyntaxCase} allow for expression of
+\cite{steele-90} and Scheme \cite{SyntaxCase}, allow for expression of
 many forms of abstraction, as do extremely flexible object systems
@@
 In contrast, the inclusion of metaclasses does not allow for arbitrary
-additional abstraction, but instead makes the abstractions commonly
-needed easy to express.  One does not need the full power of macros to
-encode the Factory pattern.  In MCJ, we have presented a language
-where several kinds of patterns are easy to express, without adding an
-additional language on top of the original.
+additional abstraction, but instead makes the commonly needed
+abstractions easy to express.  One does not need the full power of
+macros to encode the Factory pattern.  In MCJ, we can easily express
+several kinds of patterns, without adding an additional language on
+top of the original.
 
@@
 
-As in \FJ\ and many other works, $\ox$\ stands for a possibly-empty
+As in \FJ, $\ox$\ stands for a possibly-empty
 sequence of \x.  $\SforX$ denotes a substitution of the $\oS$ for the
@@
 $@$ represents concatenation of sequences of syntactic constructs.
-Finally, $CT(\T)$ is a lookup in the class table for the definition of
-$\T$.
+Finally, $CT(\C)$ is a lookup in the class table for the definition of
+the class named $\C$.
 
@@
 directly or indirectly (however, as discussed below, cycles in the
-kind hierarchy are allowed).  Third, we assume that \this is
+kind hierarchy are allowed).  Third, we assume that \this\ is
 never used as the name of a variable, method or field.  We also take
 \Object\ to be a distinguished member of the hierarchy, with no
-specific definition, which does not have a superclass or a kind.
+specific definition that does not have a superclass or a kind.
 
@@
 In examples that follow, we omit empty type parameter lists.  Note
-that both the kind and the superclass are $\Object$.  Unlike Smalltalk
-or ObjVLisp, MCJ does not have a distinguished class
-$\Class$.\footnote{This is not the same as the \Class\ class used in
-reflection.}  Interestingly, we determined that a class $\Class$ would
-need no consideration from the type system.  Because both $\Class$ and
-$\Object$ would sit atop the hierarchy with no contents, there is no
-need to include both of them.  In a more substantial language, where
-$\Object$ might have methods or fields, there might be a use for a
-distinguished $\Class$ class, but it would not need any special
-treatment by the type system.
+that both the kind and the superclass of \C\ are $\Object$.  Unlike
+Smalltalk or ObjVLisp, MCJ does not have a distinguished class
+$\Class$.\footnote{This is not the same as the \Class\ class used for
+  reflection in the Java Programming Language.}  Interestingly, we
+determined that a class $\Class$ would need no consideration from the
+type system.  Because both $\Class$ and $\Object$ would sit atop the
+hierarchy with no contents, there is no need to include both of them.
+In a more substantial language, where $\Object$ might have methods or
+fields, there might be a use for a distinguished $\Class$ class, but
+it would not need any special treatment by the type system.
 
@@
 {
- class Pair<A,B> make(A a, B b){ 
+ static Pair<A,B> make(A a, B b){ 
     return new <A,B> (a,b) 
-} ;
+ } 
  A fst;
@@
      return new <B,B> (b, this.snd);
- };
+ }
 }
@@
 Second, we have an instance method, {\txt{setfst}}, which creates a new
-pair with the old second element, and a new first element which has
+pair with the old second element, and a new first element that has
 the same type as the second element.
@@
 {
- class O make(){return new () ()};
+ static O make(){return new () ()};
 }
@@
 
-This program, which creates two instances of {\txt O}, then creates
-a {\txt{Pair}} to hold them both, and finally selects one of the two to
+This program, which creates two instances of {\txt O}, then creates a
+{\txt{Pair}} to hold them both, and finally selects one of the two to
 become the value of the program, demonstrates three of the expression
-forms in MCJ.  {\txt{Pair<O,O>}} is a type, which, by itself, can
-be a value.  {\txt{O.make()}} is a method invocation (here with a
-class, not an instance, as the receiver).  The entire expression is a
-field access, of the {\txt{fst}} instance field.
+forms in MCJ.  {\txt{Pair<O,O>}} is a type that by itself can be a
+value.  {\txt{O.make()}} is a method invocation (here with a class, as
+the receiver).  The entire expression is a field access, of the
+{\txt{fst}} instance field.
 
@@
 With an understanding of expressions, we can examine the rest of the
-{\txt{Pair}} class.  There are two methods: a class method
+{\txt{Pair}} class.  There are two methods: a static method
 ({\txt{make}}) and an instance method ({\txt{setfst}}).  There are
@@
 instance of the syntactically enclosing class.\footnote{In the grammar
-of figure \ref{fig:syntax}, {\txt{new}} expressions are annotated with
+of Figure \ref{fig:syntax}, {\txt{new}} expressions are annotated with
 this enclosing class.  In the examples, this redundant information is
@@
 A mapping denoting an instance class {\txt C} consists of a sequence
-of class fields mapped to a sequence of expressions and is annotated
+of static fields mapped to a sequence of expressions and is annotated
 with the type \typeof{\C}. Note that the right hand sides of such
 maps need not always be values, and thus computation can take place
-inside of a mappings.  Mappings are not available to the programmer,
+inside of a mapping.  Mappings are not available to the programmer,
 and thus can only be created by operation of the reduction rules.
@@
 written $\typ \ox \oT$ respectively.  A bounds environment contains
-two bounds for each type variable, so that if $\Delta = \X \subtypeSym
-(\A,\B)$, then $\Delta(\X) = \A$ and $\Delta({\typeof \X}) = \B$.
+two bounds for each type variable, so that if $\Delta = \X \superSym
+\A \kindSym \B$, then $\Delta(\X) = \A$ and $\Delta({\typeof \X}) = \B$.
 
 The notation $\A \subtypeSym \B$ means $\A$ is a subtype of $\B$.
-Subtyping judgments are made in the context of a bounds environment,
-which relates a type to the declared bound of that type from the class
+Subtyping judgments are made in the context of a bounds environment
+that relates a type to the declared bound of that type from the class
 header.
@@
 use of the latter portion ($\GG$) of this judgment as a second argument,
-allowing us to use this rule to check both instance and class methods.
+allowing us to use this rule to check both instance and static methods.
 Method well-formedness consists mostly in fitting the body to the
@@
 Class well-formedness involves all of the above checks.
-All methods must be well-formed, and all class fields must
-have correct initialization expressions.  Further, class fields
+All methods must be well-formed, and all static fields must
+have correct initialization expressions.  Further, static fields
 must contain the instance fields of the kind.  Finally, all
@@
 expression or method call, or for evaluating the initializers for
-class fields.  There are also a number of places where either
+static fields.  There are also a number of places where either
 congruence or reduction rules can be applied.  Therefore, in the
@@
 
+\newpage
 \fieldsPreservedLemma
@@
 \methodTypingLemma
+\newpage
 
@@
 
-From these, we can conclude the desired type safety result.
+From these, we can conclude the desired type soundness result.
 
@@
 
+\paragraph{Other Systems with Metaclasses}
 A number of object-oriented languages have included some form of
@@
 in which each class has its own freely generated metaclass, defined by
-the class methods and fields of the class.  In contrast, MCJ
+the static methods and fields of the class.  In contrast, MCJ
 provides a hierarchy for structuring metaclass relationships, which
@@
 The metaclass system present in Python \cite{PythonMeta} is
-similar to that provided here, where classes inheritance the instance methods
-of their metaclasses as class methods.  However,
+similar to that provided here, where classes inherit the instance methods
+of their metaclasses as static methods.  However,
 Python is dynamically-typed, and many of the uses to which Python
@@
 metaclasses \cite{Starkiller}.
-
-Several type systems have also been proposed for languages with
-metaclasses, including the Strongtalk language \cite{Strongtalk},
-which turns Smalltalk into a structurally-typed language with
-static checking.  However, because the Smalltalk metaclass
-system is so different from the one in MCJ, many of the interesting
-aspects of the type system do not carry over.  Furthermore, the
-Strongtalk papers do not provide a formal semantics and analysis of
-the system.  A formal analysis of inheritance in Smalltalk is provided
-in \cite{cook:thesis} but this again does not consider the hierarchy
-of metaclasses presented here.
-
-Graver and Johnson \cite{graver:90} present another type system for
-Smalltalk, with a formalism and sketch of a safety proof.  Again, the
-metaclass type system is substantially different, reflecting the
-underlying Smalltalk system.  Additionally, the paper is concerned
-primarily with optimization as opposed to static checking.
 
@@
 
+\paragraph{Type Systems for Metaclasses}
+Several type systems have also been proposed for languages with
+metaclasses, including the Strongtalk language \cite{Strongtalk},
+which turns Smalltalk into a structurally-typed language with
+static checking.  However, because the Small\-talk metaclass
+system is so different from the one in MCJ, many of the interesting
+aspects of the type system do not carry over.  Furthermore, the
+Strongtalk papers do not provide a formal semantics and analysis of
+the system.  A formal analysis of inheritance in Smalltalk is provided
+in \cite{cook:thesis} but this again does not consider the hierarchy
+of metaclasses presented here.
+
+Graver and Johnson \cite{graver:90} present another type system for
+Smalltalk, with a formalism and sketch of a safety proof.  Again, the
+metaclass type system is substantially different, reflecting the
+underlying Smalltalk system.  Additionally, the paper is concerned
+primarily with optimization as opposed to static checking.
+
+\paragraph{Metaclasses and Nested Classes}
+Some of the problems are alleviated in a language with nested, or
+inner classes.  In the language Scala \cite{Scala} the following
+program expresses some of the desired invariants:\footnote{This
+  example, and this discussion, was suggested by an anonymous
+  reviewer.}
+
+\begin{verbatim}
+class Species {
+  class Member {}
+}
+object Eagles extends Species;
+class Eagle() extends Eagles.Member;
+val harry = new Eagle;
+\end{verbatim}
+
+This program automates many of the invariants required in a language
+without metaclasses.  However, two separate classes must still be
+constructed per \txt{Species}.  Further, the fundamental problem of
+not being able to group classes by functionality in the same way as
+instances remains.  Inner classes are not constrained by
+invariants imposed by their enclosing class.  In MCJ, the invariants of the
+kind are enforced by the type system.
+
+\paragraph{Prototype Systems}
 When viewed as ``instance generators'', our metaclasses are similar to
@@
 programming languages community today, including \FJ\ \cite{FJ}, upon
-which MCJ is based.  However, none of them have considered static
-methods, which are the closest analogue of the metaclass functionality
-in MCJ.
+which MCJ is based.  However, none of them have considered
+metaclasses, or even static methods, which are the closest analogue in
+the Java Programming Language of the metaclass functionality in MCJ.
 
@@
 soundness'' in the resulting system.  Further, the calculus presented
-here is sufficent for demonstrating the techincal detail, but not for
-practical programming.  Many extensions would be neccessary for MCJ to
+here is sufficient for demonstrating the technical detail, but not for
+practical programming.  Many extensions would be necessary for MCJ to
 become a usable language in practice.
@@
 One of the most important extensions is a mechanism for side effects.
-Imperative features present a some difficulties for this calculus as
+Imperative features present some difficulties for this calculus as
 formulated.  The most important problem is that the initializers for
-class fields are re-evaluated every time the class is used as an
+static fields are re-evaluated every time the class is used as an
 expression.  If those expressions had side effects, this prospect
-would make class fields unusable.  One potential solution is to
-remember the previously evaluated results, but much more work remains
-in this area.
+would make static fields unusable.  One potential solution is to
+remember the previously evaluated results, as is done in the Java
+Programming Language, but more work remains in this area.
 
@@
 other languages clearer.  Further, an implementation on the Java
-Virtual Machine would be a useful are of exploration.
+Virtual Machine would be a useful area of exploration.
 
@@
 feedback on the formal rules presented in this paper.  We also thank
-David Chase for not believing in us unless we were right, Victor
+David Chase for helpful skepticism, Victor
 Luchangco for many valuable discussions, and Guy Steele for his many
-helpful comments.
+helpful comments.  Finally, several anonymous reviewers gave very
+helpful critiques.
 

### 2004-12-16 metaclasses/semantics-proofs.tex
submitted version of paper

@@
 \begin{theorem}[Subject Reduction]
+\mbox{}\\
 If \eeproves{\typ{\e}{\SSS}} and \trans{\e}{\ep}
@@
 \begin{theorem}[Progress]
+\mbox{}\\
 If \eeproves{\typ{\e}{\SSS}} then one of the following holds:
@@
 
-Immediate from the premises of \rulename{WF-ClassDef} and \rulename{R-Class}.
+Immediate from the premises of \\\rulename{WF-ClassDef} and \rulename{R-Class}.
 
@@
 
-We know that $\e = \invoke {\Phi_{\GG}} {\mof \dd}$ and that $\ep =
+We know that $\e = \invoke {\Phi_{\GG}} {\mof \dd}$ and that \\$\ep =
 {\substituteTwoArg \ox \od \this {\Phi_{\GG}}} \e_0$. Further, $\e$
@@
 
-By the lemma $mtype$ and $mbody$ agree, $\proves
-{\emptyset;{\typ{\ox}{\oT}},{\typ{\this}{\GG}}} {\typ{\e_0'}{\Up}}$
-where $\eproves {\subtype \Up \U}$.  Then by the lemma Substitution
-Preserves Typing, $\eeproves {\typ {\ep = {\substituteTwoArg \ox \od
-\this {\Phi_{\GG}}} \e_0} \Upp}$ where $\eproves {\subtype \Upp \Up}$.
+By the lemma $mtype$ and $mbody$ agree, $\emptyset;{\typ{\ox}{\oT}},$\\
+$\proves {{\typ{\this}{\GG}}} {\typ{\e_0'}{\Up}}$ where $\eproves
+{\subtype \Up \U}$.  Then by the lemma Substitution Preserves Typing,
+$\eeproves {\typ \ep \Upp}$ where $\ep = {\substituteTwoArg \ox \od
+    \this {\Phi_{\GG}}} \e_0$ and $\eproves {\subtype \Upp \Up}$.
 
@@
 
-\proofcase{\rulename{T-Class}} In this case $\e = \CSp$ where $\CSp$ is ground.
-Then by the lemma  Agreement of {\it field-vals} and {\it fields}, $\typeparams{\CSp} = \oX \implSym \parens {\oI, \oJ}$ for some     Further, $\fieldVals{\typeof \CSp}  = \Tfe$ for some $\oT$, $\of$, and $\oe$.
-Therefore, \rulename{R-Class} applies and $\trans \SSS {\mapping \of {\SpforX \CSp} {\typeof \CSp}}$.
+\proofcase{\rulename{T-Class}} In this case $\e = \CSp$ where $\CSp$
+is ground.  Then by the lemma Agreement of {\it field-vals} and {\it
+  fields}, $\fieldVals{\typeof \CSp} = \Tfe$ for some $\oT$, $\of$,
+and $\oe$.  Further, $\typeparams{\CSp} = \oX \superSym \oI \kindSym
+  \oJ$ for some $\oX, \oI, \oJ$. Therefore, \rulename{R-Class}
+applies and \\$\trans \SSS {\mapping \of {\SpforX \CSp} {\typeof
+    \CSp}}$.
 
-\proofcase{\rulename{T-Mapping}} Either $\e$ is already a value, or $\e = \mapping \of \oe \GG$ where not all of the $\oe$ are values.
-Then by the induction hypothesis, there is some $i$ such that either $\trans {\e_i} {\ep_i}$, in which case \rulename{C-Mapping} applies, or $\e_i$ contains a bad cast, and the case is complete.
+\proofcase{\rulename{T-Mapping}} Either $\e$ is already a value, or
+$\e = \mapping \of \oe \GG$ where not all of the $\oe$ are values.
+Then by the induction hypothesis, there is some $i$ such that either
+$\trans {\e_i} {\ep_i}$, in which case \rulename{C-Mapping} applies,
+or $\e_i$ contains a bad cast, and the case is complete.
 
@@
 \item $\e = \invoke \rr {\mof \dd}$ where $\rr$ is not a mapping.  Then by the induction hypothesis, either $\rr$ contains a bad cast or $\trans \rr \rr'$ and \rulename{C-Rcvr} applies.
-\item $\e = \invoke {\Phi_{\T}} {\mof \dd}$  We know from the antecedent that $\mtype \m {\bound \T} = \arrType \oU \V$ and therefore $\mtype \m \T = \arrType \oU \V$ since $\T$ is ground.
+\item $\e = \invoke {\Phi_{\T}} {\mof \dd}$  We know from the antecedent that \\$\mtype \m {\bound \T} = \arrType \oU \V$ and therefore $\mtype \m \T = \arrType \oU \V$ since $\T$ is ground.
 Therefore, since {\it mbody} is defined everywhere {\it mtype} is defined, $\mbody \m \T = (\ox, \e_0)$ for some $\ox$ and $\e_0$.
@@
 
-\proofcase{\rulename{T-Field}} Here there are two cases, either the reciever is a mapping or not.
+\proofcase{\rulename{T-Field}} Here there are two cases, either the receiver is a mapping or not.
 In the first, by the antecedent of the typing rule, we can lookup the field successfully and apply \rulename{R-Field}.
@@
 \begin{lemma}[Substitution Preserves Typing]
+\mbox{}\\
 If $\dgprove {\typ \e \T}$ and $\Gamma = \typ{\ox}{\oS}$ and
@@
 \begin{lemma}[Subtyping Preserves Method Typing]
+\mbox{}\\
 If $\mtype \m \U = \arrType \oT \SSS$ and $\proves \emptyset {\subtype \V \U}$ then
@@
 
-In this case, $\V = \CR$, where $CT(\C) = {\typeFront {\typeArgsBounds \C \X \I \J} \B \A}
-\braces{... \oM}$ and $\U = \RforX \A$.  We need to show that $\mtype
+In this case, $\V = \CR$, where \\$CT(\C) = {\typeFront {\typeArgsBounds \C \X \I \J} \B \A}
+\braces{... \oM}$ \\and $\U = \RforX \A$.  We need to show that $\mtype
 \m \CR = \arrType \oT \SSS$.   By lemma $methods$ is well-defined, it
@@
 
-We know that $\simpleMethod \in \methods \U$.  So, we proceed by
+Since we know that $\simpleMethod \in$ \\$\methods \U$, we proceed by
 induction on the derivation of $\methods \U$.
@@
 
-\subcase{$\U = \typeArgs \D \W$}
+\subcasenl{$\U = \typeArgs \D \W$}\\
 Then, $\U = \RforX \A$.  By \rulename{MethodsClass}, $\methods \V =
@@
 
-Analagous to \rulename{S-Super}.
+Analogous to \rulename{S-Super}.
 
@@
 \begin{lemma}[Fields are Preserved by Subtypes]
+\mbox{}\\
 If $\fields \U = \oF$ and ${\proves {\emptyset} {\subtype \V \U}}$ then $\exists\ \oG$ where $\fields \V = \oF\ @\ \oG$.
@@
 {
-
 \myProof
@@
 
-\proofcase{$\V = \CR$}
+\proofcasenl{$\V = \CR$}
 
@@
 
-Let $\T$ be the intermediate type. Then $\fields \V = {\fields \T}\ @\
+Let $\T$ be the intermediate type. Then \\$\fields \V = {\fields \T}\ @\
 {\ov {\HH'}}$ and $\fields \V = {\fields \T}\ @\ {\ov {\HH''}}$ by the
@@
 
-\proofcase{$\V = \typeof{\CS}$}
+\proofcasenl{$\V = \typeof{\CS}$}
 
@@
 As above.
-
-\subcase{\rulename {S-Kind}}
+\newpage
+\subcasenl{\rulename {S-Kind}}
 
@@
 well-formedness of $\C$, we know that $\fields \B \subseteq \oW\ \oh$.
-Then $\SforX \fields \B \subseteq \SforX \oW\ \oh$, and $\fields {\SforX \B} \subseteq
-\SforX \oW\ \oh$ by lemma Substitution Distributes over $fields$.
+Then $\SforX \fields \B \subseteq \SforX \oW\ \oh$, and $\fields
+{\SforX \B} \subseteq$\\$\SforX \oW\ \oh$ by lemma Substitution
+Distributes over $fields$.
 \end{description}
@@
 \begin{lemma}[Agreement of $fields$ and {\it field-vals}]
+\mbox{}\\
 If $\fields \GG = \Tf$ and $\fieldVals \GG = \oTp\ \ofp\ \oe$ then
@@
 \newcommand{\fieldsAgreeProof}{
-\myProof The proof is exactly analagous to the proof of lemma {\it mtype} and {\it mbody} Agree.
+\myProof The proof is exactly analogous to the proof of lemma {\it mtype} and {\it mbody} Agree.
 }
@@
 
-\proofcase{\rulename{S-Kind}} Analagous to \rulename{S-Super}.
+\proofcase{\rulename{S-Kind}} Analogous to \rulename{S-Super}.
 

### 2005-01-19 classes/semantics/assn1.tex
first semantics assignment

\documentclass{article}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{define}{Definition}
\newcommand{\proofcase}[1]{\item[Case {\textsc {#1}}:]}

\begin{document}

\title{Assignment 1}
\author{Sam Tobin-Hochstadt}

\maketitle

\begin{lemma}[Substitution Preserves Typing]\mbox{}\\
If $e_1 : t'$ and $e_2 : t$ then $e_1[x^t/e_2]:t$.
\end{lemma}

\begin{proof}
  We proceed by structural induction and then by cases on the final
  rule in the derivation of $e_1 : t'$.

\begin{itemize}
  \proofcase{T-App} In this case, $e_1 = (e\ e')$.  Therefore
  $e_1[x^t/e_2] = (e\ e')[x^t/e_2] = (e[x^t/e_2]\ e'[x^t/e_2])$.  By the
  induction hypothesis, if $(e\ e') : t'$ then $(e[x^t/e_2]\
  e'[x^t/e_2]) : t'$.
  \proofcase{T-Abs} In this case, $e_1 = \lambda y.e$.  Then there are
  two cases.  If $y \neq x$, then substitution has no effect, and the
  statement is trivially true.  Otherwise, by the induction
  hypothesis, if $e : s$ then $e[e_2/x^t] : s$, and the premises of
  the rule are still satisfied.
  \proofcase{T-Const}  Here, substitution has no effect, and thus the
  lemma is trivially true.  
  \proofcase{T-Var}  Here, $e_1 = y$.  There are two cases.  If $y
  \neq x$, then substitution has no effect, and we are done.
  Otherwise, $y = x$, and $t = t'$, and therefore $e_1[x^t/e_2] = y$,
  and $y : t'$.  
\end{itemize}

\end{proof}

If $e_0 = ((\lambda x^t.e_1)\ e_2) : t'$, then the desired theorem is

\begin{theorem}[$\beta$-reduction Preserves Typing] \mbox{}\\
If $((\lambda x^t.e_1)\ e_2) : t'$ then $e_1[x^t/e_2]:t$.
\end{theorem}

To prove this, simply consider the typing rules for application and abstraction, which
shows that the abstraction must have type $t \rightarrow t'$, and therefore
$e_1 : t'$ and $e_2 : t$, from the typing of the application.  Thus we
can apply our lemma to obtain the desired conclusion.

\end{document}

### 2005-01-19 classes/semantics/assn1.tex
formatting change

@@
 
+\begin{proof}
 To prove this, simply consider the typing rules for application and abstraction, which
@@
 can apply our lemma to obtain the desired conclusion.
+\end{proof}
 

### 2005-02-02 classes/semantics/assn3.tex
semantics assignment 3

\documentclass{article}
\usepackage{mmm}
\usepackage{sample}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{fullpage}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{define}{Definition}

\begin{document}

\title{Assignment 3}
\author{Sam Tobin-Hochstadt}

\newmeta\e{e}

\maketitle

\subsection*{Problem 1}

Let $\P1 = \mathtt{Y\ (\l f.\l x. \l y.if\ (zero?\ x)\ y\ (+1\ (f\
  (-1\ x)\ y)))}$.

Then we would like to prove

\begin{theorem}
If $\e1 \rightarrow^* \c{n}$ and $\e2 \rightarrow^* \c{m}$ then $(\P1\
\e1\ \e2) \rightarrow^* \c{n+m}.$
\end{theorem}

\begin{proof}
  We prove this by induction on $n$.

\paragraph{Base Case}  Here $n=0$.  Then, the test in the conditional
will be true, and therefore $(\P1\
\e1\ \e2) \rightarrow^* \e2 \rightarrow^* \c{m} = \c{0+m}$.

\paragraph{Inductive Step} Here we assume that our hypothesis is true
for all $k \leq n$.  We prove the fact at $n+1$.  Since $n+1 > 0$, the
test in the conditional must be false.  Therefore, we are left with
the following term: $(+1\ (\P1\ (-1\ x)\ y))$.  By the induction
hypothesis at $n$, we have that $(\P1\ (-1\ x)\ y) \rightarrow^*
\c{n+m}$.  Therefore,  $(+1\ (\P1\ (-1\ x)\ y)) \rightarrow^* (+1\
\c{n+m}) \rightarrow \c{n+1+m}$.

\end{proof}

\subsection*{Problem 2}

Let $\P2 = \mathtt{Y\ (\l f.\l x. \l y.if\ (zero?\ x)\ y\ (f\
  (-1\ x)\ (+1\ y))))}$.

Then we would like to prove

\begin{theorem}
If $\e1 \rightarrow^* \c{n}$ and $\e2 \rightarrow^* \c{m}$ then $(\P2\
\e1\ \e2) \rightarrow^* \c{n+m}.$
\end{theorem}

\begin{proof}
  We prove this by induction on $n$.

\paragraph{Base Case}  Here $n=0$.  Then, the test in the conditional
will be true, and therefore $(\P1\
\e1\ \e2) \rightarrow^* \e2 \rightarrow^* \c{m} = \c{0+m}$.

\paragraph{Inductive Step} Here we assume that our hypothesis is true
for all $k \leq n$.  We prove the fact at $n+1$.  Since $n+1 > 0$, the
test in the conditional must be false.  Therefore, we are left with
the following term: $(\P2\ (-1\ x)\ (+1\ y))$.  But the induction
hypothesis at $n$, we have that $(\P2\ (-1\ x)\ (+1\ y)) \rightarrow^*
\c{n+m+1}$.

\end{proof}

\newcommand\red{\rightarrow^*} 
\newcommand\app[2]{(#1\ #2)}
\newcommand\PP[1]{\app{\P{}}{#1}}

\newpage

\subsection*{Problem 3}

Consider the following function: $\P{} = \mathtt{Y(\l f.\l x.\ if\ (>\
  x\ 100)\ (-\ x\ 10)\ (f\ (f\ (+\ x\ 11))))}$.

\begin{theorem}
   If $\e{} \red \c{n}$, and $\PP{\e{}} \red \c{k}$ then the following holds:
   \begin{enumerate}
   \item if $n > 100$, then $k = n-10$
   \item otherwise, $k = 91$ 
   \end{enumerate}

\end{theorem}

\begin{proof}

It is obvious by inspection that for any $\e{} \rightarrow^* \c{k}$
where $k \geq 100$, $(P\ k) \rightarrow^* \c{k-10}$.

Further, since we are only considering the cases when the function
terminates, and the computation produces no other side effects, we can
treat the semantics as if they were applicative-order.  This
simplifies the reasoning process.

Then, consider $90 \leq k \leq 100$.  In this case, $\PP{k}
\red \PP{\PP {(k+11)}} \red \PP{(k+1)}$ since $k+11>100$.  Therefore,
$\PP{k} \red \PP{101} \red 91$.

Alternatively, $k < 90$.  Then $k+11 \leq 100$.  Therefore, $\PP{k}
\red \PP{\PP{k+11}}$.  Now, if $k+11 \geq 90$, the previous discussion
applies, and $\PP{\PP{k+11}} \red \PP{91} \red 91$.  So, we have that
$\PP{k} \red 91$ if $k \geq 79$.  But if $k+11 \geq 79$, then  by
the foregoing discussion, $\PP{k+11} \red 91$ and therefore
$\PP{\PP{k+11}} \red 91$.  Therefore, $\PP{k} \red 91$ if $k \geq 68$.

Identical arguments prove that $\PP{k} \red 91$ if $k \geq 57$, and if
$k \geq 46$, and if $k \geq 35$, and if $k \geq 24$, and if $k \geq
13$, and if $k \geq 2$.  If $k = 1$ or $k = 0$, then $PP{k} \red
\PP{\PP{k+11}}$.  Since $k+11 = 12$ or $11$ respectively,
$\PP{\PP{k+11}} \red 91$.

\end{proof}

\end{document}

### 2005-02-09 classes/semantics/assn4.tex
assignment 4

\documentclass{article}[12pt]
\usepackage{mmm}
\usepackage{sample}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{fullpage}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{define}{Definition}

\newcommand\red{\rightarrow^*}

\begin{document}

\title{Assignment 4}
\author{Sam Tobin-Hochstadt}

\newmeta\e{e}
\newmeta\vv{v}
\newmeta\kk{k}
%\newcommand\v{\vv{}}

\maketitle

\begin{theorem}
  If $\e{} \Downarrow v$ then $<e,E> \rightarrow^* <v,E>$.
\end{theorem}

\begin{proof}
  We proceed by induction the derivation of $\e{} \Downarrow v$ and by
  cases on the last rule of the derivation.

\item[[$v \Downarrow v$]] Then $<e,E> \rightarrow^* <v,E>$ in zero steps.

\item[[$\beta$-reduction]] In this case $e = (\e1\ \e2)$.  Then by focus-down,
  $<e,E> \rightarrow^* <\e1,E[([]\ \e2)]>$.  Then by the induction
  hypothesis, this reduces to $<\lambda x.e',E[([]\ \e2)]>$.  Then we
  can apply focus-up to get $<(\lambda x.e'\ \e2),E>$, which reduces
  by the $\beta$-reduction rule to $<e'[\e2/x],E>$, which by the
  induction hypothesis reduces to $<v,E>$.

\item[[if true]] In this case $e = (\e1\ \e2)$.  Then by focus-down,
  $<e,E> \rightarrow^* <\e1,E[([]\ \e2)]>$. By the induction
  hypothesis, this reduce to $<if true e',E[([]\ \e2)]>$.  By applying
  focus-up, we get $<if true e' \e2,E>$, which takes a step to
  $<e',E>$.  By the induction hypothesis, this reduces to $<v,E>$.

\item[[if false]] This is exactly analagous to the above case.

\item[[zero?]] In this case, $e = (\e1\ \e2)$.  By focus-down, we have
  $<e,E> \rightarrow^* <\e1,E[([]\ \e2)]>$. By the induction
  hypothesis, we then have $<\e1,E[([]\ \e2)]> \red <zero?,E[([]\
  \e2)]>$.  Then by focus-up and focus-down, we have $<zero?,E[([]\
  \e2)]> \red <\e2,E[(zero?\ [])]>$.  By the induction hypothesis and
  focus-up, this reduces to $<(zero?\ \kk{m},E>$ which reduces to
  either true or false.  Since both true and false are values, we now
  have $<v,E>$ as desired.

\item[[+1, -1]] These cases are exactly analagous to zero?.

\item[[if]] In this case, $e = (\e1\ \e2)$.  By focus-down, we have
  $<e,E> \rightarrow^* <\e1,E[([]\ \e2)]>$. By the induction
  hypothesis, we then have $<\e1,E[([]\ \e2)]> \red <if,E[([]\
  \e2)]>$. Then by focus-up and focus-down, we have $<if,E[([]\
  \e2)]> \red <\e2,E[(if\ [])]>$.  By the induction hypothesis, this
  reduces to $<v',E[(if\ [])]>$.  By applying focus-up, we get $<if
  v,E>$, which is a value.

\item[[Other cases for if]] The other case for if proceeds as above.

\item[[Y]] In this case $e = (\e1\ \e2)$.  Then by focus-down,
  $<e,E> \rightarrow^* <\e1,E[([]\ \e2)]>$.  Then by the induction
  hypothesis, this reduces to $<Y,E[([]\ \e2)]>$.  By applying
  focus-up, we then get $<(Y\ \e2),E>$ which reduces by the reduction
  rule for $Y$ to $<(\e2\ (Y\ \e2)),E>$.  By the induction hypothesis,
  this reduces to $<v,E>$.

\end{proof}

\end{document}
